---
title: "Introduction to JoiNMe"
output-dir: ../docs/articles
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Installation

```r
# install.packages("remotes")
# remotes::install_github("trinhdhk/JoiNMe")
```

For Stan backends:

```r
# CmdStanR backend (recommended)
# install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
# cmdstanr::install_cmdstan()

# RStan backend (optional)
# install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

## Quick start

## Conceptual overview

The `JoiNMe` model links two components:

1. A multivariate longitudinal model for repeated biomarker measurements.
2. A survival model whose hazard depends on summaries of the longitudinal
   process at time $t$.

For the longitudinal part, each marker can follow its own distribution
(Gaussian, Student-$t$, Poisson, negative binomial, Bernoulli, beta,
skew families, or ordinal). The linear predictor is decomposed into fixed
effects, subject-level random effects, marker-level random effects, and
optional marker-by-subject random effects. The survival part uses a
proportional hazards model with a flexible spline baseline hazard and
time-dependent association terms.

The association terms implemented in `joinme` package are:

- Current value (CV) of the latent linear predictor.
- Current slope (CS) of the latent linear predictor.
- Correlation association (CORR) derived from off-diagonal marker-by-subject
  correlations.
- Variance-covariance association (VCOV) derived from lower-triangular
  marker-by-subject Cholesky-factor entries taken directly from `L`.

Each association term can be left on its original scale or transformed
using a functional expression, a monotone I-spline, or a piecewise linear
mapping. These transformations are configured in R and passed to Stan as
data so that the likelihood remains fully deterministic.

## Family-shared distributional semantics

For mixed longitudinal families, `JoiNMe` treats distributional parameters as
**shared by family** when no distributional regression is supplied.

Concretely:

- Markers with the same family share one latent baseline process for each
  applicable distributional parameter.
- Markers from different families do not share those baselines.
- Parameters are instantiated only when a family uses them
  (for example `nu` for Student-`t`, `phi` for negative-binomial).

Example interpretation:

- If markers `A` and `B` are both Student-`t` and marker `C` is Gaussian,
  then `A` and `B` share the same baseline `sigma` and `nu` processes,
  while `C` has its own `sigma` process and no `nu`.

### What changes when `formulaDist` is provided?

Supplying a distributional regression (for example `sigma ~ 1 + time`) switches
that parameter from a family-level constant baseline to a regression-driven
process. In that mode, summaries report distributional regression coefficients
(`(Intercept)`, covariate effects, and random-effect SD terms when present)
rather than only baseline family constants.

Illustrative `formulaDist` specification:

```r
formulaDist <- list(
  sigma ~ 1 + time,
  nu ~ 1
)
```

This means:

- `sigma` is modelled through regression (fixed/random effects if included),
- `nu` is regression-driven with an intercept-only structure,
- other distributional parameters without formulas continue using family-shared
  baselines when applicable.

## Data requirements (minimal)

You need two data frames:

- `dataLong`: long-format longitudinal data with columns for subject id,
  marker id, time, outcome, and covariates.
- `dataEvent`: one row per subject with event/censoring time, event
  indicator, and baseline covariates.

Time is internally scaled to $[0, 1]$ for numerical stability. The model
returns coefficients on the original time scale.

### Step 1: Load packages and set a reproducible seed

```{r setup}
if (requireNamespace("pkgload", quietly = TRUE)) {
  repo_root <- normalizePath("..", mustWork = FALSE)
  if (file.exists(file.path(repo_root, "DESCRIPTION"))) {
    pkgload::load_all(repo_root, quiet = TRUE, compile = FALSE)
  } else {
    library(joinme)
  }
} else {
  library(joinme)
}
library(survival)

set.seed(2026)
```

### Step 2: Simulate a small dataset

We generate a modest dataset so the workflow runs quickly while still
containing both longitudinal and survival information.

```{r intro-simulate}
sim <- simulate_joinme(
  n_id = 10,
  families = list(
    jm_family("student_t"),
    jm_family("student_t")
  ),
  n_obs_per_marker_per_id = 4,
  times_obs = seq(0, 5, length.out = 8),
  baseline_hazard = list(type='weibull', shape = 2, scale = 20),
  quadrature_nodes = 31,
  seed = 2026,
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.6)
)

# Gauss-Kronrod nodes/weights are fixed in Stan; `quadrature_nodes` selects
# the node count used to build the design matrices in R.

head(sim$dataLong)
head(sim$dataEvent)
```

### Step 3: Specify longitudinal and survival models

The longitudinal formula defines fixed effects and random effects. The survival
formula defines baseline covariates in the hazard model.

Formula-scoped weighting can be declared directly in grouping terms via
`weighted(group, weights = <column>)`, including marker-level grouping declarations.

```{r intro-formulas}
formulaLong <- y ~ 1 + time + x1 +
  (1 + time || id) +
  (0 + x1 + (1 + time || id) || marker)

formulaEvent <- survival::Surv(time, event) ~ 1 + x1 + x2
```

```{r intro-weighted-formula, eval=FALSE}
formulaLong_weighted <- y ~ 1 + time + x1 +
  (1 + time | weighted(id, weights = id_w)) +
  (0 + x1 + (1 + time | weighted(id, weights = id_w)) | weighted(marker, weights = marker_w))
```

### Step 4: Fit the joint model

We link the survival hazard to the current value (CV) of the longitudinal
process using an identity transformation.

```{r intro-fit}
fit <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  families = "student_t",
  transforms = joinme_tf(cv_total = "identity"),
  basehaz = 'bs', 
  control = list(
    engine = "rstan",
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 500,
    iter_sampling = 500,
    seed = 2026,
    refresh = 0,
    adapt_delta = 0.85
  )
)
```

### Step 5: Summarise fixed effects and associations

```{r intro-summaries}
summary(fit) # or posterior_summary(fit)fixef(fit)
coef(fit)
posterior_assoc(fit, summary = TRUE)
```

`posterior_summary()` is an alias for `summary()`. `posterior_assoc()` returns
association-specific posterior output in a format that stays close to the main
summary tables. For covariance-style channels (`corr`, `vcov`), the printed and
tabular output now uses readable matrix labels instead of raw numeric indices.
`coef()` returns coefficients on the model-matrix scale after adding the
population-level contribution and the matching group-level deviation. As with
`fixef()` and `ranef()`, use `summary = FALSE` to retrieve posterior draws
instead of posterior summaries.

### Step 6: Dynamic prediction for one subject

We condition on the observed history up to the last measurement for subject 1
and obtain predictive summaries beyond that time.

```{r intro-predict}
ndL <- sim$dataLong[sim$dataLong$id == 1, ]
ndE <- sim$dataEvent[sim$dataEvent$id == 1, ]
time_start <- max(ndL$time)

pred <- posterior_predict(
  fit,
  newdataLong = ndL,
  newdataEvent = ndE,
  time_start = time_start,
  times = seq(time_start, time_start + 1, length.out = 20),
  control = list(
    n_samples = 50, 
    n_pred_draws = 1000,
    iter_warmup = 200,
    chains = 1
    ),
  seed = 2026
)

# control$n_samples: extracted posterior parameter draws
# control$n_pred_draws: dynpred output draw count

# use predict(..., scale = c("epred", "linpred", "predict"))
# to return multiple longitudinal scales in one call (default = all scales)
```

### Step 7: Plot longitudinal and survival predictions

```{r intro-plot}
plot(pred, type = c("longitudinal", "survival"), combined = TRUE)
```

If `patchwork`/`cowplot` is unavailable, `combined = TRUE` falls back to the
standard per-subject structured return (grouped by subject and outcome) instead
of any flattened helper list.

The heatmap is an alternative to the usual separated marker curves. It orders
markers by the magnitude of posterior mean change and makes tiles transparent
when the posterior sign certainty does not exceed `1 - threshold`.

`plot(fit, ...)` uses fitted posterior samples directly. For conditioning at
named baseline profiles, call `predict(fit, ..., condition = make_conditions(...))`
and then plot the resulting `JoiNMeDynPred` object.

### Step 8: Diagnostics and extraction helpers

```{r intro-diagnosis}
diagnosis(fit)
diagnosis(pred)
```

`ranef(fit)`, `coef(fit)`, and `vcov(fit)` are nested by `formulaLong` and
`formulaDist`. The posterior aliases `posterior_fixef()`,
`posterior_ranef()`, and `posterior_coef()` are short forms for requesting the
draw-level output of those extractors. For prediction objects, `ranef(pred)`
and `vcov(pred)` are available only when marker covariance is configured to
depend on id.

## Next steps

- See the modelling workflow in `JoiNMe-workflow`.
- Explore association structures in `JoiNMe-associations`.
- Review the statistical details in `JoiNMe-model` and `JoiNMe-theory`.
