---
title: "joinme: Association Construction"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
options(width = 80)
```

# Purpose

This vignette focuses on **association term construction**, including transformation selection, interpretability, and practical trade-offs.

Association transforms apply to all supported longitudinal families (including
double_exponential, skew_double_exponential, beta, and cumulative_logit) and to
competing-risk survival models.

## Step-by-step: how association summaries are constructed

At the event time and at the default 15 Gauss-Kronrod nodes used for integration
(configurable), the
longitudinal linear predictor is decomposed into a mean component and a marker
component. From these components, `joinme` constructs:

- **CV_total**: total current value (mean + marker components).
- **CV_mean**: current value of the mean component only.
- **CV_marker**: current value of the marker component only.
- **CS_total**: total current slope, approximated by a forward finite difference
  and rescaled to the original time units.
- **CS_mean**: slope of the mean component.
- **CS_marker**: slope of the marker component.
- **CORR**: a subject-specific variance summary of the marker-by-id covariance
  structure, constant across time.
- **VCOV**: a subject-specific lower-triangular summary of the marker-by-id
  Cholesky factor `L`, also constant across time. The channel uses the entries
  of `L` directly rather than the reconstructed covariance matrix `L L'`.

Each component can be included or excluded using the `assoc` argument. If
`assoc` does not include a component, the corresponding coefficient is prior-only
and does not enter the likelihood.

The CORR and VCOV components require marker-by-id random effects. If the marker
block does not include an inner `( ... | id )` term, then `Q_idm = 0` and these
covariance-style associations are not available.

# Marker-weighted CV and CS

When multiple biomarkers are modelled, `joinme` constructs marker-average association
components using weights $\omega_d$ (one per marker). For marker-aggregated
terms (`cv_total`, `cv_marker`, `cs_total`, `cs_marker`), transforms are applied
at the marker level first and then averaged with weights. This matters whenever
transforms are nonlinear.

For example, with marker-specific total current values $CV_{id}(t)$ and transform
$f_{cv}$:

$$
  ext{cv\_total}(t) = \frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cv}\!\left(CV_{id}(t)\right),
$$

which differs from $f_{cv}(\frac{1}{D}\sum_d \omega_d CV_{id}(t))$ unless $f_{cv}$
is linear.

For slopes, let $CS_{id}(t)$ be a marker-specific slope feature and $f_{cs}$ the
configured CS transform. Then

$$
  ext{cs\_total}(t) = \frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cs}\!\left(CS_{id}(t)\right),
$$

and analogously for `cs_marker`. This is intentionally not equal to
$f_{cs}(\frac{1}{D}\sum_d \omega_d CS_{id}(t))$ unless $f_{cs}$ is linear.

In practice, provide fixed weights via `marker_weights`, or set
`fixed_marker_weights = FALSE` to estimate signed perturbations around base
weights via `marker_weights + z_marker_weights` with
`z_marker_weights ~ N(0, 1)`. The resulting weights are used directly in
marker-aggregated association terms. The simulator reports them in
`truth$marker_weights`.

## Plotting fitted association curves

`plot(fit, type = "association")` uses a compact association plotting bundle
stored on the fitted `JoinMeFit` object. The bundle keeps only the posterior
quantities needed for plotting: active association coefficients, marker weights,
transform coefficients, and cached model-implied raw support ranges.

This has two practical consequences:

- association plots remain usable after serializing a fitted object without
  depending on transient CmdStan CSV files for this feature,
- default x-axis grids are taken from model-implied feature support when
  available, with spline plots falling back to knot support only when the fitted
  spline support is narrower than the model-implied raw range.

# Association coefficients and naming

Association coefficients are denoted with the $\alpha$ prefix (e.g., `alpha_cv_total`).
This follows joint-model conventions and avoids confusion with the linear predictor
$\eta$ used throughout the longitudinal and survival submodels.
Total, mean, and marker associations use a positive non-centred parameterisation
$\alpha = z_{\alpha} \cdot sd_{\alpha}$ for stable sampling.

When an association component is not selected in `assoc`, its coefficient is
prior-only and does not enter the likelihood. Summaries only report active
components to keep interpretation focused on the fitted association structure.

# Transform modes

`joinme` supports four transformation modes for association features:

- **Identity** (`type = "identity"`): no transformation.
- **Functional** (`type = "functional"`): a user-defined expression converted
  into a fixed bytecode program.
- **Monotone I-spline** (`type = "ispline"` or `"ispline_penalised"`): a smooth
  monotone mapping.
- **Bounded-domain monotone I-spline** (`type = "ispline_expit"` or
  `"ispline_expit_penalised"`): the same monotone spline machinery, but applied
  to `plogis(x)` rather than directly to `x`.
- **Piecewise linear** (`type = "pwlin"`): linear interpolation across knots.

Interpretation note for recovery experiments:

- If you use a fixed spline (`type = "ispline"` or an expit variant with
  explicit `coeff`), the association coefficient is interpretable on that fixed
  transformed scale and can be checked directly in simulate-fit recovery tests.
- If you use a jointly estimated penalised spline (`*_penalised` without `y`),
  the transform shape and the association coefficient are learned together. The
  fitted alpha is still meaningful inside the model, but coefficient recovery is
  typically weaker than in the fixed-transform case because the data must learn
  both the transformed scale and the hazard weight simultaneously.

## Default values and minimal parameterisations

- **Identity**:
  - Minimal form: `list(type = "identity")`
  - Default when a transform term is omitted or set to `NULL`.
  - No other parameters are used.
- **Functional**:
  - Minimal form: `list(type = "functional", expr = ~ log1p(x))`
  - Required field: `expr`.
  - No other defaults beyond the supplied expression.
- **I-spline**:
  - Minimal form:
    `list(type = "ispline", knots = c(-1, 0, 1), coeff = c(0, 0.3, 0.8, 1.1, 1.3))`
  - Required fields: `knots`, `coeff`.
  - Default: `degree = 3` if omitted.
- **Penalised I-spline**:
  - Legacy plug-in mode:
    `list(type = "ispline_penalised", x = seq(-2, 2, length.out = 50), y = exp(seq(-2, 2, length.out = 50)), n_knots = 6, degree = 3, lambda = 1)`
  - Stan-estimated mode:
    `list(type = "ispline_penalised", x = seq(-2, 2, length.out = 50), n_knots = 6, degree = 3, lambda = 1)`
  - Defaults: `n_knots = 6`, `degree = 3`, `lambda = 1`.
  - If `knots` are omitted in Stan-estimated mode, they are derived from `x`
    quantiles.
- **Expit-based penalised I-spline**:
  - Legacy plug-in mode:
    `list(type = "ispline_expit_penalised", x = seq(0.02, 0.98, length.out = 50), y = seq(0.02, 0.98, length.out = 50)^0.8, n_knots = 6, degree = 3, lambda = 1)`
  - Stan-estimated mode:
    `list(type = "ispline_expit_penalised", x = seq(0.02, 0.98, length.out = 50), n_knots = 6, degree = 3, lambda = 1)`
  - Defaults: `n_knots = 6`, `degree = 3`, `lambda = 1`.
  - The raw association feature is mapped through `plogis(x)` at runtime, and
    the spline basis is built on that bounded domain. Supply training `x` and
    explicit `knots` directly on the expit scale in `[0, 1]` so the fitted
    spline domain matches the runtime basis.
- **Piecewise linear**:
  - Minimal form:
    `list(type = "pwlin", x = c(-2, -1, 0, 1, 2), y = c(0.2, 0.5, 1, 0.5, 0.2))`
  - Required fields: `x`, `y`.
  - No additional defaults.

# Functional Transform

```{r functional-transform, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "functional",
    expr = ~ sinh(x) + tanh(x)
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```

The functional evaluator supports common arithmetic and functions such as
`log`, `exp`, `sqrt`, `sin`, `cos`, `tanh`, plus aliases `sigmoid`/`expit`,
`softplus`/`log1p_exp`, and helpers like `cbrt` and `power`.

Defaults:

- Required field: `expr`.
- No automatic defaults beyond the expression you provide.

```{r functional-transform-mixed-demo, eval=FALSE}
transforms <- list(
  cv_total = list(type = "functional", expr = ~ inv_logit(x)),
  cv_mean = list(type = "functional", expr = ~ exp(0.2 * x)),
  cv_marker = list(type = "functional", expr = ~ power(x, 2)),
  cs_total = list(type = "functional", expr = ~ sqrt(abs(x) + 1e-6)),
  corr = list(type = "functional", expr = ~ cbrt(x))
)
```

**When to use**

- You want a fixed functional form with no spline fitting.
- You need fast reconfiguration without recompiling Stan.

# Monotone I-Spline

For `type = "ispline"`, there is no fitting step. You specify the transform
shape directly:

- `knots`: knot locations (first and last are boundary knots),
- `coeff`: non-negative I-spline basis coefficients,
- `degree`: spline degree (usually `3`).

In this mode, `x`, `y`, and `lambda` are **not used**.

Defaults:

- `degree = 3` if omitted.
- `knots` and `coeff` must be supplied explicitly.

For `type = "ispline_expit"`, the same direct specification applies, except the
I-spline basis is evaluated on `plogis(x)`. The runtime association feature is
still shown on its raw scale, but the transform's `knots` must be supplied on
the expit scale because that is the spline domain.

```{r monotone-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.5, 1, 1.2, 1.5, 1.6),
    degree = 3
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```

**When to use**

- You want monotone, smooth effects with constrained shape.

# Penalised Monotone I-Spline

For `type = "ispline_penalised"`, there are now two supported modes:

- **Legacy plug-in mode**: provide `x` and `y`, and the package first learns a
  monotone spline in R before passing fixed coefficients into Stan.
- **Stan-estimated mode**: omit `y`, keep `knots` (or `n_knots`) plus `lambda`,
  and Stan estimates the monotone spline coefficients jointly with the rest of
  the model.

Meaning of the arguments:

- `x`: raw association-feature values, used either as training inputs or to
  help choose knots.
- `y`: optional target transformed values at those `x` points. If present, you
  are telling the package the shape you want to approximate.
- `lambda`: smoothness penalty weight (higher = smoother, lower = more flexible).

Defaults:

- `n_knots = 6` if `knots` are omitted.
- `degree = 3` if omitted.
- `lambda = 1.0` if omitted.
- In helper-based plug-in fitting, `weights = NULL` means equal weights and
  `diff_order = 2` is the default penalty order.

Practical tip: start with `lambda = 1.0`, then decrease if too flat or increase
if too wiggly.

```{r penalised-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "ispline_penalised",
    x = seq(-2, 2, length.out = 40),
    y = exp(seq(-2, 2, length.out = 40)),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

```{r penalised-ispline-stan, eval=FALSE}
transforms <- list(
  corr = list(
    type = "ispline_penalised",
    x = seq(-2, 2, length.out = 40),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

**When to use**

- You want a monotone transform with smoothing, either anchored by `(x, y)` or
  estimated jointly in Stan.
- Penalisation avoids overfitting when knot counts are moderate.

# Piecewise-Linear

```{r piecewise-linear, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "pwlin",
    x = c(-2, -1, 0, 1, 2),
    y = c(0.2, 0.5, 1, 0.5, 0.2)
  )
)
```

Defaults:

- `x` and `y` must both be supplied.
- No additional defaults are added.

**When to use**

- You need simple, interpretable transformations.

# Association Diagram

```{mermaid}
%%| fig-cap: "Association components mapped into the survival predictor."
graph TD
  CV[CV total] --> FCV[f_cv]
  CS[CS total] --> FCS[f_cs]
  V[Corr] --> FCORR[f_corr]
  FCV --> ETA[eta assoc]
  FCS --> ETA
  FCORR --> ETA
```

# Hierarchical Covariance Regression

```{mermaid}
%%| fig-cap: "Covariance regression structure for marker-by-id effects."
graph TD
  X[Covariates] --> LP[Linear predictor]
  UL[Latent effect] --> LP
  LP --> L[Cholesky factor]
  L --> W[Marker-by-id effects]
  Z[Latent z] --> W
```

# Fit with Custom Association

```{r fit-custom-association, eval=FALSE}
fit <- joinme(
  formulaLong = y ~ 1 + time + x1 +
    (1 + time || id) +
    (0 + x1 + (1 + time || id) || marker),
  dataLong = sim$dataLong,
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  dataEvent = sim$dataEvent,
  transforms = transforms,
  assoc = c("cv_total", "corr", "vcov")
)
```
