---
title: "Block-specific priors and shrinkage in JoiNMe"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

# Purpose

A joint multivariate model contains parameter blocks with quite different
scientific meanings. A longitudinal fixed effect, a survival-association
coefficient, an affine shift inside a nonlinear transformation, and a
standardised marker effect should not acquire the same prior merely because
they are fitted in one Stan programme. `jm_prior()` therefore declares the
family of each block independently.

The available declarations are `prior_normal()`, `prior_student_t()`,
`prior_laplace()`, `prior_horseshoe()`, and `prior_lkj()`. The `prior_` prefix
is deliberate: it makes code explicit and avoids collisions with constructors
from other modelling packages.

# Complete declaration

```{r}
model_priors <- jm_prior(
  beta = prior_normal(
    mu = list(0, 1, 2, 3),
    scale = list(1, 1, 2, 0.5)
  ),
  alpha = prior_student_t(df = 4, mu = 0, scale = 1),
  iota = prior_laplace(mu = 0, scale = 0.75),
  marker = prior_normal(),
  marker_weight = prior_student_t(df = 6),
  vcov_sd = prior_normal(mu = 0, scale = 1),
  vcov_corr = prior_student_t(df = 4, mu = 0, scale = 0.75),
  lkj = prior_lkj(eta = 2),
  class_probability = 1,
  class_regression = prior_normal(mu = 0, scale = 1)
)
```

This object can be supplied unchanged to `joinme()` or `joinme_mix()`. The
ordinary and latent-class entry points share the prior encoder and the common
fit-only Stan modules.

# Statistical definitions

For an ordinary coefficient (	heta_j), the Normal, Student-(t), and Laplace
declarations mean, respectively,

[
	heta_j\sim N(\mu_j,s_j^2),\qquad
	heta_j\sim t_\nu(\mu_j,s_j),\qquad
	heta_j\sim \operatorname{Laplace}(\mu_j,s_j).
]

The Laplace scale is its exponential-decay scale, so its standard deviation is
(sqrt{2}s_j). Student-(t) degrees of freedom are fixed for the complete
block. They are not estimated. A scalar `mu` or `scale` is repeated over the
block. A non-scalar vector must contain exactly one value per coefficient;
partial recycling is rejected. Numeric vectors and lists of scalar numbers
are equivalent and retain their supplied order.

The regularised horseshoe follows the hierarchy used by brms and rstanarm. For
raw (z_j\sim N(0,1)), positive local scale (lambda_j), positive global
scale (	au), and finite slab variance (c^2), JoiNMe uses

[
\widetilde\lambda_j
=
\sqrt{\frac{c^2\lambda_j^2}
{c^2+\tau^2\lambda_j^2}},
\qquad
\theta_j=\mu_j+s_j\tau\widetilde\lambda_jz_j.
]

The local and global scales have half-Student-(t) distributions. The random
slab multiplier has the inverse-gamma representation corresponding to the
declared `slab_df`; `slab_scale` gives the finite slab width. This hierarchy
shrinks small coefficients strongly whilst preventing the very large tails of
an unregularised horseshoe from dominating weakly identified joint models.

# Meaning of each block

| `jm_prior()` argument | Scientific quantity | Vector order |
|---|---|---|
| `beta` | longitudinal population fixed effects | columns of the longitudinal fixed-effect model matrix |
| `alpha` | coefficients multiplying transformed longitudinal summaries in the log hazard | `cv_total`, `cs_total`, `cv_mean`, `cs_mean`, `cv_marker`, `cs_marker`, then packed `corr`, then packed `vcov` |
| `iota` | fitted intercepts and slopes inside functional association transforms | channel order `cv`, `cs`, `corr`, `vcov`, `cv_mean`, `cv_marker`, `cs_mean`, `cs_marker`; intercept precedes slope within each channel |
| `marker` | standardised marker-level random effects before the marker covariance factor | marker-major order, then random-effect coordinate |
| `marker_weight` | signed perturbations added to supplied base marker weights | weight-set order, then marker order |
| `vcov_sd` | intercepts and observed-covariate slopes for subject-specific standard deviations | all (Q) intercepts, then the (Q\times K_{sd}) slope matrix in row-major order |
| `vcov_corr` | intercepts and observed-covariate slopes for off-diagonal partial correlations | all (Q(Q-1)/2) intercepts in `(2,1), (3,1), (3,2), ...` order, then the corresponding slope matrix in row-major order |
| `class_regression` | multinomial-logit coefficients introduced by `formulaClass` | all subject-domain coefficients, then all marker-domain coefficients |
| `lkj` | Cholesky correlation factors for subject and marker random-effect blocks | one concentration shared by the fitted correlation matrices |

All six scalar `alpha` positions exist in the Stan parameter block so the
ordering remains stable across association specifications. Inactive
coefficients do not enter the likelihood. The `corr` and `vcov` tails have the
dimensions established by the covariance design. Likewise, only requested
functional intercepts and slopes contribute positions to `iota`. A practical
workflow is to construct the data once with `fit = FALSE`, inspect the model
matrix and transformation metadata, and then supply coefficient-wise vectors.

The declared `alpha` location and scale describe the coefficient used on the
log-hazard scale. No second random association-scale parameter is multiplied
into it. Where estimated marker weights require a sign convention, the
relevant coefficient is folded to the required sign after its prior
transformation.

# Covariance-regression priors

`vcov_sd` and `vcov_corr` are separate because the positive marginal scale and
the dependence structure answer different scientific questions and may have
different design dimensions. Each accepts Normal, fixed-degrees-of-freedom
Student-(t), Laplace, or regularised-horseshoe shrinkage:

```{r}
covariance_priors <- jm_prior(
  vcov_sd = prior_normal(
    mu = 0,
    scale = c(1, 1, 0.5, 0.5)
  ),
  vcov_corr = prior_horseshoe(
    global_scale = 0.25,
    slab_df = 4,
    slab_scale = 1
  )
)
```

The coefficient dimension is known only after both `formulaVCov` model
matrices have been constructed. A scalar location or scale is repeated over
its block. A non-scalar vector must have exactly the documented packed length;
partial recycling is rejected. The non-negative `lambda_L` values are residual
heterogeneity hyperparameters, not observed-covariate coefficients, and retain
their identified half-Student-$t_6(0,1)$ prior.

# Standardised marker blocks

`marker` and `marker_weight` are deliberately different from `beta`, `alpha`,
and `iota`. Their location is always zero and their ordinary scale is always
one. For example, these are valid:

```{r}
jm_prior(
  marker = prior_laplace(),
  marker_weight = prior_student_t(df = 4)
)
```

These are rejected:

```{r}
jm_prior(marker = prior_normal(scale = 2))
jm_prior(marker_weight = prior_laplace(mu = 0.5))
```

The restriction preserves two identifications. First, the marker covariance
factor carries the scale of realised marker random effects, so a second freely
chosen latent scale would be redundant. Secondly, marker weights are fitted as
unit-scale perturbations around the user-supplied base weights. The family may
change their tail behaviour, but not the unit in which the base weights are
defined. `marker` is the sole public name for this prior block.

# Class-membership regression

`class_regression` accepts the same Normal, fixed-degrees-of-freedom Student-t,
Laplace, and regularised horseshoe declarations as `beta`, `alpha`, and `iota`.
The prior is materialised after `formulaClass` has constructed both allocation-
domain designs. A scalar location or scale is repeated; a vector must match the
combined number of subject-domain and marker-domain coefficients exactly.

```{r}
jm_prior(
  class_regression = prior_student_t(
    df = 4,
    mu = c(0, 0),
    scale = c(1, 0.5)
  )
)
```

For compatibility, `class_regression = 1` continues to mean
`prior_normal(mu = 0, scale = 1)`. It does not select a Student-t prior.

# LKJ correlation prior

`prior_lkj(eta)` is separate because an LKJ prior is a distribution on a
correlation matrix, not a scalar shrinkage family. `eta = 1` is uniform over
correlation matrices. Values greater than one increasingly favour correlations
near zero. The same declared concentration is passed to each fitted Cholesky
correlation factor.

# Relationship to latent classes

In `joinme_mix()`, `shrinkage` still selects the Student-(t_6), Laplace, or
Normal *component density* for selected latent-progress coordinates. It no
longer acts as a global prior switch. The ordinary prior for a selected marker
coordinate is removed before the mixture density is added, so choosing
`marker = prior_laplace()` does not accidentally retain both a Laplace prior
and a mixture prior. Subject and covariance-regression latent coordinates have
their established standard-Normal ordinary densities. Priors for `beta`,
`alpha`, and `iota` remain independent of the class component family.

For a regularised-horseshoe marker declaration, the latent-class mixture acts
on the raw standardised coordinate and the local/global horseshoe scale maps
that coordinate into the realised marker effect. This ordering matches the
non-centred longitudinal predictor and is stated explicitly because both the
mixture scale and the random-effect covariance scale can otherwise appear to
describe the same variation.

# Prior calibration and reporting

Locations and scales should be chosen on the scale where the coefficient enters
the model. Standardising continuous covariates makes this substantially easier.
Association coefficients are log-hazard coefficients after transformation, so
a prior can be calibrated by translating plausible values into hazard ratios.
A horseshoe is most useful when a block is genuinely sparse and contains enough
coefficients for global--local shrinkage to be identified; it is rarely helpful
for a two-coefficient block.

A defensible analysis should report the family and all fixed hyperparameters
for every non-default block, the LKJ concentration, and a sensitivity analysis
for scientifically influential coefficients. Prior predictive simulation
should precede long posterior runs. In a latent-class analysis, inspect class
separation and energy diagnostics as well as ordinary coefficient diagnostics,
because a broad component scale and a broad random-effect scale may produce a
weakly identified posterior ridge even when each prior is proper.

# Implementation map

R validates declarations in `R/specs.R`, encodes them in `R/family.R`, and
materialises scalar or coefficient-wise values in `R/standata.R` only after all
formula dimensions are known. Stan data declarations live in
`helper/data/fit_data.stan`. The reusable distribution transformation and
density functions are in `helper/functions/prior_families.stanfunctions`.
Fit-only parameters, transformations, and densities remain split across the
existing parameter, transformed-parameter, and model include files. Both
ordinary and latent-class fit entry points include those same modules.

# Related implementations

The regularised-horseshoe parameterisation and the decision to keep correlation
priors separate follow the public interfaces documented by
[brms](https://paulbuerkner.com/brms/reference/horseshoe.html) and
[rstanarm](https://mc-stan.org/rstanarm/reference/priors.html). JoiNMe retains
its own explicit block names because its nested marker and association
structures do not map one-to-one to the coefficient classes in either package.
