---
title: "A Scientifically Rigorous Workflow for Joint Modeling"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
# Ensure the package is available for the vignette to run if possible
repo_root <- normalizePath(if (dir.exists("R")) "." else "..")
if (requireNamespace("devtools", quietly = TRUE) && file.exists(file.path(repo_root, "DESCRIPTION"))) {
  devtools::load_all(repo_root, quiet = TRUE)
} else if (requireNamespace("joinme", quietly = TRUE)) {
  library(joinme)
}
library(dplyr)
library(ggplot2)
library(survival)
library(bayesplot)
library(posterior)
```

# 1. Introduction

Applying a joint model in a scientific publication requires more than just fitting the model. It requires a rigorous validation process to ensure the results are robust, converged, and meaningful. This vignette outlines a "best practice" workflow for using `joinme` in a research context, suitable for inclusion in high-impact statistical or clinical journals.

We will walk through a complete end-to-end simulation study:
1.  Generating synthetic data with known ground truth.
2.  Fitting the joint model.
3.  Assessing convergence and model fit.
4.  Performing dynamic predictions for individual subjects.

## 1.1 Program flow in practice

The `joinme` workflow follows a fixed sequence that mirrors the Stan program:

1. **Data assembly (R)**: build design matrices, scale time to $[0, 1]$,
  and prepare Gauss-Kronrod nodes for survival integration.
2. **Parameter declarations (Stan)**: define fixed effects, random effects,
  distributional regression coefficients, baseline hazard parameters, and
  association coefficients.
3. **Transformed parameters (Stan)**: rescale time-related coefficients and
  construct the random effects used in the likelihood.
4. **Model block (Stan)**: apply priors and evaluate the likelihood using
  `reduce_sum`, which splits subjects across threads when requested.
5. **Generated quantities (Stan)**: compute per-observation log-likelihoods,
  fitted values, and prediction summaries for posterior diagnostics.

This sequence is important because it clarifies where each modeling choice
enters the inference: formula parsing happens in R, while all probability
calculations occur in Stan.

# 2. Simulation and Data Exploration

Before analyzing real data, it is highly recommended to perform a **parameter recovery simulation**. This confirms that the model is identifiable given your sample size and sampling frequency.

We use the built-in generalized simulator `simulate_joinme` to generate a dataset where the hazard depends on the "Total Current Value" of the biomarkers.

```{r simulate}
set.seed(2025)

# Simulate a dataset: 
# - 100 subjects
# - 3 biomarkers
# - ~10 observations per marker per subject
sim <- simulate_joinme(
  n_id = 40,
  families = rep("student_t", 3),
  n_obs_per_marker_per_id = 6,
  times_obs = seq(0, 8, length.out = 12),
  beta_long = c("(Intercept)" = 1.0, "time" = 0.5, "x1" = 0.4),
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.5)
)

# When estimating marker weights, compare against sim$truth$marker_weights
# because signed marker weights are used directly in Stan.

dataLong <- sim$dataLong
dataEvent <- sim$dataEvent

# Preview the data
head(dataLong)
head(dataEvent)
```

**Scientific Check**: Always visualize your longitudinal trajectories and survival distribution before modeling.

```{r explore-plot}
# Visualize trajectories for a subset of IDs
subset_ids <- base::sample(unique(dataLong$id), 6)

ggplot(dataLong %>% filter(id %in% subset_ids), aes(x = time, y = y, color = marker, group = marker)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~id) + 
  theme_bw() +
  labs(title = "Longitudinal Trajectories (Subset)", x = "Time", y = "Biomarker Value")

# Check event rate
table(dataEvent$event)
prop.table(table(dataEvent$event))
```

Ensure the event rate is sufficient (typically >10-20 events per parameter in the survival submodel) to support the complexity of the proposed association structure.

# 3. Model Specification

We specify a joint model where:
1.  **Longitudinal Submodel**: Linear growth over time with random intercepts and slopes for each subject-marker combination.
2.  **Survival Submodel**: The hazard depends on baseline covariates (`x1`, `x2`) and the current value (`cv_total`) of the biomarkers.

### Families and distributional regression

Each marker can follow a distinct family (for example Gaussian, Student-$t$,
Poisson, negative binomial, Bernoulli, beta, or ordinal). When a family includes
additional parameters (for example $\sigma$, $\nu$, or $\phi$), those parameters
can be modeled through `formulaDist`. This allows heteroscedasticity or
covariate-dependent dispersion while keeping the mean structure aligned across
markers.

### Association transformations

The association terms can be transformed using one of four modes: identity,
functional bytecode expressions, monotone I-splines, or piecewise linear
interpolation. These transformations are specified in R and passed to Stan as
data so that the likelihood remains deterministic and reproducible.

## 3.1 Marker-only random effects (no inner id term)

In some studies, you may want **marker-level random effects only**, without subject-specific marker deviations. This is now supported by omitting the inner `( ... | id )` term inside the marker block. In this configuration, marker-by-id random effects are disabled (`Q_idm = 0`), and associations that require those effects (notably `corr`) are not permitted.

```{r marker-only-example, eval=FALSE}
# Marker-only random effects: no inner ( ... | id )
formulaLong_marker_only <- y ~ time + x1 +
  (1 + time || id) +
  (0 + x1 || marker)

# Note: assoc = c("corr") is not allowed when Q_idm = 0
```

```{r priors}
# Define explicit priors for transparency (optional but recommended)
# Note: joinme exposes beta (fixed effects), alpha (associations), and lkj
priors <- list(
  beta = list(scale = 2.5),       # Fixed effects (Student-t, df = 6)
  alpha = list(scale = 1.0),      # Association coefficients
  lkj = 2.0                       # Correlation matrices (regularizing)
)
```

# 4. Model Fitting

We fit the model using Hamiltonian Monte Carlo (HMC) via the `cmdstanr` backend (recommended for speed and reliability).

```{r fit}
engine <- if (requireNamespace("rstan", quietly = TRUE)) "rstan" else "cmdstanr"

control <- list(
  engine = engine,
  chains = 1,           # Use 4 chains for publication
  iter_warmup = 200,    # Typically 1000+
  iter_sampling = 200,  # Typically 1000+
  parallel_chains = 1,
  adapt_delta = 0.90,   # Increase if divergences occur
  max_treedepth = 12,
  force_recompile = (engine == "cmdstanr")
)
if (engine == "cmdstanr") control$init <- 0

fit <- joinme(
  dataLong = dataLong,
  dataEvent = dataEvent,
  # Longitudinal formula: Fixed effects + Random effects
  formulaLong = y ~ time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  # Survival formula: Baseline covariates
  formulaEvent = survival::Surv(time, event) ~ x1 + x2,
  # Association structure: Link hazard to the total current value of markers
  assoc = c("cv_total"),
  transforms = list(cv_total = list(type = "identity")),
  priors = priors,
  control = control
)
```

# 5. Convergence Diagnostics

**Do not interpret results without checking convergence.** We use the `posterior` and `bayesplot` packages to inspect the MCMC chains.

## 5.1 R-hat and Effective Sample Size

Values of $\hat{R} < 1.01$ and ESS > 400 indicate reasonable convergence for inference.

```{r diagnostics}
# robust summary of regression coefficients
draws_obj <- if (inherits(fit$fit, "CmdStanMCMC")) {
  fit$fit$draws()
} else {
  posterior::as_draws_array(fit$fit)
}

draws_sub <- posterior::subset_draws(
  draws_obj,
  variable = c("^beta\\[", "^gamma_w\\[", "^alpha"),
  regex = TRUE
)

posterior::summarise_draws(
  draws_sub,
  default_summary_measures(),
  default_convergence_measures()
)
```

## 5.2 Traceplots

Visual inspection of traceplots helps identify mixing issues or stuck chains.

```{r trace}
# Plot traces for the association parameter
bayesplot::mcmc_trace(draws_obj, regex_pars = "alpha")
```

# 6. Posterior Predictive Checks (PPC)

A key step in validation is checking if the model can reproduce the observed data patterns.

```{r ppc, eval=FALSE}
# posterior_predict draws from the fitted model for PPC
# y_rep <- posterior_predict(fit, ndraws = 100)
# bayesplot::ppc_dens_overlay(dataLong$y, y_rep)
```

# 7. Dynamic Prediction

One of the most powerful features of joint models is **dynamic prediction**: updating survival probabilities as new biomarker data becomes available.

If competing risks are present, the dynamic survival and cumulative hazard
outputs correspond to the overall survival that integrates the sum of
cause-specific hazards. Cause-specific survival curves are not returned by the
default prediction outputs.

We select a subject who had an event at a later time and predict their survival probability conditionally on data observed up to an earlier time point.

```{r predict-setup}
# Select a subject with an event > 5
target_id <- dataEvent$id[which(dataEvent$time > 5 & dataEvent$event == 1)[1]]
if (is.na(target_id)) target_id <- dataEvent$id[1]

# Define the conditioning time (time_start)
t_cond <- 4

# You can also pass a column name in newdataEvent, e.g. time_start = "time_start".

# Filter history up to t_cond
ndLong <- dataLong %>% filter(id == target_id, time <= t_cond)
ndEvent <- dataEvent %>% filter(id == target_id)

print(paste("Predicting for ID:", target_id, "conditioned on history up to t =", t_cond))
```

Run the prediction:

```{r predict-run}
# Predict conditional survival from t_cond onwards
preds <- tryCatch(
  predict(
    fit,
    newdataLong = ndLong,
    newdataEvent = ndEvent,
    time_start = t_cond,
    times = seq(t_cond, 8, length.out = 30),
    process = "event" # Focus on survival and cumhaz
  ),
  error = function(e) {
    message("Prediction skipped: ", conditionMessage(e))
    NULL
  }
)

# The result is a list with survival and cumhaz data frames
if (!is.null(preds)) head(preds$survival)
```

Finally, we visualize the predicted survival curve. The shaded area represents the 95% credible interval, accounting for uncertainty in both the parameters and the random effects.

```{r predict-plot}
if (!is.null(preds)) {
  plot(preds) +
    ggtitle(paste("Dynamic Survival Prediction for Subject", target_id)) +
    geom_vline(xintercept = dataEvent$time[dataEvent$id == target_id], linetype = "dashed", color = "red") +
    annotate("text", x = dataEvent$time[dataEvent$id == target_id], y = 0.1, label = "Observed event time", color = "red", hjust = 1.1)
}
```

# 8. Conclusion

By following this workflow:
1.  **Simulate** (or explore) to understand data properties.
2.  **Fit** with explicit priors and robust settings.
3.  **Diagnose** convergence thoroughly using $\hat{R}$ and traceplots.
4.  **Validate** using posterior predictive checks.
5.  **Predict** to demonstrate clinical utility.

You ensure that your joint modeling analysis is scientifically sound and robust against common pitfalls in Bayesian inference.
