---
title: "joinme: Association Construction"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
options(width = 80)
```

# Purpose

This vignette focuses on **association term construction**, including transformation selection, interpretability, and practical trade-offs.

Association transforms apply to all supported longitudinal families (including
double_exponential, skew_double_exponential, beta, and cumulative_logit) and to
competing-risk survival models.

# Functional Transform

```{r functional-transform, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "functional",
    expr = ~ sinh(x) + tanh(x)
  ),
  cs_total = list(
    type = "identity"
  ),
  vcov = list(
    type = "identity"
  )
)
```

**When to use**

- You want a fixed functional form with no spline fitting.
- You need fast reconfiguration without recompiling Stan.

# Monotone I-Spline

```{r monotone-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.5, 1, 1.2, 1.5, 1.6),
    degree = 3
  ),
  cs_total = list(
    type = "identity"
  ),
  vcov = list(
    type = "identity"
  )
)
```

**When to use**

- You want monotone, smooth effects with constrained shape.

# Penalized Monotone I-Spline

```{r penalized-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  vcov = list(
    type = "ispline_penalized",
    x = seq(-2, 2, length.out = 40),
    y = exp(seq(-2, 2, length.out = 40)),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

**When to use**

- You want a monotone transform learned from (x, y) pairs with smoothing.
- Penalization avoids overfitting when knot counts are moderate.

# Piecewise-Linear

```{r piecewise-linear, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  vcov = list(
    type = "pwlin",
    x = c(-2, -1, 0, 1, 2),
    y = c(0.2, 0.5, 1, 0.5, 0.2)
  )
)
```

**When to use**

- You need simple, interpretable transformations.

# Association Diagram

```{mermaid}
%%| fig-cap: "Association components mapped into the survival predictor."
graph TD
  CV[CV total] --> FCV[f_cv]
  CS[CS total] --> FCS[f_cs]
  V[Vcov] --> FVCOV[f_vcov]
  FCV --> ETA[eta assoc]
  FCS --> ETA
  FVCOV --> ETA
```

# Hierarchical Covariance Regression

```{mermaid}
%%| fig-cap: "Covariance regression structure for marker-by-id effects."
graph TD
  X[Covariates] --> LP[Linear predictor]
  UL[Latent effect] --> LP
  LP --> L[Cholesky factor]
  L --> W[Marker-by-id effects]
  Z[Latent z] --> W
```

# Fit with Custom Association

```{r fit-custom-association, eval=FALSE}
fit <- joinme(
  formulaLong = y ~ 1 + time + x1 +
    (1 + time | id) +
    (0 + x1 + (1 + time | id) | marker),
  dataLong = sim$dataLong,
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  dataEvent = sim$dataEvent,
  transforms = transforms,
  assoc = c("cv_total", "vcov")
)
```
