---
title: "Introduction to joinme"
output-dir: ../docs/articles
format:
  html:
    toc: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Installation

```r
# install.packages("remotes")
# remotes::install_github("trinhdhk/joinme")
```

For Stan backends:

```r
# CmdStanR backend (recommended)
# install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
# cmdstanr::install_cmdstan()

# RStan backend (optional)
# install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

## Quick start: simulate → fit → predict → plot

## Conceptual overview

The `joinme` model links two components:

1. A multivariate longitudinal model for repeated biomarker measurements.
2. A survival model whose hazard depends on summaries of the longitudinal
   process at time $t$.

For the longitudinal part, each marker can follow its own distribution
(Gaussian, Student-$t$, Poisson, negative binomial, Bernoulli, beta,
skew families, or ordinal). The linear predictor is decomposed into fixed
effects, subject-level random effects, marker-level random effects, and
optional marker-by-subject random effects. The survival part uses a
proportional hazards model with a flexible spline baseline hazard and
time-dependent association terms.

The association terms implemented in `joinme` are:

- Current value (CV) of the latent linear predictor.
- Current slope (CS) of the latent linear predictor.
- Variance association (VCOV) derived from the marker-by-subject
  covariance structure.

Each association term can be left on its original scale or transformed
using a functional expression, a monotone I-spline, or a piecewise linear
mapping. These transformations are configured in R and passed to Stan as
data so that the likelihood remains fully deterministic.

## Data requirements (minimal)

You need two data frames:

- `dataLong`: long-format longitudinal data with columns for subject id,
  marker id, time, outcome, and covariates.
- `dataEvent`: one row per subject with event/censoring time, event
  indicator, and baseline covariates.

Time is internally scaled to $[0, 1]$ for numerical stability. The model
returns coefficients on the original time scale.

### Step 1: Load packages and set a reproducible seed

```{r intro-setup}
if (requireNamespace("pkgload", quietly = TRUE)) {
  repo_root <- normalizePath("..", mustWork = FALSE)
  if (file.exists(file.path(repo_root, "DESCRIPTION"))) {
    pkgload::load_all(repo_root, quiet = TRUE, compile = FALSE)
  } else {
    library(joinme)
  }
} else {
  library(joinme)
}
library(survival)

set.seed(2026)
```

### Step 2: Simulate a small dataset

We generate a modest dataset so the workflow runs quickly while still
containing both longitudinal and survival information.

```{r intro-simulate}
sim <- simulate_joinme(
  n_id = 10,
  families = rep("student_t", 2),
  n_obs_per_marker_per_id = 4,
  times_obs = seq(0, 5, length.out = 8),
  seed = 2026,
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.6)
)

head(sim$dataLong)
head(sim$dataEvent)
```

### Step 3: Specify longitudinal and survival models

The longitudinal formula defines fixed effects and random effects. The survival
formula defines baseline covariates in the hazard model.

```{r intro-formulas}
formulaLong <- y ~ 1 + time + x1 +
  (1 + time || id) +
  (0 + x1 + (1 + time || id) || marker)

formulaEvent <- survival::Surv(time, event) ~ 1 + x1 + x2
```

### Step 4: Fit the joint model

We link the survival hazard to the current value (CV) of the longitudinal
process using an identity transformation.

```{r intro-fit}
fit <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  families = rep("student_t", 2),
  transforms = list(cv_total = list(type = "identity")),
  control = list(
    engine = "rstan",
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 100,
    iter_sampling = 100,
    seed = 2026,
    refresh = 0,
    adapt_delta = 0.9
  )
)
```

### Step 5: Summarize fixed effects and associations

```{r intro-summaries}
summary(fit)
fixef(fit)
```

### Step 6: Dynamic prediction for one subject

We condition on the observed history up to the last measurement for subject 1
and obtain predictive summaries beyond that time.

```{r intro-predict}
ndL <- sim$dataLong[sim$dataLong$id == 1, ]
ndE <- sim$dataEvent[sim$dataEvent$id == 1, ]
time_start <- max(ndL$time)

pred <- rstantools::posterior_epred(
  fit,
  newdataLong = ndL,
  newdataEvent = ndE,
  time_start = time_start,
  times = seq(time_start, time_start + 1, length.out = 20),
  n_samples = 50,
  seed = 2026
)
```

### Step 7: Plot longitudinal and survival predictions

```{r intro-plot}
plot(pred, which = c("longitudinal", "survival"), combined = TRUE)
```

If `patchwork`/`cowplot` is unavailable, `combined = TRUE` falls back to the
standard per-subject structured return (grouped by subject and outcome) instead
of any flattened helper list.

## Next steps

- See the modeling workflow in `joinme-workflow`.
- Explore association structures in `joinme-associations`.
- Review the statistical details in `joinme-model` and `joinme-theory`.
