---
title: "Worked latent-progress simulation and recovery example"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  eval: false
  warning: false
  message: false
---

# Scientific question

Suppose three unobserved progress patterns differ in subject intercept and
slope and in the subject-specific marker covariance process. Baseline
treatment influences class membership, and the event hazard depends on the
current total longitudinal value. We wish to determine whether
`joinme_mix()` recovers the component structure without being given the true
labels.

# Generative specification

```{r}
library(joinme)

set.seed(5201)

longitudinal_formula <-
  y ~ 1 + time + treatment + age_c +
  (1 + time | id) +
  (0 + treatment + (1 + time | id) | marker)

event_formula <-
  survival::Surv(time, event) ~ treatment + age_c

covariance_formula <- ~ treatment + age_c

class_formula <- ~ treatment + age_c
```

The subject and covariance-regression blocks contribute two selected
coordinates each. Their common packed component locations are:

```{r}
subject_location <- rbind(
  stable  = c(-1.50, -0.70),
  central = c( 0.00,  0.10),
  rapid   = c( 1.60,  0.90)
)

covariance_location <- rbind(
  stable  = c(-0.70, -0.25),
  central = c( 0.00,  0.00),
  rapid   = c( 0.85,  0.30)
)

component_scale <- list(
  subject = rbind(
    c(0.45, 0.50),
    c(0.65, 0.60),
    c(0.50, 0.55)
  ),
  vcov = rbind(
    c(0.55, 0.60),
    c(0.70, 0.70),
    c(0.55, 0.60)
  )
)
```

The baseline probabilities are 0.25, 0.45 and 0.30. Treatment lowers the
first-class logit and raises the second-class logit relative to their baseline
probabilities.

# Simulation

```{r}
simulation <- simulate_joinme_mix(
  formulaLong = longitudinal_formula,
  formulaEvent = event_formula,
  formulaVCov = covariance_formula,
  n_id = 100,
  families = list(
    jm_family("gaussian"),
    jm_family("student_t"),
    jm_family("gaussian")
  ),
  marker_levels = c("inflammation", "renal", "functional"),
  times_obs = seq(0, 6, length.out = 10),
  obs_time_noise_sd = 0.08,
  covariate_formulas = list(
    treatment ~ rbinom(n_id, 1, 0.5),
    age_c ~ rnorm(n_id, 0, 1),
    x1 ~ rnorm(n_id),
    x2 ~ rnorm(n_id)
  ),
  assoc = "cv_total",
  assoc_coefs = c(cv_total = 0.45),
  beta_long = c(
    "(Intercept)" = 1.0,
    "time" = 0.25,
    "treatment" = -0.15,
    "age_c" = 0.10
  ),
  beta_event = c(
    treatment = -0.20,
    age_c = 0.15
  ),
  n_clusters = 3,
  formulaCluster = class_formula,
  cluster_type = c("subject", "vcov"),
  cluster_dimensions = list(
    subject = c(1, 2),
    vcov = c(1, 2)
  ),
  cluster_ordering = "intercept",
  class_parameters = list(
    probability = c(0.25, 0.45, 0.30),
    coefficient = list(
      subject = c(
        "class_1:treatment" = -0.65,
        "class_1:age_c" = -0.20,
        "class_2:treatment" = 0.45,
        "class_2:age_c" = 0.15
      )
    ),
    location = list(
      subject = subject_location,
      vcov = covariance_location
    ),
    scale = component_scale
  ),
  baseline_hazard = list(
    type = "weibull",
    shape = 1.25,
    scale = 7
  ),
  time_cens = 7,
  seed = 5201
)
```

# Inspecting simulated truth

```{r}
mixture_truth <- simulation$truth$mixture

mixture_truth$probability
mixture_truth$dimensions
mixture_truth$starts
mixture_truth$location
mixture_truth$scale
table(mixture_truth$allocation$subject)
```

The probability regression implies person-specific allocation probabilities:

```{r}
head(mixture_truth$probability_by_unit$subject)
summary(
  apply(
    mixture_truth$probability_by_unit$subject,
    1,
    max
  )
)
```

The recorded standardised draws can be compared with their component centres:

```{r}
allocation <- unname(mixture_truth$allocation$subject)

standardised_truth <- data.frame(
  id = names(mixture_truth$allocation$subject),
  class = factor(allocation),
  subject_intercept =
    mixture_truth$standardised_draws$subject[, 1],
  subject_slope =
    mixture_truth$standardised_draws$subject[, 2],
  covariance_coordinate =
    mixture_truth$standardised_draws$vcov[, 1]
)
```

# Fit without true labels

```{r}
fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = simulation$dataLong,
  formulaEvent = event_formula,
  dataEvent = simulation$dataEvent,
  formulaVCov = covariance_formula,
  families = simulation$marker_info$families,
  transforms = simulation$truth$transforms,
  n_clusters = 3,
  formulaCluster = class_formula,
  cluster_type = c("subject", "vcov"),
  cluster_dimensions = list(
    subject = c(1, 2),
    vcov = c(1, 2)
  ),
  cluster_ordering = "intercept",
  priors = jm_prior(
    class_probability = c(2, 2, 2),
    class_regression = 1
  ),
  assoc = "cv_total",
  shrinkage = simulation$truth$shrinkage,
  seed = 5201,
  control = list(
    chains = 2,
    parallel_chains = 2,
    iter_warmup = 2000,
    iter_sampling = 1000,
    adapt_delta = 0.7
  )
)
```

# Computational assessment

```{r}
summary(fit)
diagnosis(fit)
mcmc_plot(fit, variable = "mix_probability", type = "trace")
mcmc_plot(fit, variable = "mix_location", type = "trace")
mcmc_plot(fit, variable = "mix_scale", type = "trace")
```

Assess effective sample sizes and convergence for the ordinary covariance and
association parameters as well as the mixture parameters. A mixture trace can
appear stable even when the underlying covariance regression is weakly
identified.

# Component parameter recovery

```{r}
posterior_summary(
  fit,
  variables = c(
    "mix_probability",
    "mix_location",
    "mix_scale",
    "mix_class_coefficient_subject"
  )
)
```

Under intercept ordering, posterior rows can be compared directly with the
ordered truth. Under no ordering, first align posterior draws or summaries by
permutation.

# Allocation recovery

```{r}
posterior_membership <- posterior_class(fit)
head(posterior_membership$subject)
```

Join posterior summaries to the truth only for evaluation:

```{r}
allocation_assessment <- transform(
  posterior_membership$subject,
  true_class = mixture_truth$allocation$subject[
    match(unit, names(mixture_truth$allocation$subject))
  ]
)
```

Useful summaries include:

```{r}
mean(
  allocation_assessment$class ==
    paste0("class_", allocation_assessment$true_class)
)
```

Hard accuracy should be accompanied by the posterior probability assigned to
the true class and a multiclass logarithmic or Brier score.

# Longitudinal class trajectories

```{r}
longitudinal_plot(
  fit,
  estimand = "mean_per_class"
)

longitudinal_plot(
  fit,
  estimand = "marginal_per_class"
)
```

`mean_per_class` uses fixed effects plus the clustered block's component
centre. `marginal_per_class` integrates the inverse-link response trajectory
over the remaining random-effect distribution. They answer different
questions and need not coincide under nonlinear links.

# Association trajectories

```{r}
posterior_assoc(
  fit,
  times = seq(0, 6, length.out = 80),
  by_class = TRUE
)

plot(
  posterior_assoc(
    fit,
    times = seq(0, 6, length.out = 80),
    by_class = TRUE
  )
)
```

Although the hazard has no explicit class coefficient, class-specific
longitudinal trajectories induce class-specific association trajectories.

# Posterior predictive assessment

```{r}
pp_check(fit)
concordance(fit)
tvAUC(fit, Tstart = 3, Dt = 2)
tvROC(fit, Tstart = 3, Dt = 2)
```

Good allocation recovery does not guarantee good outcome calibration.
Conversely, predictive performance can be adequate when subject class
labels remain uncertain.

# Sensitivity to the fitted class count

```{r}
fits <- lapply(2:4, function(n_clusters) {
  joinme_mix(
    formulaLong = longitudinal_formula,
    dataLong = simulation$dataLong,
    formulaEvent = event_formula,
    dataEvent = simulation$dataEvent,
    formulaVCov = covariance_formula,
    families = simulation$marker_info$families,
    n_clusters = n_clusters,
    formulaCluster = class_formula,
    cluster_type = c("subject", "vcov"),
    cluster_dimensions = list(
      subject = c(1, 2),
      vcov = c(1, 2)
    ),
    cluster_ordering = "intercept",
    assoc = "cv_total",
    seed = 5300 + n_clusters
  )
})
```

Compare out-of-sample predictive measures, posterior predictive checks,
effective class sizes, entropy, component separation and computation. The
known generating count is useful for validation, but applied analyses do not
have this privilege.

# Repeated-replicate study

```{r}
run_replicate <- function(replicate_id) {
  simulated <- simulate_joinme_mix(
    n_id = 240,
    families = rep("gaussian", 3),
    n_clusters = 3,
    cluster_type = c("subject", "vcov"),
    cluster_dimensions = list(
      subject = c(1, 2),
      vcov = c(1, 2)
    ),
    class_parameters = list(
      probability = c(0.25, 0.45, 0.30),
      location = list(
        subject = subject_location,
        vcov = covariance_location
      ),
      scale = component_scale
    ),
    seed = 6000 + replicate_id
  )

  # Fit, diagnose and return compact recovery summaries here.
}
```

Simulation conclusions require repeated datasets. A single recovered example
is a software demonstration, not evidence about repeated-sampling
performance.
