---
title: "Discrimination analysis: concordance, time-dependent ROC curves, and AUC"
subtitle: "Statistical estimands, censoring rules, and public usage"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
```

This vignette gives a transparent account of the discrimination measures
implemented for a model.

The two public functions answer different scientific questions:

- `concordance()` asks whether the fitted **conditional survival curves order
  observed failure times correctly over the available follow-up**. It returns
  one follow-up-wide index.
- `tvROC()` asks how the fitted **conditional event risk at a specified
  horizon separates cases from controls at a chosen landmark** over all
  classification thresholds. `tvAUC()` integrates that receiver operating
  characteristic curve and returns one cumulative/dynamic area for every
  requested landmark--horizon combination.

These are measures of **discrimination**, not calibration. A model may order subjects correctly while systematically overestimating or underestimating their absolute risks. Conversely, well-calibrated group averages do not imply that individual subjects are ordered well.

This distinction is the reason that `concordance()` does not require `time_horizon` or `Dt`, whereas `tvROC()` and `tvAUC()` do. Supplying a horizon changes the scientific question from follow-up-wide concordance tohorizon-specific classification.

## One prediction pathway, two estimands

Both measures begin with the same fitted joint model and the same prospective restriction of each longitudinal history. They separate only after posterior mean conditional survival has been obtained.

```{mermaid}
flowchart LR
    A["Fitted joint model<br/>and evaluation data"] -->
    B["One terminal outcome<br/>per subject"]
    B --> C["Choose landmark and retain<br/>history observed by that time"]
    C --> D["Predict posterior mean<br/>conditional survival"]
    D --> E["Concordance:<br/>compare curves at each<br/>observable event time"]
    D --> F["tvROC/tvAUC:<br/>classify risks at one<br/>specified horizon"]
```

The shared prediction pathway is important: a difference between the two reported numbers arises from their comparison sets and time summaries, not from fitting two different survival models.

# Statistical setting and notation

## Observed event data

For subject $i=1,\ldots,n$, let

- $xT_i^\ast$ be the latent event time;
- $C_i$ be the censoring time;
- $T_i=\min(T_i^\ast,C_i)$ be the observed event or censoring time;
- $\delta_i=1$ indicate an observed event and $\delta_i=0$ indicate
  censoring;
- $K_i$ be the observed event cause when competing events are present;
- $\mathcal H_i(s)$ be the longitudinal history observed through time $s$;
  and
- $\mathbf X_i$ be the event-process covariates.

The fitted event formula may use either a right-censored `Surv(time, status)` response or counting-process `Surv(start, stop, status)` data. Counting-process data can contain several rows for one subject. Discrimination nevertheless concerns one terminal outcome per subject. `joinme` therefore orders the intervals and retains the terminal row before forming any case, control, or comparable-pair indicator. Left- and interval-censored outcomes are not silently reinterpreted; they are rejected because they require different comparison rules.

## Conditional prediction origin

Let $T_{0i}$ denote the time at which prediction is made for subject $i$.
Only longitudinal observations at or before $T_{0i}$ are allowed to inform the prediction. Define residual follow-up by

$$
R_i = T_i-T_{0i}.
$$

The conditional survival function on this residual time scale is

$$
S_i(u\mid T_{0i})
=
\Pr\!\left(
T_i^\ast>T_{0i}+u
\mid
T_i^\ast>T_{0i},
\mathcal H_i(T_{0i}),
\mathbf X_i,
\mathcal D_{\mathrm{fit}}
\right),
\qquad u\geq 0,
$$

where $\mathcal D_{\mathrm{fit}}$ denotes the data used to obtain the posterior distribution. By construction, $S_i(0\mid T_{0i})=1$.

`joinme` calculates survival over posterior draws and uses the posterior mean
ordinate

$$
\widehat S_i(u\mid T_{0i})
=
\operatorname{E}\!\left[
S_i(u\mid T_{0i},\boldsymbol\theta)
\mid \mathcal D_{\mathrm{fit}}
\right]
$$

for discrimination. The present functions therefore report a plug-in discrimination estimate from posterior mean curves. They do not report the posterior distribution, standard error, or confidence interval of the discrimination measure.

## How the joint model enters a survival prediction

For cause $k$, the event model may be written schematically as

$$
\lambda_{ik}(t)
=
\lambda_{0k}(t)
\exp\!\left\{
\mathbf X_i(t)^\mathsf T\boldsymbol\gamma_k
+
\mathcal A_{ik}\!\left(
t,\mathcal H_i(t)
\right)
\right\},
$$

where $\lambda_{0k}(t)$ is the fitted baseline hazard and $\mathcal A_{ik}$ is the complete longitudinal--event associationcontribution. The resulting event-free survival has the form

$$
S_i(t\mid s)
=
\exp\!\left[
-\sum_k\int_s^t \lambda_{ik}(v)\,dv
\right],
\qquad t>s,
$$

with the conditioning adjustment that survival equals one at $s$.

Consequently, the discrimination functions do not reconstruct a simplified risk score. They call `predict.JoiNMeFit()` and inherit:

- the fitted baseline-hazard basis and its original centring constants;
- event covariates and time-varying event-process rows;
- posterior random effects informed by the longitudinal history;
- marker weights;
- current-value, current-slope, functional, correlation, and covariance
  association terms; and
- monotone-spline and ordered piecewise-linear association transforms.

For an ordered piecewise-linear association, the ordered ordinates generated by the simplex and cumulative sum enter the hazard at every quadrature and prediction time. No separate approximation or unconstrained coefficient is introduced by `concordance()`, `tvROC()`, or `tvAUC()`.

# Follow-up-wide survival-curve concordance

## The estimand

Suppose subject $i$ experiences the cause of interest at residual time $R_i$, and subject $j$ is known to remain event-free beyond that residual time. Their observable order is

$$
R_i<R_j.
$$

Following the survival-curve principle proposed by [Antolini, Boracchi, and Biganzoli (2005)](https://doi.org/10.1002/sim.2427),
`joinme` evaluates **both** conditional survival curves at the earlier observed
event time $R_i$. The pair is:

$$
\begin{aligned}
\text{concordant} \quad&
\text{if}\quad
\widehat S_i(R_i\mid T_{0i})
<
\widehat S_j(R_i\mid T_{0j}),\\
\text{discordant} \quad&
\text{if}\quad
\widehat S_i(R_i\mid T_{0i})
>
\widehat S_j(R_i\mid T_{0j}),\\
\text{a prediction tie} \quad&
\text{if}\quad
\widehat S_i(R_i\mid T_{0i})
=
\widehat S_j(R_i\mid T_{0j}).
\end{aligned}
$$

Lower survival means greater predicted risk. Thus a concordant pair assigns greater risk at $R_i$ to the subject who actually experiences the event at $R_i$.

Let $\mathcal P$ be the set of observable ordered pairs and let $a_i$ be the event-time weight attached to comparisons generated by event $i$.
Define

$$
c_{ij}
=
\mathbb I\!\left\{
\widehat S_i(R_i)<\widehat S_j(R_i)
\right\}
+
\frac{1}{2}
\mathbb I\!\left\{
\widehat S_i(R_i)=\widehat S_j(R_i)
\right\}.
$$

The reported concordance is

$$
\widehat C
=
\frac{
\displaystyle
\sum_{(i,j)\in\mathcal P}a_i c_{ij}
}{
\displaystyle
\sum_{(i,j)\in\mathcal P}a_i
}.
$$

A value of $0.5$ represents chance ordering under continuous predictions, a value approaching $1$ represents increasingly correct ordering, and a value below $0.5$ indicates systematic reversal. These descriptions concern ranking only and do not establish calibration or clinical usefulness.

## Comparable pairs and censoring

`joinme` follows the risk-set convention described for
[`survival::concordance()`](https://stat.ethz.ch/R-manual/R-devel/library/survival/html/concordance.html).
For an event of interest at residual time $r$:

| Outcome for comparator $j$ | Comparable? | Reason |
|---|---:|---|
| Event or censoring after $r$ | Yes | Subject $j$ is known to survive beyond the event time |
| Censoring exactly at $r$ | Yes | Under the event-before-censor convention, $j$ is known to be event-free through $r$ |
| Censoring before $r$ | No | The order between the event and $j$'s latent event time is unknown |
| Event of interest exactly at $r$ | No | The two failure times are tied and have no observed strict order |
| Competing event before $r$ | No | It is treated as cause-specific censoring before the comparison time |
| Competing event exactly at $r$ | Yes | It follows the same event-before-censor convention as other censoring at $r$ |

This pairwise rule uses a prematurely censored subject whenever its observedhistory is sufficient for a particular earlier comparison, and excludes it after its censoring time. It neither declares the subject a late survivor nor
imputes an unobserved failure order.

The interpretation relies on the censoring assumptions appropriate to the chosen weighting method. Informative loss to follow-up that remains unexplained by the fitted data can bias discrimination estimates. No comparison rule alone can identify the ordering of two latent event times after observation has ceased.

## Event-time weighting

The `type_weights` argument selects an event-time weighting schedule. `joinme` asks `survival::concordance(..., ranks = TRUE)` for its documented event-time weights and divides each event weight by the number at risk at that event time. This produces the pair weight $a_i$ used in the survival-curve comparisons.

| `type_weights` | Interpretation in `joinme` |
|---|---|
| `"none"` or `"n"` | Harrell-style pair weighting; every observable pair has weight one |
| `"S"` | Weight events using the estimated survival schedule |
| `"S/G"` | Weight events using the survival and censoring schedules |
| `"n/G2"` | Uno-style inverse-censoring event weighting supplied by `survival` |
| `"I"` | Equal event-time weighting supplied by `survival` |

For `"n"`, the event-time weight returned by `survival` equals the risk-set
size, so division by that size gives unit weight to every pair. The other
options change the relative contribution of event times.

The `"n/G2"` option reuses the event-time weighting associated with Uno's
concordance in `survival`; the subject scores in `joinme` remain the
time-dependent survival-curve comparisons defined above. The resulting
estimand is an Uno-style censoring-weighted Antolini comparison rather than a
conventional static-score Uno estimator.

## Choice of prediction origin

By default, `concordance(fit)` chooses

$$
T_{0i}
=
\max\{t<T_i: Y_i(t)\text{ was observed}\}.
$$

The strict inequality prevents a measurement recorded at the outcome time from informing a prediction assessed against that same outcome. Subjects without a valid pre-outcome longitudinal history cannot contribute.

This default answers a residual-prognosis question: **given each subject's most recent pre-outcome information, does the model order their remaining follow-up?** Because $T_{0i}$ may differ between subjects, residual timesrather than absolute study times are compared.

A common landmark gives a different and often more directly comparable
estimand:

```{r common-landmark}
concordance(fit, time_start = 2)
```

This asks whether predictions made at study time 2 order subsequent even times among subjects still under observation after time 2. The landmark may lso be a named vector or the name of a subject-constant column in the event
data:

```{r subject-specific-landmark}
subject_landmark <- c("1" = 1.5, "2" = 2.0, "3" = 1.0)

concordance(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = subject_landmark
)

concordance(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = "clinical_landmark"
)
```

Subject-specific origins define remaining time from those origins, which is
distinct from concordance on a common absolute study-time scale.

## Cause-specific analysis and competing events

The `cause` argument identifies the failure type that generates comparable
event pairs. Other event causes are treated as censoring at their observed times. They may serve as comparators before their competing event, but not after it.

**Caveat** The current `predict()` result used as the score is the fitted event-free `Survival`, which reflects the complete fitted event process. Consequently, with competing risks, the pair construction is cause-specific but its ranking quantity is overall event-free survival. It is not the same estimand as a cause-specific cumulative-incidence concordance. A scientific question about cumulative incidence in the presence of competing risks requires a dedicated competing-risk discrimination estimator.

The public calculation follows the estimand and weighting definitions above. Its returned columns are `concordance`, weighted concordant, discordant and tied amounts, total comparison weight, and the retained numbers of subjects and events. Non-unit weighting means that the pair totals need not be integers.

# Cumulative/dynamic time-dependent ROC analysis

## Landmark population and horizon status

Choose a common landmark $s$ and horizon $t>s$. The relevant population contains subjects known to be event-free at $s$ and having longitudinal history available by $s$. For such a subject, define posterior mean conditional survival and event risk by

$$
\widehat S_i(s,t)
=
\Pr(T_i^\ast>t\mid T_i^\ast>s,\mathcal H_i(s),\mathbf X_i,
\mathcal D_{\mathrm{fit}})
$$

and

$$
\widehat r_i(s,t)=1-\widehat S_i(s,t).
$$

`joinme` follows the cumulative/dynamic convention used by `JMbayes2::tvROC()`. The prediction window excludes the conditioning instant and includes its horizon, $(s,t]$:

$$
D_i(s,t)
=
\begin{cases}
1,
& s<T_i\leq t,\ \delta_i=1,\ K_i=\texttt{cause},\\
0,
& T_i>t,\\
\text{unknown},
& s<T_i\leq t\ \text{without an observed event of the requested cause}.
\end{cases}
$$

An observed requested-cause event in the window is a case. A subject known remain under observation without that cause through the horizon is a dynamic control. Administrative censoring or a competing event before the horizon leaves the requested-cause status at the horizon unknown. Outcomes at or before the landmark are excluded because dynamic prediction conditions on survival beyond the landmark.

The boundary convention is important at tied recorded times. A edvent recorded exactly at $t$ is a case for the interval ending at $t$, whereas a subject must be observed beyond $t$ to be a control.

## ROC thresholds, sensitivity, and false-positive fraction

The reported curve uses 101 thresholds

$$
c_\ell\in\{0,0.01,\ldots,1\}.
$$

It follows the JMbayes2 representation and applies thresholds to conditional survival. Subject $i$ is classified as a predicted case at threshold $c$ when

$$
A_i(c)=\mathbb I\{\widehat S_i(s,t)<c\}.
$$

This is equivalent to classifying on risk with threshold $1-c$, but using survival gives the same orientation as a JMbayes2 `tvROC` curve. As $c$ increases, both sensitivity and the false-positive fraction are non-decreasing.

Let $w_i^{(1)}$ and $w_i^{(0)}$ be the case and control weights described below. The estimated true-positive and false-positive fractions are

$$
\widehat{\operatorname{TP}}(c)
=
\frac{\sum_i w_i^{(1)}A_i(c)}
{\sum_i w_i^{(1)}}
$$

and

$$
\widehat{\operatorname{FP}}(c)
=
\frac{\sum_i w_i^{(0)}A_i(c)}
{\sum_i w_i^{(0)}}.
$$

The ROC curve is the ordered set
$\{(\widehat{\operatorname{FP}}(c_\ell),
\widehat{\operatorname{TP}}(c_\ell))\}_{\ell=1}^{101}$.

## Model-based treatment of premature censoring

With `type_weights = "model-based"`, observed horizon status is represented
directly and unknown status is replaced by its fitted expectation:

$$
w_i^{(1)}
=
\begin{cases}
1, & D_i=1,\\
0, & D_i=0,\\
\widehat r_i(s,t), & D_i\ \text{unknown},
\end{cases}
\qquad
w_i^{(0)}=1-w_i^{(1)}.
$$

Thus an early-censored subject is not declared a case or a control. Instead, the subject contributes fractionally to both groups according to their predicted event probability at the landmark. This uses the fitted joint modelto integrate over the unobserved horizon status. It is coherent only to the extent that the event model and its conditioning information are adequate for the censoring setting. It does not turn an informative dropout process into independent censoring by assumption.

The same posterior mean risk determines both the uncertain-status weight and the primary threshold classification. This matches the plug-in character of the main ROC curve. Posterior-draw ROC curves retain the status weights and vary the subject classification over posterior survival draws.

## Inverse-probability-of-censoring treatment

With `type_weights = "IPCW"`, subjects of unknown horizon status receive zero case and control weight. Let

$$
G(u\mid s)=\Pr(C_i\geq u\mid C_i>s)
$$

be the conditional probability of remaining uncensored. JoiNMe estimates $G$ by a Kaplan--Meier curve in the landmark risk set. Requested-cause failures are failures rather than censoring events in this auxiliary distribution; administrative censoring and competing events terminate observation of the requested cause.

The weights are

$$
w_i^{(1)}
=
\frac{\mathbb I(D_i=1)}{\widehat G(T_i-\mid s)}
$$

and

$$
w_i^{(0)}
=
\frac{\mathbb I(D_i=0)}{\widehat G(t\mid s)}.
$$

The immediately preceding censoring probability is used for a case so that a censoring jump tied at its event time is not allowed to precede the observed event. IPCW targets the landmark population under conditional independent censoring and positivity. Very small values of $\widehat G$ produce unstable weights when the horizon extends beyond reliable follow-up.

## Posterior-draw and posterior-mean curves

Dynamic prediction retains a survival matrix with posterior draws in rows and prediction times in columns for every subject. At the exact horizon, `joinme` aligns these values into a subject-by-draw event-risk matrix. For draw $m$, the classification is

$$
A_i^{(m)}(c)
=
\mathbb I\{1-r_i^{(m)}(s,t)<c\}.
$$

The `tp` and `fp` matrices in the returned object contain one ROC curve per retained posterior draw. The upper-case `TP` and `FP` vectors are the primary curve calculated from posterior mean subject risks. They are not obtained by averaging the posterior-draw ROC curves. This distinction is retained because it is the data structure expected by `JMbayes2::tvAUC.tvROC()` and its plotting method.

A posterior draw containing a non-finite horizon risk is discarded as a whole curve. A subject with a non-finite posterior mean risk is removed before the case and control totals are formed. If no complete posterior-draw column remains, the posterior mean curve is stored as a single deterministic column.

# Area under the time-dependent ROC curve

## Trapezoidal definition

`tvAUC()` first runs `tvROC()` when supplied with a fitted model and then applies trapezoidal integration through `JMbayes2::tvAUC()`, which evaluates

$$
\widehat{\operatorname{AUC}}(s,t)
=
\sum_{\ell=1}^{100}
\frac{
\widehat{\operatorname{TP}}(c_{\ell+1})
+
\widehat{\operatorname{TP}}(c_{\ell})
}{2}
\left\{
\widehat{\operatorname{FP}}(c_{\ell+1})
-
\widehat{\operatorname{FP}}(c_{\ell})
\right\}.
$$

This is the trapezoidal area under the weighted cumulative/dynamic ROC curve. In the absence of censoring weights and with sufficiently complete threshold coverage, it has the familiar interpretation as the probability that a case has a greater predicted event risk than a control, with half credit to risk ties. Under model-based or IPCW weighting, that probability is taken with respects to the corresponding weighted landmark population.

**Caveat** The area is calculated from `TP` and `FP`, and therefore from posterior mean subject risks. It is not a posterior mean of draw-specific areas and it is not a credible interval. A resampling or posterior procedure would be required for uncertainty in predictive performance.

## Optimal threshold summaries

For descriptive use, `tvROC()` also returns two survival-probability cut-offs:

- `F1score` is the median threshold among all finite maxima of $2\,n_{\mathrm{TP}}/(2n_{\mathrm{TP}}+n_{\mathrm{FN}}+n_{\mathrm{FP}})$;
- `Youden` is the median threshold among all maxima of $\operatorname{TP}-\operatorname{FP}$.

These are data-dependent summaries, not automatically recommended clinical decision thresholds. Their consequences depend on event prevalence, the relative costs of errors, calibration, and the intended intervention.

The public methods use the arguments shown in the examples below. A single landmark returns a `tvROC` object that can be plotted or passed to `tvAUC()`; several landmarks return one separately identified curve and area for each landmark--horizon pair.

# Concordance and horizon-specific ROC answer different questions

| Feature | `concordance()` | `tvROC()` and `tvAUC()` |
|---|---|---|
| Scientific question | Are earlier failures assigned lower survival than subjects observed to survive longer? | Can horizon cases be separated from horizon controls? |
| Time summary | One follow-up-wide index | One curve and area per landmark--horizon pair |
| Prediction used | Survival at each observable event time | Conditional survival at one horizon |
| Comparison unit | Earlier-event subject versus later-observed subject | Weighted case and control classification |
| Premature censoring | Pair remains usable only before censoring | Model-based expected status or IPCW |
| Prediction ties | Half credit in pair counts | Represented by the ROC steps and trapezoidal area |
| Event-time ties | No strict failure order; excluded | Status follows the $(s,t]$ prediction window |
| Weighting | Selectable event-time weighting | `"model-based"` or `"IPCW"` censoring weights |
| Competing event | Cause-specific censoring at its event time | Cause-specific censoring for horizon status |
| Principal output | Concordance and auditable pair totals | ROC ordinates, thresholds, and trapezoidal AUC |

AUC can vary markedly with the chosen horizon even when follow-up-wide
concordance changes little. Concordance combines comparisons made at many
event times and can therefore mask horizon-specific behaviour.

# Reproducible use

## Concordance

```{r concordance-examples}
# Subject-specific origin at the final pre-outcome measurement.
concordance(fit)

# Common study-time landmark and unit pair weights.
concordance(
  fit,
  time_start = 2,
  type_weights = "none",
  predict_control = list(n_samples = 500, n_pred_draws = 100),
  seed = 2026
)

# Uno-style censoring event weights.
concordance(
  fit,
  time_start = 2,
  type_weights = "n/G2",
  predict_control = list(n_samples = 500, n_pred_draws = 100),
  seed = 2026
)
```

Concordance controls posterior extraction and dynamic-prediction draws through `predict_control`. Stability can be checked by repeating the calculation with more posterior and prediction draws while retaining the seed.

## Time-dependent ROC curve and AUC

```{r roc-examples}
# Risk through time 5 among subjects under observation after time 2.
roc_2_5 <- tvROC(
  fit,
  time_start = 2,
  time_horizon = 5,
  type_weights = "model-based",
  n_samples = 500,
  seed = 2026
)

plot(roc_2_5)

# Integrate an already calculated curve without repeating prediction.
area_2_5 <- tvAUC(roc_2_5)

# Or calculate the curve and area from the fitted model in one call.
area_2_5_direct <- tvAUC(
  fit,
  time_start = 2,
  time_horizon = 5,
  type_weights = "IPCW",
  n_samples = 500,
  seed = 2026
)

# A three-year prediction window from several landmarks.
areas <- tvAUC(
  fit,
  time_start = c(0, 1, 2),
  Dt = 3,
  type_weights = "model-based",
  n_samples = 500,
  seed = 2026
)
```

For several landmarks, each row is a separate cumulative/dynamic estimand. To plot them, select the intended compatible curve explicitly, for example
`plot(rocs$curves[[1]])`.

## Evaluation on new subjects

```{r validation-example}
cIndex <- concordance(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = 2,
  predict_control = list(n_samples = 500, n_pred_draws = 100),
  seed = 2026
)

validation_roc <- tvROC(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = 2,
  time_horizon = 5,
  type_weights = "IPCW",
  n_samples = 500,
  seed = 2026
)

validation_auc <- tvAUC(validation_roc)
```

When new data are omitted, the fitting data are used. Held-out evaluation requires both `newdataLong` and `newdataEvent`.

# References

Antolini L, Boracchi P, Biganzoli E (2005). A time-dependent discrimination
index for survival data. *Statistics in Medicine*, 24, 3927--3944.
[doi:10.1002/sim.2427](https://doi.org/10.1002/sim.2427).

Heagerty PJ, Zheng Y (2005). Survival model predictive accuracy and ROC
curves. *Biometrics*, 61, 92--105.
[doi:10.1111/j.0006-341X.2005.030814.x](https://doi.org/10.1111/j.0006-341X.2005.030814.x).

Rizopoulos D (2011). Dynamic predictions and prospective accuracy in joint
models for longitudinal and time-to-event data. *Biometrics*, 67, 819--829.
[Publisher record](https://onlinelibrary.wiley.com/doi/10.1111/j.1541-0420.2010.01546.x).

Rizopoulos D, Papageorgiou G, Miranda Afonso P (2024). `JMbayes2` accuracy
measures. The package reference defines the `tvROC` and `tvAUC` interfaces,
censoring choices, returned structures, and associated methods.
[R documentation](https://search.r-project.org/CRAN/refmans/JMbayes2/html/accuracy.html).

Therneau TM. `survival::concordance`: concordance for survival and regression
models. The official documentation describes risk-set comparability, event
ties, and the available time-weighting rules.
[R documentation](https://stat.ethz.ch/R-manual/R-devel/library/survival/html/concordance.html).

Therneau TM. *Concordance* vignette for the `survival` package. The vignette
gives the computational and censoring conventions used by the event-time
weighting engine.
[Official vignette](https://stat.ethz.ch/R-manual/R-devel/library/survival/doc/concordance.pdf).
