---
title: "A Example Workflow for Joint Modelling with mutlivariate Longitudinal and Survival Data using joinme"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
# Ensure the package is available for the vignette to run if possible
repo_root <- normalizePath(if (dir.exists("R")) "." else "..")
if (!require('joinme', quietly = TRUE)) {
  if (requireNamespace("devtools", quietly = TRUE) && file.exists(file.path(repo_root, "DESCRIPTION"))) {
    devtools::load_all(repo_root, quiet = TRUE)
  } else {
    try(base::attachNamespace("joinme"), silent = TRUE)
  }
}
library(dplyr)
library(ggplot2)
library(survival)
library(bayesplot)
library(posterior)
```


This vignette outlines a workflow for using `joinme` in a research context, suitable for inclusion in high-impact statistical or clinical journals.

# Simulation and Data Exploration

We use the built-in generalised simulator `simulate_joinme` to generate a dataset where the hazard depends on the "Total Current Value" of the biomarkers.

```{r simulate}
set.seed(2025)

# Simulate a dataset: 
# - 100 subjects
# - 3 biomarkers
# - ~10 observations per marker per subject
sim <- simulate_joinme(
  n_id = 40,
  families = list(
    jm_family("student_t"),
    jm_family("student_t"),
    jm_family("student_t", inv_link = ~ inv_logit(x / 2))
  ),
  times_obs = seq(0, 8, length.out = 12),
  quadrature_nodes = 31,
  assoc = c("cv_total"),
  truth = jm_truth(
    longitudinal = c("(Intercept)" = 1.0, "time" = 0.5, "x1" = 0.4),
    assoc_coef = list(slope = c(cv_total = 0.5)),
    marker_weights = list(family = "student_t")
  )
)

dataLong <- sim$dataLong
dataEvent <- sim$dataEvent

# Optional: generate counting-process event rows with delayed entry.
sim_split <- simulate_joinme(
  n_id = 20,
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x1 + x2,
  covariate_formulas = list(
    x1 ~ time_varyring(rnorm, c(2, 5), mean = 0, sd = 1),
    x2 ~ rnorm(n_id)
  ),
  left_truncation_max = 1.0,
  seed = 2026
)
dataEvent_split <- sim_split$dataEvent

# Preview the data
head(dataLong)
head(dataEvent)
```

`jm_truth()` describes the data-generating population; `jm_prior()` describes
the probability distributions used in fitting. They share scientific
component names but are not interchangeable. A numerical value in `jm_truth()`
is fixed for the whole data set; a `prior_*()` declaration is sampled once and
its realised value is stored in `sim$truth`. The truth object also contains
`basehaz`, `assoc_coef`, and `re_params`. If an ordinary random-effect standard
deviation is omitted, it is drawn once from `Exponential(1)`. If its
correlation matrix is omitted, it is drawn once from the LKJ distribution
declared by `lkj = prior_lkj(...)`.

Visualise your longitudinal trajectories and survival distribution before modelling.

```{r explore-plot}
# Visualise trajectories for a subset of IDs
subset_ids <- base::sample(unique(dataLong$id), 6)

ggplot(dataLong %>% filter(id %in% subset_ids), aes(x = time, y = y, color = marker, group = marker)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~id) + 
  theme_bw() +
  labs(title = "Longitudinal Trajectories (Subset)", x = "Time", y = "Biomarker Value")

# Check event rate
table(dataEvent$event)
prop.table(table(dataEvent$event))
```

# Model Specification

We specify a joint model where:
1.  **Longitudinal Submodel**: Linear growth over time with random intercepts and slopes for each subject-marker combination.
2.  **Survival Submodel**: The hazard depends on baseline covariates (`x1`, `x2`) and the current value (`cv_total`) of the biomarkers.

### Families and distributional regression

Each marker can follow a distinct family (for example Gaussian, Student-$t$,
Poisson, negative binomial, Bernoulli, beta, or ordinal). When a family includes
additional parameters (for example $\sigma$, $\nu$, or $\phi$), those parameters
can be modelled through `formulaDist`. This allows heteroscedasticity or
covariate-dependent dispersion while keeping the mean structure aligned across
markers.

### Association transformations

The association terms can be transformed using identity, a user-defined
mathematical expression, monotone I-splines, or ordered piecewise-linear
interpolation.

## Marker-only random effects (no inner id term)

In some studies, marker-level random effects may be required without
subject-specific marker departures. Omit the inner `( ... | id )` term inside
the marker block. Because `corr` and `vcov` describe the covariance of those
subject-specific marker departures, they are unavailable in this model.

```{r marker-only-example, eval=FALSE}
# Marker-only random effects: no inner ( ... | id )
formulaLong_marker_only <- y ~ time + x1 +
  (1 + time || id) +
  (0 + x1 || marker)
```

Prior families are declared by model block through `jm_prior()`. Thus changing
the prior for an association coefficient does not silently change the fixed
effects, affine transformation shifts, marker effects, or marker weights.
```{r priors}
# Define explicit, block-specific priors for transparency.
priors <- jm_prior(
  intercept = prior_student_t(df = 6, mu = 0, scale = 2),
  slope = prior_normal(mu = 0, scale = 1),
  longitudinal = list(
    slope = prior_normal(
      mu = list(0, 0),
      scale = list(1, 0.5)
    )
  ),
  assoc = list(slope = prior_student_t(df = 4, mu = 0, scale = 1)),
  functional = list(slope = prior_laplace(mu = 0, scale = 0.75)),
  marker = list(family = prior_normal()),
  marker_weights = list(
    offset = c(m1 = 0, m2 = 0, m3 = 0),
    intercept = prior_normal(mu = 0, scale = 0.75),
    family = "student_t"
  ),
  lkj = prior_lkj(2)
)
```

The global `intercept` and `slope` declarations are inherited by every
ordinary regression component unless that component replaces the named role.
Numeric vectors and lists of scalar numbers are equivalent. Partial recycling
is refused: a non-scalar declaration must have exactly the dimension of that
component and role after the formula has been parsed. 

When `formulaDist` contains family-scoped regressions, the matching prior may
use the same scope. Because it is an argument name, quote it with backticks:

```{r scoped-distributional-prior}
scoped_priors <- jm_prior(
  `sigma[family='student']` = list(
    intercept = prior_student_t(df = 4, scale = 1),
    slope = prior_normal(scale = 0.25)
  )
)
```

<!-- The alias `student` is canonicalised to `student_t`. A marker selector such as
`` `sigma[marker='m2']` `` is available when `m2` uniquely identifies that
family-scoped coefficient block. If several markers share the same block, use
the family selector because the likelihood contains only one shared
coefficient vector. -->

# Model Fitting

We fit the model using Hamiltonian Monte Carlo (HMC) via Stan. By default, `joinme` prefers `cmdstanr` if compiled, otherwise it falls back to `rstan`. The control parameters can be adjusted using `cmdstanr` syntax.

```{r fit}
has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  !is.null(tryCatch(cmdstanr::cmdstan_version(error_on_NA = FALSE), error = function(e) NULL))
engine <- if (has_cmdstan) "cmdstanr" else "rstan"

control <- list(
  engine = engine,
  chains = 1,           # Use 4 chains for publication
  iter_warmup = 100,    # Typically 1000+
  iter_sampling = 100,  # Typically 1000+
  parallel_chains = 1,
  adapt_delta = 0.8,   # Increase if divergences occur
  max_treedepth = 12
)

fit <- joinme(
  dataLong = dataLong,
  dataEvent = dataEvent,
  # Longitudinal formula: Fixed effects + Random effects
  formulaLong = y ~ time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  # Survival formula: Baseline covariates
  formulaEvent = survival::Surv(time, event) ~ x1 + x2,
  # Association structure: Link hazard to the total current value of markers
  assoc = c("cv_total"),
  transforms = joinme_tf(cv_total = "identity"),
  priors = priors,
  control = control
)

# Counting-process fit with left truncation/time-split covariates:
fit_split <- joinme(
  dataLong = sim_split$dataLong,
  dataEvent = dataEvent_split,
  formulaLong = y ~ time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x1 + x2,
  assoc = c("cv_total"),
  control = control
)
```

# Convergence Diagnostics

We use the `posterior` and `bayesplot` packages to inspect the MCMC chains.

## R-hat and Effective Sample Size

Values of $\hat{R} < 1.01$ and ESS > 400 indicate reasonable convergence for inference, as recommended by Stan developers. 

```{r diagnostics}
# robust summary of regression coefficients on the renamed user-facing scale
draws_obj <- posterior_draws(fit, format = "draws_array")

draws_sub <- posterior::subset_draws(
  draws_obj,
  variable = c("^(Intercept)", "^time$", "^event", "^cv_total"),
  regex = TRUE
)

posterior::summarise_draws(
  draws_sub,
  default_summary_measures(),
  default_convergence_measures()
)
```

## Traceplots

Visual inspection of traceplots helps identify mixing issues or stuck chains. Note that the marker-specific association term for the first association (if marker weights are shared) is forced to be positive. If `marker_weights$shared` is `FALSE`, then the marker-specific association terms are all constrained to be positive.

```{r trace}
# Plot traces for the association parameter using the helper API
mcmc_plot(fit, variable = "cv_total", type = "trace")
```

# Dynamic Prediction

One of the most powerful features of joint models is **dynamic prediction**: updating survival probabilities as new biomarker data becomes available.

If competing risks are present, the dynamic survival and cumulative hazard outputs correspond to the overall survival that integrates the sum of cause-specific hazards. Cause-specific survival curves are not returned by the default prediction outputs.

We select a subject who had an event at a later time and predict their survival probability conditionally on data observed up to an earlier time point.

```{r predict-setup}
# Select a subject with an event > 5
target_id <- dataEvent$id[which(dataEvent$time > 5 & dataEvent$event == 1)[1]]
if (is.na(target_id)) target_id <- dataEvent$id[1]

# Define the conditioning time (time_start)
t_cond <- 4

# You can also pass a column name in newdataEvent, e.g. time_start = "time_start".

# Filter history up to t_cond
ndLong <- dataLong |> filter(id == target_id, time <= t_cond)
ndEvent <- dataEvent |> filter(id == target_id)

print(paste("Predicting for ID:", target_id, "conditioned on history up to t =", t_cond))
```

Run the prediction:

```{r predict-run}
# Predict conditional survival from t_cond onwards
preds <- tryCatch(
  predict(
    fit,
    newdataLong = ndLong,
    newdataEvent = ndEvent,
    time_start = t_cond,
    times = seq(t_cond, 8, length.out = 30),
    process = "event", # Focus on survival and cumhaz
    control = list(progress = FALSE) # just to silence the progress bar
  ),
  error = function(e) {
    message("Prediction skipped: ", conditionMessage(e))
    NULL
  }
)

head(preds$predictions$survival)
```

When the identifier occurred during fitting, the posterior random effects can be reused directly. This targets a fitted-subject conditional prediction and avoids estimating another set of effects from the same history:

```{r predict-fitted-effects, eval=FALSE}
fitted_subject_prediction <- posterior_epred(
  fit,
  newdataLong = ndLong,
  newdataEvent = ndEvent,
  process = c("longitudinal", "event"),
  times = seq(t_cond, 8, length.out = 30),
  reuse_fitted_re = TRUE
)

```

The identifier is matched to the fitted subject index. An unseen identifier is rejected; dynamic prediction for a new subject uses `reuse_fitted_re = FALSE`.`posterior_linpred()` and `posterior_predict()` accept the same argument. 
<!-- For a latent-class fit, fitted allocation probabilities remain paired with the reused effects. -->

Conditional effects and contrasts can use the same subject-conditional estimand by placing the fitted identifier in `conditions`. Both contrast profiles deliberately share that identifier:

```{r conditional-fitted-effects, eval=FALSE}
conditional_effects(
  fit,
  effects = "time",
  conditions = data.frame(id = target_id),
  process = "longitudinal",
  reuse_fitted_re = TRUE
)

conditional_contrast(
  fit,
  groupA = c(treatment = "active"),
  groupB = c(treatment = "control"),
  conditions = data.frame(id = target_id, time = seq(0, 8, length.out = 40)),
  process = "longitudinal",
  reuse_fitted_re = TRUE
)
```

Both conditional interfaces can retain the posterior calculation before it is summarised. This is useful for posterior probabilities, nonlinear functions of a contrast, or bespoke figures whose uncertainty must be propagated draw by draw:

```{r conditional-mcmc-samples, eval=FALSE}
effect_draws <- conditional_effects(
  fit,
  effects = "time",
  process = "longitudinal",
  summary = FALSE
)

time_draws <- effect_draws$longitudinal$time

contrast_draws <- conditional_contrast(
  fit,
  groupA = c(treatment = "active"),
  groupB = c(treatment = "control"),
  conditions = data.frame(time = seq(0, 8, length.out = 40)),
  process = "longitudinal",
  summary = FALSE
)

time_contrast_draws <- contrast_draws$longitudinal
```

`concordance(fit)` is a follow-up-wide survival-curve concordance rather than a ROC analysis at one selected horizon. By default, residual follow-up for subject $i$ starts at their final longitudinal measurement strictly before the observed event or censoring time, $T_{0i}$. For an event at residual time $r_i=T_i-T_{0i}$, the event subject is compared with every subject known to survive longer. The pair is concordant when $\widehat S_i(r_i)<\widehat S_j(r_i)$: both conditional survival curves are evaluated at the same earlier event time. This is the survival-curve concordance proposed by Antolini and colleagues and does not use the outcome time to construct a fixed subject score.

Premature censoring is handled through observability. A subject censored before $r_i$ is not comparable with that event; censoring at $r_i$ establishes survival through that time and remains comparable. `type_weights = "n/G2"` requests Uno's inverse-censoring weighting through `survival::concordance()`. Fitted monotone splines and ordered piecewise-linear associations enter every curve through their posterior ordinates.

<!-- `tvROC(fit, ...)` retains the horizon-specific estimand. At landmark $s$ and horizon $t$, it evaluates sensitivity and the false-positive fraction over 101 thresholds of conditional survival. Cases experience the requested cause in the interval $(s,t]$, while controls are known to remain without that cause through $t$. Premature censoring is handled either bymodel-based expected case/control status or inverse Kaplan--Meier censoring weights. `tvAUC(fit, ...)` first calculates that curve and then uses `JMbayes2::tvAUC.tvROC()` to obtain its trapezoidal area. Dynamic prediction for all methods reuses the exact fitted B-spline, natural-spline, or formula baseline-hazard basis, including its original centring constants and any ordered piecewise-linear association. -->


```{r discrimination-example, eval=FALSE}
concordance(fit)
concordance(fit, time_start = 2, type_weights = "n/G2")
roc <- tvROC(
  fit,
  time_start = 2,
  time_horizon = 5,
  type_weights = "model-based"
)
plot(roc)
tvAUC(roc)
```

Finally, we visualise the predicted survival curve. The shaded area represents the 95% credible interval, accounting for uncertainty in both the parameters and the random effects.

```{r predict-plot}
if (!is.null(preds)) {
  survival_plot(preds) +
    ggtitle(paste("Dynamic Survival Prediction for Subject", target_id)) +
    geom_vline(xintercept = dataEvent$time[dataEvent$id == target_id], linetype = "dashed", color = "red") +
    annotate("text", x = dataEvent$time[dataEvent$id == target_id], y = 0.1, label = "Observed event time", color = "red", hjust = 1.1)
}
```

# Posterior draws and visualisation

`posterior_draws()` and `mcmc_plot()` use the same parameter names as `summary(fit)`.

```{r posterior-reporting}
renamed_draws <- posterior_draws(fit, variables = c("time", "cv_total"), format = "draws_df")
head(renamed_draws)

mcmc_plot(fit, variable = c("time", "cv_total"), type = "intervals")
```

The plotting wrappers expose the same fitted and predicted views as the main `plot()` methods while making the intent explicit in scripts and reports:

```{r plot-wrappers, eval=FALSE}
longitudinal_plot(fit, subject = 1, scale = "epred")
association_plot(fit)
diagnostic_plot(fit, type = "ess_bulk")

if (!is.null(preds)) {
  longitudinal_plot(preds)
  survival_plot(preds)
  cumhaz_plot(preds)
}
```

For fitted longitudinal displays, the heatmap is an alternative to the usual separated marker curves:

```{r predict-heatmap, eval=FALSE}
plot(
  fit,
  type = "longitudinal",
  longitudinal_style = "heatmap",
  scale = "epred",
  threshold = 0.05
)
```

`plot(fit, ...)` uses fitted posterior samples directly and does not accept a conditioning argument. Named covariate profiles belong to the conditional- effects estimand. Build those profiles with `make_conditions()` and pass them to `conditional_effects()`:

```{r predict-conditions, eval=FALSE}
profiles <- make_conditions(dataEvent, vars = c("x1", "x2"))

effects_data <- conditional_effects(
  fit,
  effects = list(
    longitudinal = "time",
    event = c("x1", "x2")
  ),
  conditions = profiles,
  process = c("longitudinal", "event"),
  longitudinal_estimand = "population",
  plot = FALSE
)

plot(effects_data, ask = FALSE)
```

The longitudinal calculation has three estimands. `"population"` uses only the fixed-effect contribution, evaluates that contribution under every selected marker's inverse link, and averages the resulting trajectories within each posterior draw. `"marker"` adds the fitted marker-level deviation and retains a separate trajectory for every selected marker, whilst still excluding subject and marker-by-subject deviations. `"marginal_marker"` adds the marker-level deviation, calculates each selected marker response, and averages those responses within every posterior draw. Population and marker-marginal results therefore contain one trajectory per condition; only `"marker"` produces marker-specific curves. Applying the inverse link before averaging is important for nonlinear links: it estimates the average expected response rather than the expected response at an average marker effect. The `markers` argument restricts the equally weighted set entering either average, or the panels retained by the marker-specific estimand.

```{r predict-conditions-marker-average, eval=FALSE}
average_marker_effect <- conditional_effects(
  fit,
  effects = "time",
  process = "longitudinal",
  longitudinal_estimand = "marginal_marker",
  markers = c("m1", "m2"),
  plot = FALSE
)
```

The event tables contain the covariate-specific relative hazard
`exp(W * gamma)`, holding the baseline hazard and longitudinal association contribution fixed.

A paired comparison of two covariate profiles is obtained with
`conditional_contrast()`. This differs from subtracting two rows of a printed
conditional-effects table: predictions for A and B are paired within each MCMC
draw, the contrast is formed, and only that posterior contrast is summarised.

```{r predict-conditional-contrast, eval=FALSE}
contrast_conditions <- data.frame(
  time = seq(0, max(dataLong$time), length.out = 60),
  x2 = 0,
  cond__ = paste0("time=", round(seq(0, max(dataLong$time), length.out = 60), 2))
)

profile_contrast <- conditional_contrast(
  fit,
  groupA = list(x1 = 1),
  groupB = list(x1 = -1),
  conditions = contrast_conditions,
  process = c("longitudinal", "event"),
  method = "posterior_epred",
  longitudinal_estimand = "marginal_marker",
  event_scale = "hazard_ratio",
  plot = FALSE
)

plot(profile_contrast, condition_variable = "time", ask = FALSE)
```

Longitudinal contrasts are differences, A minus B, on the selected scale. Event contrasts are hazard ratios, A divided by B, unless the log-hazard-ratio scale is requested. Group assignments override matching columns in `conditions`; all remaining predictors use the condition value or their reference value from the fitting data.
