---
title: "Independent standard-deviation and correlation regression"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

# Scientific purpose

Marker-by-subject random effects describe departures specific to one subject
and one longitudinal marker. Their covariance need not be constant across
subjects. A treatment may alter the variability of these departures without
altering their dependence, whilst a baseline physiological measurement may
alter dependence without changing marginal variability. A single covariance
formula cannot express these hypotheses independently.

`joinme` therefore separates covariance regression into:

1. a standard-deviation regression, declared by `formulaVCov$sd`; and
2. an off-diagonal unit-lower-triangular factor regression, declared by `formulaVCov$corr`.

# Formula interface

The explicit form is:

```{r}
covariance_formula <- list(
  sd = ~ treatment + age,
  corr = ~ baseline_score
)
```

A single formula is an exact shorthand for two identical formulae:

```{r}
formulaVCov = ~ treatment + age

# exactly equivalent to
formulaVCov = list(
  sd = ~ treatment + age,
  corr = ~ treatment + age
)
```

Both named components must be present when a list is used. Random-effect and the longitudinal time variable are excluded because covariance predictors are evaluated once per subject. 

# Statistical model

Let $Q$ be the dimension of the marker-by-subject random-effect basis. For
subject $i$, `joinme` reconstructs

$$
L_i=\operatorname{diag}(\boldsymbol\sigma_i)K_i,
$$

where $\boldsymbol{\sigma}_i=(\sigma_{i1},\ldots,\sigma_{iQ})^\top$ contains marginal standard deviations and $K_i$ is a Cholesky correlation factor. The marker-by-subject effect for marker $d$ is

$$
\boldsymbol w_{id}=L_i\boldsymbol z_{w,id},
\qquad
\boldsymbol z_{w,id}\sim N_Q(\boldsymbol0,I_Q).
$$

## Standard deviations

Let $\boldsymbol x_{sd,i}$ be the design row produced from `formulaVCov$sd`.
For coordinate $r=1,\ldots,Q$,

$$
\eta^{sd}_{ir} =
\alpha^{sd}_r
+\boldsymbol x_{sd,i}^{\top}\boldsymbol\beta^{sd}_r
+\lambda_{m(r)}z_{L,i,m(r)}.
$$

The marginal standard deviation is

$$
\sigma_{ir}=g_{sd}(\eta^{sd}_{ir}),
$$

where $g_{sd}$ is softplus by default and may be changed to the exponential
link with `control = list(vcov_diag_link = "exp")`.

## Off-diagonal dependence

Let $\boldsymbol x_{corr,i}$ be the independent design row produced from
`formulaVCov$corr`. For row-major off-diagonal coordinate $h=(r,c)$, $c<r$,

$$
\eta^{K}_{ih}
=
\alpha^{K}_h
+\boldsymbol x_{corr,i}^{\top}\boldsymbol\beta^{K}_h
+\lambda_{m(h)}z_{L,i,m(h)},
\qquad
p_{i,rc}=\tanh(\eta^{K}_{ih}),\ c<r.
$$

These partial correlations form one row of the Cholesky correlation factor by

$$
K_{i,rc}=p_{i,rc}\prod_{h<c}\sqrt{1-p_{i,rh}^2},\quad c<r,
\qquad
K_{i,rr}=\prod_{h<r}\sqrt{1-p_{i,rh}^2}.
$$

The upper triangle is zero and every row has norm one.

The covariance matrix is then

$$
\Sigma_i=L_iL_i^{\top}
=\operatorname{diag}(\boldsymbol\sigma_i)K_iK_i^{\top}\operatorname{diag}(\boldsymbol\sigma_i).
$$

This is the independent regression design: the SD model sets
$\boldsymbol\sigma_i$, and the dependence model sets row-wise partial
correlations. The off-diagonal regression is absent when the nested
marker-by-subject term uses `||`, signalling an iid random effect structure.

## Meaning of the latent loading

There is one non-negative scalar $\lambda_m$ for each covariance
coordinate. Collectively, $\boldsymbol\lambda_L$ is a vector, not a dense
matrix. Its action is

$$
\operatorname{diag}(\boldsymbol\lambda_L)\boldsymbol z_{L,i},
$$

so latent coordinate $m$ affects only predictor coordinate $m$. In an
ordinary model, $\lambda_m^2$ is the conditional residual variance of that
untransformed predictor. It is not itself a final standard deviation,
correlation, or covariance.

# Priors

Standard-deviation and correlation regressions can have independent priors:

```{r}
covariance_priors <- jm_prior(
  vcov = list(
    sd = list(
      intercept = prior_normal(mu = 0, scale = 1),
      slope = prior_normal(mu = 0, scale = 0.5)
    ),
    corr = list(slope = prior_horseshoe(
      global_scale = 0.25,
      slab_df = 4,
      slab_scale = 1
    ))
  )
)
```

Normal, fixed-degrees-of-freedom Student-$t$, Laplace, and regularised
horseshoe families are available. A scalar location or scale applies across
its coefficient block; a vector must match that block exactly. The positive
`lambda_L` loadings retain a half-Student-$t_6$ prior
because they describe unexplained heterogeneity rather than observed-
covariate regression coefficients.

# Fitting

```{r}
fitted <- joinme(
  formulaLong = y ~ treatment * time +
    (1 + time | id) +
    (1 + time + (1 + time | id) | marker),
  dataLong = data_long,
  formulaEvent = survival::Surv(event_time, event) ~ treatment,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  priors = covariance_priors,
  families = marker_families,
  assoc = "vcov"
)
```

`summary(fitted)` labels SD and correlation slopes separately using their own
model-matrix column names. Dynamic prediction evaluates both formulae for the
new subject and applies the same unit-lower-triangular factor reconstruction.

The latent-progress entry point uses precisely the same covariance syntax:

```{r}
mixture_fit <- joinme_mix(
  formulaLong = y ~ treatment * time +
    (1 + time | id) +
    (1 + time + (1 + time | id) | marker),
  dataLong = data_long,
  formulaEvent = survival::Surv(event_time, event) ~ treatment,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  priors = covariance_priors,
  n_classes = 3,
  formulaClass = ~ treatment,
  class_type = "vcov"
)
```

The mixture acts on selected standardised covariance coordinates before the SD or
correlation link. It does not form a Cartesian product of class allocations
for the two regressions. `class_type = "corr"` selects off-diagonal
coordinates; `class_type = "vcov"` may select diagonal and off-diagonal
coordinates.

# Exact simulation-to-fit translation

The simulator uses the same formula object and component names:

```{r}
simulation <- simulate_joinme(
  formulaLong = y ~ treatment * time +
    (1 + time | id) +
    (1 + time + (1 + time | id) | marker),
  formulaEvent = survival::Surv(event_time, event) ~ treatment,
  formulaVCov = covariance_formula,
  truth = jm_truth(
    vcov = list(
      sd = list(
        intercept = c(-0.2, -0.4),
        slope = c(0.35, 0.20, -0.10, 0.05),
        latent = c(0.25, 0.20)
      ),
      corr = list(intercept = 0.15, slope = -0.30, latent = 0.20)
    ),
    re_params = list(
      id = list(sd = c(0.6, 0.2), corr = diag(2)),
      marker = list(sd = c(0.4, 0.15), corr = diag(2))
    )
  ),
  vcov_diag_link = "exp"
)

# Every fitting argument, including data, formulae and fitting priors, is retained.
recovery_fit <- do.call(
  joinme,
  c(simulation$truth$recovery$arguments, list(seed = 401))
)
```

