---
title: "Discrimination analysis: concordance, time-dependent ROC curves, and AUC"
subtitle: "Statistical estimands, censoring rules, computation, and interpretation"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
```

# Purpose and scope

This vignette gives a transparent account of the discrimination measures
implemented for a fitted `JoiNMeFit` object. It is intended to provide enough
detail for a statistical methods section, independent reproduction of the
estimators, and critical assessment of their assumptions.

The two public functions answer different scientific questions:

- `concordance()` asks whether the fitted **conditional survival curves order
  observed failure times correctly over the available follow-up**. It returns
  one follow-up-wide index.
- `tvROC()` asks how the fitted **conditional event risk at a specified
  horizon separates cases from controls at a chosen landmark** over all
  classification thresholds. `tvAUC()` integrates that receiver operating
  characteristic curve and returns one cumulative/dynamic area for every
  requested landmark--horizon combination.

These are measures of **discrimination**, not calibration. A model may order
subjects correctly while systematically overestimating or underestimating
their absolute risks. Conversely, well-calibrated group averages do not imply
that individual subjects are ordered well.

This distinction is the reason that `concordance()` does not require
`time_horizon` or `Dt`, whereas `tvROC()` and `tvAUC()` do. Supplying a
horizon changes the scientific question from follow-up-wide concordance to
horizon-specific classification.

## One prediction pathway, two estimands

Both measures begin with the same fitted joint model and the same prospective
restriction of each longitudinal history. They separate only after posterior
mean conditional survival has been obtained.

```{mermaid}
flowchart LR
    A["Fitted joint model<br/>and evaluation data"] -->
    B["One terminal outcome<br/>per subject"]
    B --> C["Choose landmark and retain<br/>history observed by that time"]
    C --> D["Predict posterior mean<br/>conditional survival"]
    D --> E["Concordance:<br/>compare curves at each<br/>observable event time"]
    D --> F["tvROC/tvAUC:<br/>classify risks at one<br/>specified horizon"]
```

The shared prediction pathway is important: a difference between the two
reported numbers arises from their comparison sets and time summaries, not
from fitting two different survival models.

# Statistical setting and notation

## Observed event data

For subject $i=1,\ldots,n$, let

- $xT_i^\ast$ be the latent event time;
- $C_i$ be the censoring time;
- $T_i=\min(T_i^\ast,C_i)$ be the observed event or censoring time;
- $\delta_i=1$ indicate an observed event and $\delta_i=0$ indicate
  censoring;
- $K_i$ be the observed event cause when competing events are present;
- $\mathcal H_i(s)$ be the longitudinal history observed through time $s$;
  and
- $\mathbf X_i$ be the event-process covariates.

The fitted event formula may use either a right-censored
`Surv(time, status)` response or counting-process
`Surv(start, stop, status)` data. Counting-process data can contain several
rows for one subject. Discrimination nevertheless concerns one terminal
outcome per subject. JoiNMe therefore orders the intervals and retains the
terminal row before forming any case, control, or comparable-pair indicator.
Left- and interval-censored outcomes are not silently reinterpreted; they are
rejected because they require different comparison rules.

## Conditional prediction origin

Let $T_{0i}$ denote the time at which prediction is made for subject $i$.
Only longitudinal observations at or before $T_{0i}$ are allowed to inform
the prediction. Define residual follow-up by

$$
R_i = T_i-T_{0i}.
$$

The conditional survival function on this residual time scale is

$$
S_i(u\mid T_{0i})
=
\Pr\!\left(
T_i^\ast>T_{0i}+u
\mid
T_i^\ast>T_{0i},
\mathcal H_i(T_{0i}),
\mathbf X_i,
\mathcal D_{\mathrm{fit}}
\right),
\qquad u\geq 0,
$$

where $\mathcal D_{\mathrm{fit}}$ denotes the data used to obtain the
posterior distribution. By construction,
$S_i(0\mid T_{0i})=1$.

JoiNMe calculates survival over posterior draws and uses the posterior mean
ordinate

$$
\widehat S_i(u\mid T_{0i})
=
\operatorname{E}\!\left[
S_i(u\mid T_{0i},\boldsymbol\theta)
\mid \mathcal D_{\mathrm{fit}}
\right]
$$

for discrimination. The present functions therefore report a plug-in
discrimination estimate from posterior mean curves. They do not report the
posterior distribution, standard error, or confidence interval of the
discrimination measure.

## How the joint model enters a survival prediction

For cause $k$, the event model may be written schematically as

$$
\lambda_{ik}(t)
=
\lambda_{0k}(t)
\exp\!\left\{
\mathbf X_i(t)^\mathsf T\boldsymbol\gamma_k
+
\mathcal A_{ik}\!\left(
t,\mathcal H_i(t)
\right)
\right\},
$$

where $\lambda_{0k}(t)$ is the fitted baseline hazard and
$\mathcal A_{ik}$ is the complete longitudinal--event association
contribution. The resulting event-free survival has the form

$$
S_i(t\mid s)
=
\exp\!\left[
-\sum_k\int_s^t \lambda_{ik}(v)\,dv
\right],
\qquad t>s,
$$

with the conditioning adjustment that survival equals one at $s$.

Consequently, the discrimination functions do not reconstruct a simplified
risk score. They call `predict.JoiNMeFit()` and inherit:

- the fitted baseline-hazard basis and its original centring constants;
- event covariates and time-varying event-process rows;
- posterior random effects informed by the longitudinal history;
- marker weights;
- current-value, current-slope, functional, correlation, and covariance
  association terms; and
- monotone-spline and ordered piecewise-linear association transforms.

For an ordered piecewise-linear association, the ordered ordinates generated
by the simplex and cumulative sum enter the hazard at every quadrature and
prediction time. No separate approximation or unconstrained coefficient is
introduced by `concordance()`, `tvROC()`, or `tvAUC()`.

# Follow-up-wide survival-curve concordance

## The estimand

Suppose subject $i$ experiences the cause of interest at residual time
$R_i$, and subject $j$ is known to remain event-free beyond that residual
time. Their observable order is

$$
R_i<R_j.
$$

Following the survival-curve principle proposed by
[Antolini, Boracchi, and Biganzoli (2005)](https://doi.org/10.1002/sim.2427),
JoiNMe evaluates **both** conditional survival curves at the earlier observed
event time $R_i$. The pair is:

$$
\begin{aligned}
\text{concordant} \quad&
\text{if}\quad
\widehat S_i(R_i\mid T_{0i})
<
\widehat S_j(R_i\mid T_{0j}),\\
\text{discordant} \quad&
\text{if}\quad
\widehat S_i(R_i\mid T_{0i})
>
\widehat S_j(R_i\mid T_{0j}),\\
\text{a prediction tie} \quad&
\text{if}\quad
\widehat S_i(R_i\mid T_{0i})
=
\widehat S_j(R_i\mid T_{0j}).
\end{aligned}
$$

Lower survival means greater predicted risk. Thus a concordant pair assigns
greater risk at $R_i$ to the subject who actually experiences the event at
$R_i$.

Let $\mathcal P$ be the set of observable ordered pairs and let $a_i$ be
the event-time weight attached to comparisons generated by event $i$.
Define

$$
c_{ij}
=
\mathbb I\!\left\{
\widehat S_i(R_i)<\widehat S_j(R_i)
\right\}
+
\frac{1}{2}
\mathbb I\!\left\{
\widehat S_i(R_i)=\widehat S_j(R_i)
\right\}.
$$

The reported concordance is

$$
\widehat C
=
\frac{
\displaystyle
\sum_{(i,j)\in\mathcal P}a_i c_{ij}
}{
\displaystyle
\sum_{(i,j)\in\mathcal P}a_i
}.
$$

A value of $0.5$ represents chance ordering under continuous predictions, a
value approaching $1$ represents increasingly correct ordering, and a value
below $0.5$ indicates systematic reversal. These descriptions concern
ranking only and do not establish calibration or clinical usefulness.

## Comparable pairs and censoring

Censoring does not imply long survival. It only limits which temporal
orderings are observed. JoiNMe follows the risk-set convention described for
[`survival::concordance()`](https://stat.ethz.ch/R-manual/R-devel/library/survival/html/concordance.html).
For an event of interest at residual time $r$:

| Outcome for comparator $j$ | Comparable? | Reason |
|---|---:|---|
| Event or censoring after $r$ | Yes | Subject $j$ is known to survive beyond the event time |
| Censoring exactly at $r$ | Yes | Under the event-before-censor convention, $j$ is known to be event-free through $r$ |
| Censoring before $r$ | No | The order between the event and $j$'s latent event time is unknown |
| Event of interest exactly at $r$ | No | The two failure times are tied and have no observed strict order |
| Competing event before $r$ | No | It is treated as cause-specific censoring before the comparison time |
| Competing event exactly at $r$ | Yes | It follows the same event-before-censor convention as other censoring at $r$ |

This pairwise rule uses a prematurely censored subject whenever its observed
history is sufficient for a particular earlier comparison, and excludes it
after its censoring time. It neither declares the subject a late survivor nor
imputes an unobserved failure order.

The interpretation relies on the censoring assumptions appropriate to the
chosen weighting method. Informative loss to follow-up that remains
unexplained by the fitted data can bias discrimination estimates. No
comparison rule alone can identify the ordering of two latent event times
after observation has ceased.

## Event-time weighting

The `type_weights` argument selects an event-time weighting schedule. JoiNMe
asks `survival::concordance(..., ranks = TRUE)` for its documented event-time
weights and divides each event weight by the number at risk at that event
time. This produces the pair weight $a_i$ used in the survival-curve
comparisons.

| `type_weights` | Interpretation in JoiNMe |
|---|---|
| `"none"` or `"n"` | Harrell-style pair weighting; every observable pair has weight one |
| `"S"` | Weight events using the estimated survival schedule |
| `"S/G"` | Weight events using the survival and censoring schedules |
| `"n/G2"` | Uno-style inverse-censoring event weighting supplied by `survival` |
| `"I"` | Equal event-time weighting supplied by `survival` |

For `"n"`, the event-time weight returned by `survival` equals the risk-set
size, so division by that size gives unit weight to every pair. The other
options change the relative contribution of event times.

The `"n/G2"` option reuses the event-time weighting associated with Uno's
concordance in `survival`; the subject scores in JoiNMe remain the
time-dependent survival-curve comparisons defined above. It should therefore
be described as an Uno-style censoring-weighted Antolini comparison, not as a
claim that the implementation is identical in every respect to a conventional
static-score Uno estimator.

## Choice of prediction origin

By default, `concordance(fit)` chooses

$$
T_{0i}
=
\max\{t<T_i: Y_i(t)\text{ was observed}\}.
$$

The strict inequality prevents a measurement recorded at the outcome time
from informing a prediction assessed against that same outcome. Subjects
without a valid pre-outcome longitudinal history cannot contribute.

This default answers a residual-prognosis question: **given each subject's
most recent pre-outcome information, does the model order their remaining
follow-up?** Because $T_{0i}$ may differ between subjects, residual times
rather than absolute study times are compared.

A common landmark gives a different and often more directly comparable
estimand:

```{r common-landmark}
concordance(fit, time_start = 2)
```

This asks whether predictions made at study time 2 order subsequent event
times among subjects still under observation after time 2. The landmark may
also be a named vector or the name of a subject-constant column in the event
data:

```{r subject-specific-landmark}
subject_landmark <- c("1" = 1.5, "2" = 2.0, "3" = 1.0)

concordance(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = subject_landmark
)

concordance(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = "clinical_landmark"
)
```

Subject-specific origins are appropriate only when the scientific target is
remaining time from those origins. They should not be presented as
concordance on a common absolute study-time scale.

## Cause-specific analysis and competing events

The `cause` argument identifies the failure type that generates comparable
event pairs. Other event causes are treated as censoring at their observed
times. They may serve as comparators before their competing event, but not
after it.

The current `predict()` result used as the score is the fitted event-free
`Survival`, which reflects the complete fitted event process. Consequently,
with competing risks, the pair construction is cause-specific but its ranking
quantity is overall event-free survival. This is transparent and reproducible,
but it is not the same estimand as a cause-specific cumulative-incidence
concordance. A scientific question about cumulative incidence in the presence
of competing risks requires a dedicated competing-risk discrimination
estimator.

# Concordance: implementation from call to result

The following sequence describes every public and internal function used by
the concordance calculation. Internal functions begin with a full stop and are
not part of the stable user interface.

## Step 1: validate the request and resolve the data

`concordance.JoiNMeFit()`:

1. verifies that `object` is a fitted `JoiNMeFit`;
2. checks that `cause` is a single positive integer;
3. asks `.concordance_time_weight()` to validate and translate
   `type_weights`;
4. calls `.get_train_data()`; and
5. passes the aligned data to `.concordance_survival_curves()`.

`.get_train_data()` uses the stored fitting data only when both
`newdataLong` and `newdataEvent` are omitted. If either new data set is
provided, both are required. This prevents an accidental mixture of validation
longitudinal histories with training event outcomes, or the converse.

`.concordance_time_weight()` maps `"none"` to the `survival` spelling `"n"`
and rejects any unknown weighting rule before prediction begins.

## Step 2: construct one terminal outcome per subject

`.subject_event_outcomes()`:

1. finds the `Surv()` response in the fitted event formula;
2. verifies that the subject identifier is present and non-missing;
3. accepts right-censored and counting-process responses;
4. orders counting-process intervals by subject, stop time, and start time;
5. retains the final interval for each subject; and
6. decodes the event indicator and event cause into `event_status` and
   `event_type`.

The ordering makes the reduction invariant to the row order supplied by the
user. Retaining one row avoids counting a subject once for every
time-dependent covariate interval.

## Step 3: determine and validate landmarks

When `time_start` is omitted, `.last_preoutcome_measurement()` finds the final
longitudinal measurement strictly before each observed event or censoring
time. A numerical tolerance prevents a value that differs only through
floating-point representation from being treated as earlier.

When `time_start` is supplied, `.subject_time_map()` converts it to one finite,
named value per subject. It accepts:

- one scalar shared by all subjects;
- a named vector matched by subject identifier;
- an unnamed vector already in subject order; or
- a character value naming a subject-constant event-data column.

For counting-process event data, `.constant_finite_value()` verifies that a
named landmark column is finite and constant within subject.

Subjects are eligible only when their observed outcome occurs strictly after
their landmark. The cause-specific event indicator is then

$$
d_i=\mathbb I(\delta_i=1,\ K_i=\texttt{cause}).
$$

## Step 4: restrict the longitudinal history

`.concordance_survival_curves()` retains only measurements satisfying

$$
t_{i\ell}\leq T_{0i}.
$$

This is an essential prospective-prediction condition. Measurements after the
landmark could improve estimation of a subject's random effects, but they were
not available when the prediction was notionally made and would introduce
future information. Subjects without any usable history after this
restriction are removed.

## Step 5: build prediction times

For each subject, `.concordance_prediction_grid()` constructs an absolute-time
grid containing:

- residual time zero, expressed as the subject's landmark;
- 50 regularly spaced points between the landmark and observed follow-up end;
- every distinct cause-specific residual event time at which that subject can
  be a comparator; and
- the subject's observed follow-up end.

Including every required event time makes the concordance ordinate exact with
respect to the prediction grid. The regular points retain a stable curve for
other prediction summaries and protect against sparse interpolation.

## Step 6: obtain conditional survival predictions

`.concordance_survival_curves()` calls:

```r
predict(
  object,
  process = "event",
  times = subject_specific_time_grids,
  time_start = subject_specific_landmarks,
  control = list(n_samples = n_samples),
  seed = seed
)
```

`predict.JoiNMeFit()` evaluates the fitted hazard and its integral for the
selected posterior draws. `n_samples` controls the number of posterior draws
used to estimate each posterior mean survival curve; `seed` makes draw
subsampling reproducible. Additional prediction arguments supplied through
`...` are passed on unchanged.

## Step 7: align curves on residual event times

`.concordance_survival_matrix()` constructs a matrix whose rows are distinct
cause-specific residual event times and whose columns are subjects. For every
subject it:

1. subtracts that subject's landmark from the absolute prediction times;
2. orders the residual grid;
3. linearly interpolates at the required event times; and
4. returns `NA` outside the subject's predicted support.

The necessary event times were inserted explicitly in Step 5, so interpolation
normally returns a stored ordinate. It mainly guards against harmless
floating-point changes during time rescaling. Prohibiting extrapolation ensures
that a subject is not compared beyond their observed support.

## Step 8: obtain event weights

`.concordance_event_weights()` constructs a right-censored residual-time
response and calls `survival::concordance(..., ranks = TRUE)` with an arbitrary
fixed working score. The score is used only because the `survival` interface
requires one; it cannot affect the risk-set sizes or event-time weights.

The function aligns the returned rank table to event rows, including the
single-event case in which R may drop a matrix dimension. It divides each
event-time weight by the number of subjects whose residual follow-up reaches
that time. Non-event rows receive weight zero.

## Step 9: visit every observable pair once

`.concordance_from_survival_curves()` visits each cause-specific event and
selects comparators who:

- have residual follow-up greater than the event time; or
- are censored exactly at the event time; and
- have a finite survival ordinate at that event time.

It subtracts the event subject's survival from every comparator's survival,
counts positive, negative, and zero differences, and multiplies the counts by
the event's pair weight. Each strict temporal ordering is visited once. The
final estimate gives half credit to a predicted tie.

The implementation stores an event-time-by-subject matrix rather than a
posterior-draw-by-time-by-subject array. If there are $E$ distinct event
times, its principal storage is $O(En)$. Pair comparison is
$O(n^2)$ in the worst case, which is unavoidable for an exact all-pairs
empirical concordance without further structure.

## Step 10: return an auditable summary

The result contains:

| Column | Meaning |
|---|---|
| `concordance` | $( \text{concordant}+0.5 \times \text{tied} )/ n_{\text{pairs}}$ |
| `concordant` | Weighted number of correctly ordered pairs |
| `discordant` | Weighted number of incorrectly ordered pairs |
| `tied` | Weighted number of survival-probability ties |
| `n_pairs` | Total comparison weight |
| `n_subjects` | Number of subjects retained after landmark and history checks |
| `n_events` | Number of observed events of the requested cause |

With non-unit event weighting, the first four pair-count columns are weighted
amounts and need not be integers.

# Cumulative/dynamic time-dependent ROC analysis

## Landmark population and horizon status

Choose a common landmark $s$ and horizon $t>s$. The relevant population
contains subjects known to be event-free at $s$ and having longitudinal
history available by $s$. For such a subject, define posterior mean
conditional survival and event risk by

$$
\widehat S_i(s,t)
=
\Pr(T_i^\ast>t\mid T_i^\ast>s,\mathcal H_i(s),\mathbf X_i,
\mathcal D_{\mathrm{fit}})
$$

and

$$
\widehat r_i(s,t)=1-\widehat S_i(s,t).
$$

JoiNMe follows the cumulative/dynamic convention used by `JMbayes2::tvROC()`.
The prediction window excludes the conditioning instant and includes its
horizon, $(s,t]$:

$$
D_i(s,t)
=
\begin{cases}
1,
& s<T_i\leq t,\ \delta_i=1,\ K_i=\texttt{cause},\\
0,
& T_i>t,\\
\text{unknown},
& s<T_i\leq t\ \text{without an observed event of the requested cause}.
\end{cases}
$$

An observed requested-cause event in the window is a case. A subject known to
remain under observation without that cause through the horizon is a dynamic
control. Administrative censoring or a competing event before the horizon
leaves the requested-cause status at the horizon unknown. Outcomes at or
before the landmark are excluded because dynamic prediction conditions on
survival beyond the landmark.

The boundary convention is important at tied recorded times. A requested
event recorded exactly at $t$ is a case for the interval ending at $t$,
whereas a subject must be observed beyond $t$ to be a control. Users should
choose time units and horizons before analysis rather than move a horizon
after seeing the recorded event times.

## ROC thresholds, sensitivity, and false-positive fraction

The implementation places 101 thresholds

$$
c_\ell\in\{0,0.01,\ldots,1\}.
$$

It follows the JMbayes2 representation and applies thresholds to conditional
survival. Subject $i$ is classified as a predicted case at threshold $c$ when

$$
A_i(c)=\mathbb I\{\widehat S_i(s,t)<c\}.
$$

This is equivalent to classifying on risk with threshold $1-c$, but using
survival gives the same orientation and stored fields as a JMbayes2 `tvROC`
object. As $c$ increases, both sensitivity and the false-positive fraction
are non-decreasing.

Let $w_i^{(1)}$ and $w_i^{(0)}$ be the case and control weights described
below. The estimated true-positive and false-positive fractions are

$$
\widehat{\operatorname{TP}}(c)
=
\frac{\sum_i w_i^{(1)}A_i(c)}
{\sum_i w_i^{(1)}}
$$

and

$$
\widehat{\operatorname{FP}}(c)
=
\frac{\sum_i w_i^{(0)}A_i(c)}
{\sum_i w_i^{(0)}}.
$$

The ROC curve is the ordered set
$\{(\widehat{\operatorname{FP}}(c_\ell),
\widehat{\operatorname{TP}}(c_\ell))\}_{\ell=1}^{101}$.

## Model-based treatment of premature censoring

With `type_weights = "model-based"`, observed horizon status is represented
directly and unknown status is replaced by its fitted expectation:

$$
w_i^{(1)}
=
\begin{cases}
1, & D_i=1,\\
0, & D_i=0,\\
\widehat r_i(s,t), & D_i\ \text{unknown},
\end{cases}
\qquad
w_i^{(0)}=1-w_i^{(1)}.
$$

Thus an early-censored subject is not declared a case or a control. Instead,
the subject contributes fractionally to both groups according to their
predicted event probability at the landmark. This uses the fitted joint model
to integrate over the unobserved horizon status. It is coherent only to the
extent that the event model and its conditioning information are adequate for
the censoring setting. It does not turn an informative dropout process into
independent censoring by assumption.

The same posterior mean risk determines both the uncertain-status weight and
the primary threshold classification. This matches the plug-in character of
the main ROC curve. Posterior-draw ROC curves retain the status weights and
vary the subject classification over posterior survival draws.

## Inverse-probability-of-censoring treatment

With `type_weights = "IPCW"`, subjects of unknown horizon status receive zero
case and control weight. Let

$$
G(u\mid s)=\Pr(C_i\geq u\mid C_i>s)
$$

be the conditional probability of remaining uncensored. JoiNMe estimates
$G$ by a Kaplan--Meier curve in the landmark risk set. Requested-cause
failures are failures rather than censoring events in this auxiliary
distribution; administrative censoring and competing events terminate
observation of the requested cause.

The weights are

$$
w_i^{(1)}
=
\frac{\mathbb I(D_i=1)}{\widehat G(T_i-\mid s)}
$$

and

$$
w_i^{(0)}
=
\frac{\mathbb I(D_i=0)}{\widehat G(t\mid s)}.
$$

The immediately preceding censoring probability is used for a case so that a
censoring jump tied at its event time is not allowed to precede the observed
event. IPCW targets the landmark population under conditional independent
censoring and positivity. Very small values of $\widehat G$ produce unstable
weights; users should inspect the number at risk and avoid horizons beyond
reliable follow-up.

## Posterior-draw and posterior-mean curves

Dynamic prediction retains a survival matrix with posterior draws in rows and
prediction times in columns for every subject. At the exact horizon, JoiNMe
aligns these values into a subject-by-draw event-risk matrix. For draw $m$,
the classification is

$$
A_i^{(m)}(c)
=
\mathbb I\{1-r_i^{(m)}(s,t)<c\}.
$$

The `tp` and `fp` matrices in the returned object contain one ROC curve per
retained posterior draw. The upper-case `TP` and `FP` vectors are the primary
curve calculated from posterior mean subject risks. They are not obtained by
averaging the posterior-draw ROC curves. This distinction is retained because
it is the data structure expected by `JMbayes2::tvAUC.tvROC()` and its
plotting method.

A posterior draw containing a non-finite horizon risk is discarded as a
whole curve. A subject with a non-finite posterior mean risk is removed before
the case and control totals are formed. If no complete posterior-draw column
remains, the posterior mean curve is stored as a single deterministic column.

# Area under the time-dependent ROC curve

## Trapezoidal definition

`tvAUC()` first runs `tvROC()` when supplied with a `JoiNMeFit` object. It then
passes the resulting `tvROC` object to `JMbayes2::tvAUC()`. S3 dispatch selects
`JMbayes2::tvAUC.tvROC()`, which evaluates

$$
\widehat{\operatorname{AUC}}(s,t)
=
\sum_{\ell=1}^{100}
\frac{
\widehat{\operatorname{TP}}(c_{\ell+1})
+
\widehat{\operatorname{TP}}(c_{\ell})
}{2}
\left\{
\widehat{\operatorname{FP}}(c_{\ell+1})
-
\widehat{\operatorname{FP}}(c_{\ell})
\right\}.
$$

This is the trapezoidal area under the weighted cumulative/dynamic ROC curve.
In the absence of censoring weights and with sufficiently complete threshold
coverage, it has the familiar interpretation as the probability that a case
has a greater predicted event risk than a control, with half credit to risk
ties. Under model-based or IPCW weighting, that probability is taken with
respect to the corresponding weighted landmark population.

The area is calculated from `TP` and `FP`, and therefore from posterior mean
subject risks. It is not a posterior mean of draw-specific areas and it is not
a credible interval. A resampling or posterior procedure would be required
for uncertainty in predictive performance.

## Optimal threshold summaries

For descriptive use, `tvROC()` also returns two survival-probability cut-offs:

- `F1score` is the median threshold among all finite maxima of
  $2\,n_{\mathrm{TP}}/(2n_{\mathrm{TP}}+n_{\mathrm{FN}}+n_{\mathrm{FP}})$;
- `Youden` is the median threshold among all maxima of
  $\operatorname{TP}-\operatorname{FP}$.

These are data-dependent summaries, not automatically recommended clinical
decision thresholds. Their consequences depend on event prevalence, the
relative costs of errors, calibration, and the intended intervention.

# ROC implementation from call to result

## Step 1: dispatch and preserve the JoiNMe argument recipe

JoiNMe re-exports the `JMbayes2::tvROC()` and `JMbayes2::tvAUC()` generics.
The `JoiNMeFit` methods preserve the argument names used by the former JoiNMe
AUC interface:

```r
tvROC(
  object,
  newdataLong = NULL,
  newdataEvent = NULL,
  time_start = 0,
  time_horizon = NULL,
  Dt = NULL,
  cause = 1,
  n_samples = 200,
  seed = 123,
  type_weights = c("model-based", "IPCW"),
  ...
)
```

`tvAUC.JoiNMeFit()` has the same recipe. The only new choice is
`type_weights`, which makes the censoring treatment explicit.

`.get_train_data()` selects the stored fitting data when both new-data
arguments are absent. If either new-data argument is supplied, both must be
supplied so that longitudinal history and event outcome refer to the same
evaluation cohort.

## Step 2: resolve landmarks and horizons

`.get_tvroc_times()` validates finite numeric landmarks. When
`time_horizon` is omitted:

- `Dt` defines $t=s+\texttt{Dt}$; or
- when `Dt` is also absent, `.default_discrimination_horizon()` uses the
  fitted maximum follow-up, falling back to the largest finite observed event
  or censoring time.

A scalar horizon is repeated across several landmarks. Otherwise there must
be one horizon for each landmark. Every horizon must be strictly greater than
its landmark. Several landmarks are evaluated separately; they do not form
repeated observations of one common ROC curve.

## Step 3: recover one terminal event outcome per subject

`.subject_event_outcomes()` reads the event response from the fitted formula.
For right-censored data, the event time and status are used directly. For
counting-process data, rows are ordered within subject and reduced to the
terminal interval. Event type is retained when a cause-coded status is used.
Left- and interval-censored responses are rejected because their horizon
status requires a different estimator.

`.subject_time_map()` aligns scalar, named, and event-column time declarations
by subject identifier. A time column repeated over counting-process rows must
be finite and constant within subject.

## Step 4: retain prospective longitudinal history

`.dynamic_discrimination_risk_set()` retains longitudinal observations whose
recorded time is no later than the landmark. A subject whose outcome occurred
at or before the landmark is removed. These restrictions ensure that a
horizon classification is based only on information available when the
prediction would have been made.

The filtering uses indexed vectors rather than repeatedly splitting and
joining subject data. Its cost is linear in the number of longitudinal rows.

## Step 5: obtain conditional survival at the exact horizon

`.discrimination_time_grid()` creates 50 equally spaced prediction times from
the landmark through the horizon, including both endpoints.
`predict.JoiNMeFit(process = "event")` then evaluates the complete fitted
joint model. The horizon row is selected with a relative numerical tolerance;
a neighbouring grid point is never substituted.

The risk-set data retain the posterior mean risk, terminal time, event status,
event type, landmark, horizon, and truncated follow-up. The prediction call
also retains raw posterior survival draws. Consequently identity,
formula-based, monotone-spline, and ordered piecewise-linear associations are
handled in exactly the same way as ordinary dynamic survival prediction.

## Step 6: align raw posterior draws

`.discrimination_horizon_risk_draws()` finds the exact horizon column in each
subject's raw survival matrix and converts it to event risk. Older prediction
objects may contain unequal numbers of draws; the common minimum is used to
form a rectangular matrix without recycling values. If raw draws are absent,
ROC estimation remains available from posterior mean risks.

## Step 7: construct censoring weights

`.tvroc_status_weights()` implements the equations above. Model-based weights
use fractional expected case status. IPCW weights first estimate the
cause-specific censoring survival curve, then weight observed cases at their
event time and controls at the horizon. The helper returns separate
non-negative case and control vectors, making the denominator of each ROC
ordinate explicit.

## Step 8: calculate every threshold efficiently

`.tvroc_from_risk_set()` forms a subject-by-threshold logical classification
matrix with `outer()`. Matrix cross-products with the case and control weight
vectors calculate all 101 numerators without constructing any case--control
pair matrix. This reduces the central calculation to $O(101n)$ operations and
uses a compact $n\times101$ temporary matrix.

Posterior-draw curves are evaluated one draw at a time. This avoids a
three-dimensional subject-by-threshold-by-draw array, whose memory use can be
substantial in external validation samples.

## Step 9: return a JMbayes2-compatible object

For one landmark, `tvROC.JoiNMeFit()` returns class `"tvROC"` with the
JMbayes2 fields:

| Field | Statistical content |
|---|---|
| `TP`, `FP` | Primary sensitivity and false-positive fractions |
| `nTP`, `nFN`, `nFP`, `nTN` | Weighted classification totals |
| `tp`, `fp` | Posterior-draw ROC ordinate matrices |
| `thrs` | The 101 survival thresholds |
| `F1score`, `Youden` | Descriptive optimal threshold summaries |
| `Tstart`, `Thoriz` | Landmark and horizon |
| `nr` | Number of analysable subjects |
| `type_weights` | Censoring treatment |
| `nameObject`, `classObject` | Model labels used by JMbayes2 methods |

JoiNMe additionally retains `cause` and the `thr` spelling for transparent
inspection. The inherited `plot.tvROC()` method draws `FP` against `TP`.

For several landmarks, `tvROC.JoiNMeFit()` returns a
`"tvROC_JoiNMeFit_list"` with one standard `tvROC` object in `curves` for each
aligned landmark--horizon pair. This explicit structure prevents a plot from
silently combining distinct estimands.

## Step 10: calculate area through JMbayes2 dispatch

`tvAUC.JoiNMeFit()` calls `tvROC()` with the unchanged arguments. A scalar
result is passed directly to `JMbayes2::tvAUC()`, whose
`tvAUC.tvROC()` method performs trapezoidal integration. The returned object
has class `"tvAUC"`.

For several landmarks, each compatible curve is integrated separately and a
`"tvAUC_JoiNMeFit"` data frame returns `time_start`, `time_horizon`, `auc`,
`n_subjects`, and `type_weights`.

# Concordance and horizon-specific ROC answer different questions

| Feature | `concordance()` | `tvROC()` and `tvAUC()` |
|---|---|---|
| Scientific question | Are earlier failures assigned lower survival than subjects observed to survive longer? | Can horizon cases be separated from horizon controls? |
| Time summary | One follow-up-wide index | One curve and area per landmark--horizon pair |
| Prediction used | Survival at each observable event time | Conditional survival at one horizon |
| Comparison unit | Earlier-event subject versus later-observed subject | Weighted case and control classification |
| Premature censoring | Pair remains usable only before censoring | Model-based expected status or IPCW |
| Prediction ties | Half credit in pair counts | Represented by the ROC steps and trapezoidal area |
| Event-time ties | No strict failure order; excluded | Status follows the $(s,t]$ prediction window |
| Weighting | Selectable event-time weighting | `"model-based"` or `"IPCW"` censoring weights |
| Competing event | Cause-specific censoring at its event time | Cause-specific censoring for horizon status |
| Principal output | Concordance and auditable pair totals | ROC ordinates, thresholds, and trapezoidal AUC |

Neither measure should replace the other. AUC can vary markedly with the
chosen horizon even when follow-up-wide concordance changes little.
Concordance can conceal clinically important horizon-specific behaviour by
combining comparisons made at many event times.

# Reproducible use

## Concordance

```{r concordance-examples}
# Subject-specific origin at the final pre-outcome measurement.
concordance(fit)

# Common study-time landmark and unit pair weights.
concordance(
  fit,
  time_start = 2,
  type_weights = "none",
  predict_control = list(n_samples = 500, n_pred_draws = 100),
  seed = 2026
)

# Uno-style censoring event weights.
concordance(
  fit,
  time_start = 2,
  type_weights = "n/G2",
  predict_control = list(n_samples = 500, n_pred_draws = 100),
  seed = 2026
)
```

Concordance controls posterior extraction and dynamic-prediction draws through
`predict_control`. Stability can be checked by repeating the calculation with
more posterior and prediction draws while retaining the seed.

## Time-dependent ROC curve and AUC

```{r roc-examples}
# Risk through time 5 among subjects under observation after time 2.
roc_2_5 <- tvROC(
  fit,
  time_start = 2,
  time_horizon = 5,
  type_weights = "model-based",
  n_samples = 500,
  seed = 2026
)

plot(roc_2_5)

# Integrate an already calculated curve without repeating prediction.
area_2_5 <- tvAUC(roc_2_5)

# Or calculate the curve and area from the fitted model in one call.
area_2_5_direct <- tvAUC(
  fit,
  time_start = 2,
  time_horizon = 5,
  type_weights = "IPCW",
  n_samples = 500,
  seed = 2026
)

# A three-year prediction window from several landmarks.
areas <- tvAUC(
  fit,
  time_start = c(0, 1, 2),
  Dt = 3,
  type_weights = "model-based",
  n_samples = 500,
  seed = 2026
)
```

For several landmarks, each row is a separate cumulative/dynamic estimand.
To plot them, select the intended compatible curve explicitly, for example
`plot(rocs$curves[[1]])`.

## Evaluation on new subjects

```{r validation-example}
cIndex <- concordance(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = 2,
  predict_control = list(n_samples = 500, n_pred_draws = 100),
  seed = 2026
)

validation_roc <- tvROC(
  fit,
  newdataLong = validation_long,
  newdataEvent = validation_event,
  time_start = 2,
  time_horizon = 5,
  type_weights = "IPCW",
  n_samples = 500,
  seed = 2026
)

validation_auc <- tvAUC(validation_roc)
```

When new data are omitted, the fitting data are used. This is useful for model
checking but usually optimistic because those subjects informed the posterior
random effects and model parameters. Predictive claims should preferably use
external validation or a pre-specified cross-validation procedure in which
the complete modelling procedure is repeated within each training fold.

# Interpretation and reporting

## What should be reported for concordance

A manuscript should state:

1. whether the origin was the default final pre-outcome measurement, a common
   landmark, or a subject-specific clinical landmark;
2. the event cause and treatment of competing events;
3. the event-time weighting rule;
4. whether evaluation was apparent, internally validated, or externally
   validated;
5. the posterior prediction settings and random seed;
6. the number of retained subjects and cause-specific events;
7. the estimate and weighted comparable-pair count; and
8. that the reported value is calculated from posterior mean curves and has
   no uncertainty interval unless one was obtained separately.

An example methods statement is:

> We assessed follow-up-wide discrimination at a common landmark of two years
> using conditional survival-curve concordance. For each observed
> cause-specific event, both subjects' posterior mean survival curves were
> compared at the earlier residual event time. Subjects censored before that
> time were not comparable, prediction ties received half credit, and
> event-time weights followed the pre-specified `n/G2` rule.

## What should be reported for ROC and AUC

A manuscript should state:

1. the landmark and horizon on the original time scale;
2. the cumulative/dynamic case definition on $(s,t]$ and dynamic control
   definition;
3. the event cause and treatment of competing events;
4. whether censoring weights were model-based or IPCW;
5. the analysable sample size and, for IPCW, the support of the estimated
   censoring curve;
6. whether evaluation was apparent, internally validated, or externally
   validated;
7. `n_samples` and the random seed;
8. whether thresholds or the full ROC curve were examined; and
9. that the area is a trapezoidal plug-in estimate from posterior mean risks,
   without an uncertainty interval unless separately calculated.

An example methods statement is:

> At a landmark of two years, we estimated the cumulative/dynamic ROC curve
> for the five-year horizon. Cases experienced the cause of interest after
> two years and through five years, and controls were known to
> remain without that cause through five years. Premature censoring was
> handled using inverse Kaplan--Meier censoring weights. The AUC was the
> trapezoidal area under the ROC curve obtained from posterior mean
> conditional survival probabilities.

# Assumptions, limitations, and diagnostic checks

## Time scale and landmark comparability

All event times, longitudinal times, landmarks, and horizons must use the same
original time scale. A subject-specific concordance origin defines a
remaining-lifetime estimand; a common ROC origin defines a common-landmark
estimand. These should not be combined under one interpretation.

## Availability of longitudinal history

A dynamic prediction requires longitudinal information observed by its
landmark. Removing subjects without such history changes the analysed
population. The returned subject counts should therefore be checked against
the intended cohort.

## Censoring and positivity

Concordance uses observable orderings and may alter their event-time weights.
ROC estimation either models unknown horizon status or assumes conditional
independent censoring for IPCW. Neither choice makes unexplained informative
dropout harmless. IPCW additionally requires a positive probability of
remaining observed through every weighted time. Sensitivity analyses are
appropriate when dropout is associated with unobserved deterioration.

## Small case or control weight

Concordance requires at least one observable event--comparator pair. ROC
estimation requires positive total case and control weight and at least one
observed event of the requested cause in the window. An error in such a
window describes an undefined empirical discrimination problem; changing the
horizon solely to obtain a result would change the estimand.

## Monte Carlo stability

Posterior mean predictions and draw-specific ROC curves depend on the selected
posterior draws. Compare results at increasing `n_samples` with a fixed seed
until scientifically material digits are stable. This examines Monte Carlo
approximation only; it does not quantify sampling, validation, or posterior
uncertainty in discrimination.

## Apparent versus predictive performance

Calculating a measure on the fitting data assesses apparent discrimination.
Random effects and association parameters have been informed by those
subjects. Held-out evaluation must preserve the full modelling procedure,
including model selection and transform choices, within the training data.

## Discrimination is not calibration or utility

Neither statistic verifies absolute-risk calibration, prediction-interval
coverage, net benefit, or an acceptable clinical decision threshold.
Discrimination should be reported alongside calibration and predictive
uncertainty appropriate to the intended use.

# Function-by-function reference

The complete discrimination path is summarised below. This table is also a map
for readers auditing `R/diagnosis.R`.

| Function | Statistical role |
|---|---|
| `concordance.JoiNMeFit()` | Validates, predicts, compares observable survival orders, and returns one follow-up-wide result |
| `tvROC()` | Re-exported JMbayes2 generic for time-dependent ROC estimation |
| `tvROC.JoiNMeFit()` | Resolves JoiNMe data and times and calculates one compatible ROC curve per landmark |
| `tvAUC()` | Re-exported JMbayes2 generic for trapezoidal time-dependent AUC |
| `tvAUC.JoiNMeFit()` | Runs `tvROC()` and delegates each compatible curve to `JMbayes2::tvAUC.tvROC()` |
| `.get_train_data()` | Selects stored fitting data or a complete pair of new longitudinal and event data |
| `.constant_finite_value()` | Verifies that a subject-level time column is finite and constant within subject |
| `.discrimination_time_grid()` | Creates the 50-point landmark-to-horizon prediction grid |
| `.default_discrimination_horizon()` | Finds a fitted or observed maximum time when no horizon is supplied |
| `.get_tvroc_times()` | Aligns the existing JoiNMe landmark, horizon, and `Dt` recipe |
| `.concordance_time_weight()` | Validates concordance weighting and maps `"none"` to `"n"` |
| `.subject_time_map()` | Aligns scalar, vector, named-vector, or event-column time declarations by subject |
| `.subject_event_outcomes()` | Reduces right-censored or counting-process data to one terminal outcome per subject |
| `.last_preoutcome_measurement()` | Finds the default subject-specific concordance origin without using an outcome-time measurement |
| `.concordance_prediction_grid()` | Adds exact residual event times to each subject's regular prediction grid |
| `.concordance_survival_curves()` | Restricts histories, calls prediction, and returns aligned outcomes and curves |
| `.concordance_survival_matrix()` | Aligns posterior mean curves on distinct residual event times without extrapolation |
| `.concordance_event_weights()` | Obtains event-time weights from `survival` and converts them to pair weights |
| `.concordance_from_survival_curves()` | Counts each observable concordant, discordant, or tied pair and forms $\widehat C$ |
| `.dynamic_discrimination_risk_set()` | Builds one horizon risk set from the complete dynamic-prediction pathway |
| `.discrimination_horizon_risk_draws()` | Aligns raw posterior event-risk draws at each subject's exact horizon |
| `.censoring_survival_at()` | Evaluates the Kaplan--Meier censoring curve at event or horizon times |
| `.tvroc_status_weights()` | Constructs model-based or IPCW case and control weights |
| `.central_optimal_threshold()` | Returns the central threshold among tied finite criterion maxima |
| `.tvroc_from_risk_set()` | Calculates weighted ROC ordinates and returns the JMbayes2-compatible object |
| `predict.JoiNMeFit()` | Supplies conditional survival from the complete fitted joint model, including ordered piecewise-linear associations |

# References

Antolini L, Boracchi P, Biganzoli E (2005). A time-dependent discrimination
index for survival data. *Statistics in Medicine*, 24, 3927--3944.
[doi:10.1002/sim.2427](https://doi.org/10.1002/sim.2427).

Heagerty PJ, Zheng Y (2005). Survival model predictive accuracy and ROC
curves. *Biometrics*, 61, 92--105.
[doi:10.1111/j.0006-341X.2005.030814.x](https://doi.org/10.1111/j.0006-341X.2005.030814.x).

Rizopoulos D (2011). Dynamic predictions and prospective accuracy in joint
models for longitudinal and time-to-event data. *Biometrics*, 67, 819--829.
[Publisher record](https://onlinelibrary.wiley.com/doi/10.1111/j.1541-0420.2010.01546.x).

Rizopoulos D, Papageorgiou G, Miranda Afonso P (2024). `JMbayes2` accuracy
measures. The package reference defines the `tvROC` and `tvAUC` interfaces,
censoring choices, returned structures, and associated methods.
[R documentation](https://search.r-project.org/CRAN/refmans/JMbayes2/html/accuracy.html).

Therneau TM. `survival::concordance`: concordance for survival and regression
models. The official documentation describes risk-set comparability, event
ties, and the available time-weighting rules.
[R documentation](https://stat.ethz.ch/R-manual/R-devel/library/survival/html/concordance.html).

Therneau TM. *Concordance* vignette for the `survival` package. The vignette
gives the computational and censoring conventions used by the event-time
weighting engine.
[Official vignette](https://stat.ethz.ch/R-manual/R-devel/library/survival/doc/concordance.pdf).
