---
title: "joinme: Usage Demo"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  cache = FALSE
)
options(width = 80)

library(dplyr)
library(joinme)
options(stan_preferred_engine = 'rstan')
```

# Purpose and Reproducibility

This vignette demonstrates the usage of `joinme` with multiple scenarios, including **data simulation**, **model fitting**, **diagnostic interpretation**, **dynamic prediction**, **plotting**, and **threaded computation**. Each block is designed to be reproducible and to illustrate a specific modeling choice.

For the methodological background, see the theory vignette in [joinme-theory2.html](joinme-theory2.html).

This demo covers mixed families (including double_exponential, skew_double_exponential,
beta, and cumulative_logit) and competing-risk survival via event type columns.

```{r seed}
set.seed(123)
```

# Quick model-spec checklist

Accepted specifications include:

- **Longitudinal structure** (`formulaLong`, `families`, optional `formulaDist`)
- **Survival structure** (`formulaEvent`, and optional baseline hazard controls like `basehaz`, `n_knots`)
- **Association structure** (`assoc` plus `transforms`)
- **Priors** (`priors = list(beta = ..., alpha = ..., lkj = ...)`)
- **Computation** (`control` for sampling, and optionally threading)

# Scenario A: Multivariate Continuous Markers with CV Association

## Data simulation

We simulate a multi-marker longitudinal dataset with a survival process linked through **current value (CV)**. The simulator also returns truth values for fixed effects and survival coefficients.

```{r scenario-a-sim}
sim <- simulate_joinme_joint_student_t_cvtotal(
  n_id = 100,
  D = 5,
  n_t = 10,
  seed = 123,
  include_marker_only = TRUE
)
```

## Model specification and fit

We start with identity transforms and a conservative sampling configuration to keep runtime manageable. The nested syntax is to distinguish ourselves from `lme4` nested random effects. In their approach, the random effects 

```{r scenario-a-fit}
formulaLong <- y ~ 1 + time + x1 +
  (1 + time | id) +
  (0 + x1 + (1 + time | id) | marker)

formulaEvent <- survival::Surv(time, event) ~ 1 + x1 + x2

fit <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  families = rep('student_t',5),
  transforms = list(
    cv_total = list(type = "identity")
  ),
  priors = list(
    beta = NULL,
    alpha = NULL,
    lkj = NULL
  ),
  control = list(
    parallel_chains = 2,
    iter_warmup = 300,
    iter_sampling = 700,
    seed = 123,
    refresh = 100,
    adapt_delta = 0.75,
    max_treedepth=12,
    init = 1
    # engine = 'rstan'
  )
)
```

To customize priors or the baseline hazard basis, supply `priors = ...` and/or forward standata arguments via `...`. Currently coefficients are fit by student_t with 6 degrees of freedom. Only scales are allowed for the moment but df may be supported, who knows o__o)//

```{r scenario-a-fit2, eval=FALSE}
fit2 <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  transforms = list(cv_total = list(type = "identity")),
  priors = list(beta = list(scale = 1.5), alpha = list(scale = 1.5), lkj = 2),
  basehaz = "bs",
  n_knots = 5,
  basehaz_degree = 3,
  control = list(iter_warmup = 500, iter_sampling = 500, adapt_delta = 0.9)
)
```

## Diagnostics and parameter interpretation

We extract diagnostics (R-hat, ESS, divergences) and compare posterior summaries to known truth values. This yields **bias** and **coverage** diagnostics for fixed effects and survival coefficients.

```{r scenario-a-diagnostics}
sum_obj <- summary(fit)
print(sum_obj)

# Sampler diagnostics
diag_tbl <- as.data.frame(t(unlist(sum_obj$diagnostics)))
knitr::kable(diag_tbl, digits = 3, caption = "Sampler diagnostics (scenario A).")

# Fixed effects (beta)
fixef_tbl <- sum_obj$tables$fixef
truth_beta <- sim$truth$beta

fixef_tbl$truth <- truth_beta[fixef_tbl$term]
fixef_tbl$bias <- fixef_tbl$Estimate - fixef_tbl$truth
fixef_tbl$covered <- fixef_tbl$truth >= fixef_tbl$Q2.5 &
  fixef_tbl$truth <= fixef_tbl$Q97.5

fixef_view <- fixef_tbl[, c("term", "Estimate", "Q2.5", "Q97.5", "truth", "bias", "covered")]
knitr::kable(fixef_view, digits = 3, caption = "Fixed effects summary (scenario A).")

# Survival covariates (gamma_w)
if (!is.null(sum_obj$tables$gamma_w)) {
  gamma_tbl <- sum_obj$tables$gamma_w
  truth_gamma <- sim$truth$gamma_w

  gamma_tbl$truth <- truth_gamma[gamma_tbl$term]
  gamma_tbl$bias <- gamma_tbl$Estimate - gamma_tbl$truth
  gamma_tbl$covered <- gamma_tbl$truth >= gamma_tbl$Q2.5 &
    gamma_tbl$truth <= gamma_tbl$Q97.5

  gamma_view <- gamma_tbl[, c("term", "Estimate", "Q2.5", "Q97.5", "truth", "bias", "covered")]
  knitr::kable(gamma_view, digits = 3, caption = "Survival covariate summary (scenario A).")
}

# Association (aggregate CV total)
if (!is.null(sum_obj$tables$assoc)) {
  assoc_tbl <- sum_obj$tables$assoc
  alpha_total_hat <- sum(
    assoc_tbl$Estimate[
      assoc_tbl$term %in% c("alpha_cv_mean", "alpha_cv_marker")
    ]
  )
  alpha_total_bias <- alpha_total_hat - sim$truth$alpha_cv_total
  list(alpha_total_hat = alpha_total_hat, alpha_total_bias = alpha_total_bias)
}
```

## RSS for fitted trajectories

The `JoinMeDynPred` object includes fitted values for observed history. We compute a **residual sum of squares** (RSS) on the expected-response scale (`epred`).

```{r scenario-a-rss}
pred_fit <- posterior_epred(
  fit,
  newdataLong = sim$dataLong,
  newdataEvent = sim$dataEvent,
  Tstart = max(sim$dataLong$time),
  times = seq(0, max(sim$dataLong$time) + 1, length.out = 60),
  n_samples = 100,
  control = list(
    iter_sampling = 500,
    iter_warmup = 400,
    chains = 3,
    threads_per_chain = 5
  )
)

fitted_ep <- pred_fit$predictions$longitudinal_fitted

if (!requireNamespace("dplyr", quietly = TRUE)) {
  stop("Install dplyr to run RSS calculation.")
}

library(dplyr)

rss_tbl <- fitted_ep %>%
  filter(scale == "epred") %>%
  inner_join(
    sim$dataLong,
    by = c("id", "marker", "time")
  ) %>%
  summarise(RSS = sum((y - Estimate)^2, na.rm = TRUE))

knitr::kable(rss_tbl, digits = 3, caption = "RSS for fitted trajectories (scenario A).")
```

## Dynamic prediction and plotting

We generate a future prediction for one subject and plot longitudinal and survival outputs.

```{r scenario-a-predict-plot}
#| fig-cap: "Dynamic prediction (scenario A)."
id <- 1
ndL <- sim$dataLong[sim$dataLong$id == id, ]
ndE <- sim$dataEvent[sim$dataEvent$id == id, ]

Tstart <- max(ndL$time)
times <- seq(Tstart, Tstart + 2, length.out = 50)

pred_ep <- posterior_epred(
  fit,
  newdataLong = ndL,
  newdataEvent = ndE,
  Tstart = Tstart,
  times = times,
  n_samples = 200,
  control = list(
    iter_sampling = 500,
    iter_warmup = 400,
    threads_per_chain = 5,
    chains = 2
  )
)

plot(pred_ep, which = "longitudinal")
plot(pred_ep, which = "survival")
```

Dynamic prediction is conditional on observed history up to `Tstart` and (by construction) survival up to that landmark; the survival plot corresponds to $\Pr(T > t \mid T > T_{start}, \mathcal{H}(T_{start}))$.

# Extension Mixed Families and Nonlinear Association

This example uses **mixed families** and a **nonlinear association transform** to illustrate interpretability when markers live on distinct scales. Very experimental and tend to bug out so if you are interested, *joinme*

```{r scenario-b-mixed}
#| eval: false

sim_mixed <- simulate_joinme(
  n_id = 20,
  n_obs_per_marker_per_id = 5,
  families = c("gaussian", "bernoulli", "poisson"),
  times_obs = seq(0, 4, length.out = 5),
  seed = 101
)

formulaLong_mixed <- y ~ 1 + time + x1 +
  (1 + time | id) +
  (0 + x1 + (1 + time | id) | marker)

formulaEvent_mixed <- survival::Surv(time, event) ~ 1 + x1 + x2

transforms <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.3, 0.7, 1.1, 1.3, 1.4),
    degree = 3
  )
)

# Optional: distributional regression

If a marker family supports additional parameters (e.g., Student-$t$ scale or degrees of freedom, negative binomial overdispersion), you can model these via `formulaDist`. Example (heteroscedastic scale):

```{r scenario-b-dist}
#| eval: false
fit_dist <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  transforms = list(cv_total = list(type = "identity")),
  families = rep("student_t", 4),
  formulaDist = list(sigma = ~ 1 + time),
  control = list(iter_warmup = 300, iter_sampling = 300, seed = 123, refresh = 0)
)
```

fit_mixed <- joinme(
  formulaLong = formulaLong_mixed,
  dataLong = sim_mixed$dataLong,
  formulaEvent = formulaEvent_mixed,
  dataEvent = sim_mixed$dataEvent,
  families = sim_mixed$marker_info$families,
  transforms = transforms,
  assoc = c("cv_total"),
  control = list(
    parallel_chains = 1,
    iter_warmup = 200,
    iter_sampling = 200,
    seed = 101,
    refresh = 0,
    adapt_delta = 0.95
  )
)

sum_mixed <- summary(fit_mixed)
diag_mixed <- as.data.frame(t(unlist(sum_mixed$diagnostics)))
knitr::kable(diag_mixed, digits = 3, caption = "Sampler diagnostics (scenario B).")

if (!is.null(sum_mixed$tables$fixef)) {
  fixef_mixed <- sum_mixed$tables$fixef
  fixef_mixed <- fixef_mixed[, c("term", "Estimate", "Q2.5", "Q97.5", "Rhat", "ESS")]
  knitr::kable(fixef_mixed, digits = 3, caption = "Fixed effects summary (scenario B).")
}
```

# Scenario C: Threaded Fit and Threaded Prediction

Threading is useful when the longitudinal dimension is large. This example shows a threaded fit and dynamic prediction using two threads.

```{r scenario-c-threaded}
#| eval: false
sim_thread <- simulate_joinme_joint_student_t_cvtotal(
  n_id = 60,
  D = 6,
  n_t = 6,
  seed = 202,
  include_marker_only = TRUE
)

fit_thread <- joinme(
  formulaLong = formulaLong,
  dataLong = sim_thread$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim_thread$dataEvent,
  assoc = c("cv_total"),
  transforms = list(
    cv_total = list(type = "identity")
  ),
  control = list(
    parallel_chains = 1,
    iter_warmup = 200,
    iter_sampling = 200,
    seed = 202,
    refresh = 0,
    threads_per_chain = 2,
    adapt_delta = 0.9
  )
)

ndL_thread <- sim_thread$dataLong[sim_thread$dataLong$id %in% c(1, 2), ]
ndE_thread <- sim_thread$dataEvent[sim_thread$dataEvent$id %in% c(1, 2), ]

pred_thread <- posterior_epred(
  fit_thread,
  newdataLong = ndL_thread,
  newdataEvent = ndE_thread,
  Tstart = max(ndL_thread$time),
  times = seq(0, max(ndL_thread$time) + 2, length.out = 60),
  n_samples = 100,
  control = list(
    threads_per_chain = 2,
    chains = 1,
    iter_warmup = 100,
    iter_sampling = 1,
    seed = 202,
    refresh = 0
  )
)

plot(pred_thread, which = "longitudinal")
plot(pred_thread, which = "survival")
```

# Discussion and Reporting Guidance

These scenarios illustrate how model assumptions and association choices influence interpretation. In a report or manuscript, we recommend:

- Reporting **R-hat**, **ESS**, and divergences as primary convergence diagnostics.
- Presenting **bias and coverage** for simulated truth when validation is possible.
- Summarizing **association parameters** alongside the transform used (functional, spline, piecewise).
- Including **dynamic prediction plots** to show how uncertainty evolves beyond the observed history.
- Stating whether **threading** was used and the number of threads per chain.

This structure provides a reproducible, interpretable workflow that can be adapted to substantive scientific questions.
