---
title: "Worked latent-progress analyses with JoiNMe"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: truex
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(joinme)
library(survival)
set.seed(20260727)
```

# Aim

This worked example develops four analyses from one simulated multivariate
study:

1. shared subject progression;
2. combined subject and covariance progression;
3. marker effects combined with unshared marker weights; and
4. a longitudinal-only sensitivity analysis.

The code is not evaluated while the package vignette is built because the
models require Hamiltonian Monte Carlo. It is intended to be run interactively
with an installed CmdStan toolchain.

# Simulate a modest study

```{r}
simulated <- simulate_joinme(
  n_id = 120,
  families = c("gaussian", "student_t", "gaussian"),
  n_obs_per_marker_per_id = 5,
  times_obs = seq(0, 4, length.out = 9),
  assoc = c("cv_mean", "cv_total", "vcov"),
  assoc_coefs = c(cv_mean = 0.25, cv_total = 0.18, vcov = 0.12),
  seed = 20260727
)

data_long <- simulated$dataLong
data_event <- simulated$dataEvent
```

The nested longitudinal formula contains a shared subject intercept and
slope, a marker-level treatment deviation, and an subject-by-marker
intercept and slope:

```{r}
longitudinal_formula <-
  y ~ 1 + time + x1 +
  (1 + time | id) +
  (0 + x1 + (1 + time | id) | marker)

event_formula <- Surv(time, event) ~ 1 + x1 + x2
covariance_formula <- ~ 1 + x1
```

# Establish the ordinary model

Before introducing a mixture, fit the corresponding ordinary model:

```{r}
ordinary_fit <- joinme(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  formulaEvent = event_formula,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  assoc = c("cv_mean", "cv_total", "vcov"),
  control = list(
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1000,
    iter_sampling = 1000,
    adapt_delta = 0.95
  ),
  seed = 20260727
)

diagnosis(ordinary_fit)
pp_check(ordinary_fit)
```

This step checks the longitudinal family choices, covariance regression,
association transforms and event design without asking a mixture to compensate
for a poorly specified base model.

# Analysis 1: shared subject progression

Fit three classes on the standardised subject intercept and slope:

```{r}
subject_fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  formulaEvent = event_formula,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  n_classes = 3,
  class_type = "subject",
  class_dimensions = c(1, 2),
  formulaClass = ~ x1 + x2,
  priors = jm_prior(
    class_probability = rep(2, 3),
    class_regression = prior_normal(scale = 1)
  ),
  assoc = c("cv_mean", "cv_total", "vcov"),
  shrinkage = 0,
  control = list(
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1500,
    iter_sampling = 1500,
    adapt_delta = 0.97,
    max_treedepth = 13
  ),
  seed = 20260727
)
```

Inspect the complete summary and the allocation probabilities:

```{r}
summary(subject_fit)
subject_membership <- posterior_class(subject_fit)
subject_membership
plot(subject_fit, type = "class_membership")
```

The maximum-probability class can be obtained from `assigned_class`, but the
probabilities should be retained in any secondary analysis.

Compare class-centre and within-class marginal trajectories:

```{r}
longitudinal_plot(
  subject_fit,
  estimand = "mean_per_class",
  draws = 800
)

longitudinal_plot(
  subject_fit,
  estimand = "marginal_per_class",
  draws = 800
)
```

For a Gaussian identity-link marker, centre and marginal curves should be
close. A non-linear inverse link makes their distinction more apparent.

Display the `cv_mean` hazard contribution along each class trajectory:

```{r}
association_plot(
  subject_fit,
  estimand = "mean_per_class",
  association_term = "cv_mean",
  draws = 800
)
```

# Analysis 2: subject and covariance progression

Now combine the shared subject process with the covariance-regression
latent process:

```{r}
subject_process_fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  formulaEvent = event_formula,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  n_classes = 3,
  class_type = c("subject", "vcov"),
  class_dimensions = list(
    subject = c(1, 2),
    vcov = c(1, 2)
  ),
  assoc = c("cv_mean", "vcov"),
  priors = jm_prior(class_probability = rep(2, 3)),
  shrinkage = 0,
  control = list(
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1800,
    iter_sampling = 1800,
    adapt_delta = 0.98,
    max_treedepth = 14
  ),
  seed = 20260728
)
```

There remain three classes. Within an subject, `z_u` and `z_L` use the same
component label:

```{r}
subject_process_fit$mixture$allocation_domains
posterior_class(subject_process_fit)
```

Plot the covariance profiles:

```{r}
covariance_plots <- plot(
  subject_process_fit,
  type = "covariance_class",
  draws = 800
)
covariance_plots
```

Every returned marker plot contains all three classes in each covariance
component panel. The marker-specific plots make the shared covariance
construction explicit without separating the classes across figures.

For a covariance association, request the same class-specific display through
the association interface:

```{r}
association_plot(
  subject_process_fit,
  estimand = "mean_per_class",
  association_term = "vcov[1]"
)
```

# Analysis 3: marker effects and unshared weights

Suppose current value and current slope require distinct marker weighting
structures:

```{r}
base_weights <- list(
  cv_total = c(marker_1 = 1.0, marker_2 = 0.6, marker_3 = -0.2),
  cs_marker = c(marker_1 = 0.2, marker_2 = 1.0, marker_3 = 0.5)
)

marker_process_fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  formulaEvent = event_formula,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  n_classes = 2,
  class_type = "marker",
  assoc = c("cv_total", "cs_marker"),
  marker_weights = base_weights,
  fixed_marker_weights = FALSE,
  shared_marker_weights = FALSE,
  shrinkage = 1,
  priors = jm_prior(class_probability = c(2, 2)),
  control = list(
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1500,
    iter_sampling = 1500,
    adapt_delta = 0.97
  ),
  seed = 20260729
)
```

The component vector contains two marker-effect coordinates followed by both
weight-set coordinates. It still defines two marker classes:

```{r}
marker_process_fit$mixture
posterior_class(marker_process_fit)

association_plot(
  marker_process_fit,
  estimand = "mean_per_class",
  association_term = "cv_total"
)

association_plot(
  marker_process_fit,
  estimand = "mean_per_class",
  association_term = "cs_marker"
)
```

The class-specific marker-weight locations enter these plots before the marker
average, matching the fitted survival likelihood.

# Analysis 4: longitudinal-only sensitivity analysis

Remove the event process without changing the longitudinal formula:

```{r}
longitudinal_only_fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  formulaVCov = covariance_formula,
  n_classes = 3,
  class_type = c("subject", "vcov"),
  priors = jm_prior(class_probability = rep(2, 3)),
  shrinkage = 0,
  control = list(
    chains = 4,
    parallel_chains = 4,
    iter_warmup = 1500,
    iter_sampling = 1500
  ),
  seed = 20260730
)

summary(longitudinal_only_fit)
posterior_class(longitudinal_only_fit)
pp_check(longitudinal_only_fit)
longitudinal_plot(
  longitudinal_only_fit,
  estimand = "marginal_per_class"
)
```

The survival contribution is exactly absent. An event-only diagnostic is
therefore deliberately rejected:

```{r}
# Produces an explanatory error:
concordance(longitudinal_only_fit)
```

# Convergence review

Mixture parameters deserve their own trace review:

```{r}
mcmc_plot(
  subject_process_fit,
  variable = "mix_probability",
  type = "trace"
)

mcmc_plot(
  subject_process_fit,
  variable = "mix_location",
  type = "trace"
)

mcmc_plot(
  subject_process_fit,
  variable = "mix_scale",
  type = "trace"
)
```

Also review the random-effect Cholesky factors and covariance-regression
parameters. Stable mixture traces do not compensate for an unstable underlying
covariance model.

# Comparing plausible class counts

Fit neighbouring values with the same data, priors and sampling effort:

```{r}
fits_by_class_count <- lapply(2:4, function(n_classes) {
  joinme_mix(
    formulaLong = longitudinal_formula,
    dataLong = data_long,
    formulaEvent = event_formula,
    dataEvent = data_event,
    formulaVCov = covariance_formula,
    n_classes = n_classes,
    class_type = "subject",
    priors = jm_prior(class_probability = rep(2, n_classes)),
    assoc = c("cv_mean", "cv_total"),
    seed = 3100 + n_classes
  )
})
```

Compare more than one quantity: out-of-sample predictive measures where
available, posterior predictive checks, minimum effective class size,
allocation entropy, component separation and scientific interpretability.

# Suggested results table

A useful manuscript table has one row per class and includes:

- posterior mean class proportion and interval;
- effective number of high-probability units;
- component location and scale for every selected coordinate;
- a clinically interpretable trajectory contrast at pre-specified times;
- an association contribution or covariance contrast where relevant; and
- allocation entropy or another uncertainty summary.

The table should be accompanied by class-centre and marginal curves, not used
as a substitute for them.
