---
title: "joinme: Association Construction"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
options(width = 80)
```

# Purpose

This vignette focuses on **association term construction**, including transformation selection, interpretability, and practical trade-offs.

Association transforms apply to all supported longitudinal families (including
double_exponential, skew_double_exponential, beta, and cumulative_logit) and to
competing-risk survival models.

## Step-by-step: how association summaries are constructed

At the event time and at the default 15 Gauss-Kronrod nodes used for integration
(configurable), the
longitudinal linear predictor is decomposed into a mean component and a marker
component. From these components, `joinme` constructs:

- **CV_total**: total current value (mean + marker components).
- **CV_mean**: current value of the mean component only.
- **CV_marker**: current value of the marker component only.
- **CS_total**: total current slope, approximated by a forward finite difference
  and rescaled to the original time units.
- **CS_mean**: slope of the mean component.
- **CS_marker**: slope of the marker component.
- **CORR**: a subject-specific variance summary of the marker-by-id covariance
  structure, constant across time.

Each component can be included or excluded using the `assoc` argument. If
`assoc` does not include a component, the corresponding coefficient is prior-only
and does not enter the likelihood.

The CORR component requires marker-by-id random effects. If the marker block
does not include an inner `( ... | id )` term, then `Q_idm = 0` and CORR is not
available.

# Marker-weighted CV and CS

When multiple biomarkers are modeled, `joinme` constructs marker-average association
components using weights $\omega_d$ (one per marker). For marker-aggregated
terms (`cv_total`, `cv_marker`, `cs_total`, `cs_marker`), transforms are applied
at the marker level first and then averaged with weights. This matters whenever
transforms are nonlinear.

For example, with marker-specific total current values $CV_{id}(t)$ and transform
$f_{cv}$:

$$
  ext{cv\_total}(t) = \frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cv}\!\left(CV_{id}(t)\right),
$$

which differs from $f_{cv}(\frac{1}{D}\sum_d \omega_d CV_{id}(t))$ unless $f_{cv}$
is linear.

For slopes, let $CS_{id}(t)$ be a marker-specific slope feature and $f_{cs}$ the
configured CS transform. Then

$$
  ext{cs\_total}(t) = \frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cs}\!\left(CS_{id}(t)\right),
$$

and analogously for `cs_marker`. This is intentionally not equal to
$f_{cs}(\frac{1}{D}\sum_d \omega_d CS_{id}(t))$ unless $f_{cs}$ is linear.

In practice, provide fixed weights via `marker_weights`, or set
`fixed_marker_weights = FALSE` to estimate signed perturbations around base
weights via `marker_weights + z_marker_weights` with
`z_marker_weights ~ N(0, 1)`. The resulting weights are used directly in
marker-aggregated association terms. The simulator reports them in
`truth$marker_weights`.

# Association coefficients and naming

Association coefficients are denoted with the $\alpha$ prefix (e.g., `alpha_cv_total`).
This follows joint-model conventions and avoids confusion with the linear predictor
$\eta$ used throughout the longitudinal and survival submodels.
Total, mean, and marker associations use a positive non-centered parameterization
$\alpha = z_{\alpha} \cdot sd_{\alpha}$ for stable sampling.

When an association component is not selected in `assoc`, its coefficient is
prior-only and does not enter the likelihood. Summaries only report active
components to keep interpretation focused on the fitted association structure.

# Transform modes

`joinme` supports four transformation modes for association features:

- **Identity** (`type = "identity"`): no transformation.
- **Functional** (`type = "functional"`): a user-defined expression converted
  into a fixed bytecode program.
- **Monotone I-spline** (`type = "ispline"` or `"ispline_penalized"`): a smooth
  monotone mapping.
- **Piecewise linear** (`type = "pwlin"`): linear interpolation across knots.

# Functional Transform

```{r functional-transform, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "functional",
    expr = ~ sinh(x) + tanh(x)
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```

The functional evaluator supports common arithmetic and functions such as
`log`, `exp`, `sqrt`, `sin`, `cos`, `tanh`, plus aliases `sigmoid`/`expit`,
`softplus`/`log1p_exp`, and helpers like `cbrt` and `power`.

```{r functional-transform-mixed-demo, eval=FALSE}
transforms <- list(
  cv_total = list(type = "functional", expr = ~ inv_logit(x)),
  cv_mean = list(type = "functional", expr = ~ exp(0.2 * x)),
  cv_marker = list(type = "functional", expr = ~ power(x, 2)),
  cs_total = list(type = "functional", expr = ~ sqrt(abs(x) + 1e-6)),
  corr = list(type = "functional", expr = ~ cbrt(x))
)
```

**When to use**

- You want a fixed functional form with no spline fitting.
- You need fast reconfiguration without recompiling Stan.

# Monotone I-Spline

```{r monotone-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.5, 1, 1.2, 1.5, 1.6),
    degree = 3
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```

**When to use**

- You want monotone, smooth effects with constrained shape.

# Penalized Monotone I-Spline

```{r penalized-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "ispline_penalized",
    x = seq(-2, 2, length.out = 40),
    y = exp(seq(-2, 2, length.out = 40)),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

**When to use**

- You want a monotone transform learned from (x, y) pairs with smoothing.
- Penalization avoids overfitting when knot counts are moderate.

# Piecewise-Linear

```{r piecewise-linear, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "pwlin",
    x = c(-2, -1, 0, 1, 2),
    y = c(0.2, 0.5, 1, 0.5, 0.2)
  )
)
```

**When to use**

- You need simple, interpretable transformations.

# Association Diagram

```{mermaid}
%%| fig-cap: "Association components mapped into the survival predictor."
graph TD
  CV[CV total] --> FCV[f_cv]
  CS[CS total] --> FCS[f_cs]
  V[Corr] --> FCORR[f_corr]
  FCV --> ETA[eta assoc]
  FCS --> ETA
  FCORR --> ETA
```

# Hierarchical Covariance Regression

```{mermaid}
%%| fig-cap: "Covariance regression structure for marker-by-id effects."
graph TD
  X[Covariates] --> LP[Linear predictor]
  UL[Latent effect] --> LP
  LP --> L[Cholesky factor]
  L --> W[Marker-by-id effects]
  Z[Latent z] --> W
```

# Fit with Custom Association

```{r fit-custom-association, eval=FALSE}
fit <- joinme(
  formulaLong = y ~ 1 + time + x1 +
    (1 + time || id) +
    (0 + x1 + (1 + time || id) || marker),
  dataLong = sim$dataLong,
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  dataEvent = sim$dataEvent,
  transforms = transforms,
  assoc = c("cv_total", "corr")
)
```
