---
title: "A Example Workflow for Joint Modelling with mutlivariate Longitudinal and Survival Data using joinme"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
options(width = 100)
# Ensure the package is available for the vignette to run if possible
repo_root <- normalizePath(if (dir.exists("R")) "." else "..")
if (!require('joinme', quietly = TRUE)) {
  if (requireNamespace("devtools", quietly = TRUE) && file.exists(file.path(repo_root, "DESCRIPTION"))) {
    devtools::load_all(repo_root, quiet = TRUE)
  } else {
    try(base::attachNamespace("joinme"), silent = TRUE)
  }
}
library(dplyr)
library(ggplot2)
library(survival)
library(bayesplot)
library(posterior)
```

# Introduction

This vignette outlines a workflow for using `joinme` in a research context, suitable for inclusion in high-impact statistical or clinical journals.

In this document, we will be
1.  Generating synthetic data with known ground truth.
2.  Fitting the joint model.
3.  Assessing convergence and model fit with renamed posterior draws.
4.  Performing dynamic predictions for individual subjects.
5.  Reporting results with consistent summaries and plotting helpers.

## Program flow in practice

The `joinme` package follows the below data-processing and model-fitting sequence:

1. **Data assembly (R)**: build design matrices, scale time to $[0, 1]$,
  and prepare Gauss-Kronrod nodes (default 15, configurable) for survival integration.
2. **Parameter declarations (Stan)**: define fixed effects, random effects,
  distributional regression coefficients, baseline hazard parameters, and
  association coefficients.
3. **Transformed parameters (Stan)**: rescale time-related coefficients and
  construct the random effects used in the likelihood.
4. **Model block (Stan)**: apply priors and evaluate the likelihood using
  `reduce_sum`, which splits subjects across threads when requested.
5. **Generated quantities (Stan)**: compute per-observation log-likelihoods,
  fitted values, and prediction summaries for posterior diagnostics.
6. **Post-processing (R)**: expose the posterior through `draws()`,
  summary helpers, `mcmc_plot()`, and explicit plotting wrappers such as
  `longitudinal_plot()` and `survival_plot()`.

This sequence is important because it clarifies where each modelling choice
enters the inference: formula parsing happens in R, while all probability
calculations occur in Stan.

# Simulation and Data Exploration

Before analysing real data, it is highly recommended to perform a **parameter recovery simulation**. This confirms that the model is identifiable given your sample size and sampling frequency.

We use the built-in generalised simulator `simulate_joinme` to generate a dataset where the hazard depends on the "Total Current Value" of the biomarkers.

```{r simulate}
set.seed(2025)

# Simulate a dataset: 
# - 100 subjects
# - 3 biomarkers
# - ~10 observations per marker per subject
sim <- simulate_joinme(
  n_id = 40,
  families = list(
    jm_family("student_t"),
    jm_family("student_t"),
    jm_family("student_t", inv_link = ~ inv_logit(x / 2))
  ),
  n_obs_per_marker_per_id = 6,
  times_obs = seq(0, 8, length.out = 12),
  quadrature_nodes = 31,
  beta_long = c("(Intercept)" = 1.0, "time" = 0.5, "x1" = 0.4),
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.5),
  fixed_marker_weights = FALSE,
  shrinkage = 0L
)

# Gauss-Kronrod nodes/weights are fixed in Stan; `quadrature_nodes` selects
# the node count used to build the design matrices in R.
#
# For a probit marker, use link = "probit" or link = ~ inv_Phi(x).
# A direct inverse-link declaration instead uses inv_link = ~ Phi(x).

# `truth$marker_weights` is the effective base + latent value used in the hazard.
# The decomposition is available as marker_weights_base/marker_weights_latent.

dataLong <- sim$dataLong
dataEvent <- sim$dataEvent

# Optional: generate counting-process event rows with delayed entry.
sim_split <- simulate_joinme(
  n_id = 20,
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x1 + x2,
  covariate_formulas = list(
    x1 ~ time_varyring(rnorm, c(2, 5), mean = 0, sd = 1),
    x2 ~ rnorm(n_id)
  ),
  left_truncation_max = 1.0,
  seed = 2026
)
dataEvent_split <- sim_split$dataEvent

# Preview the data
head(dataLong)
head(dataEvent)
```

Visualise your longitudinal trajectories and survival distribution before modelling.

```{r explore-plot}
# Visualise trajectories for a subset of IDs
subset_ids <- base::sample(unique(dataLong$id), 6)

ggplot(dataLong %>% filter(id %in% subset_ids), aes(x = time, y = y, color = marker, group = marker)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 1) +
  facet_wrap(~id) + 
  theme_bw() +
  labs(title = "Longitudinal Trajectories (Subset)", x = "Time", y = "Biomarker Value")

# Check event rate
table(dataEvent$event)
prop.table(table(dataEvent$event))
```

Ensure the event rate is sufficient (typically >10-20 events per parameter in the survival submodel) to support the complexity of the proposed association structure.

# Model Specification

We specify a joint model where:
1.  **Longitudinal Submodel**: Linear growth over time with random intercepts and slopes for each subject-marker combination.
2.  **Survival Submodel**: The hazard depends on baseline covariates (`x1`, `x2`) and the current value (`cv_total`) of the biomarkers.

### Families and distributional regression

Each marker can follow a distinct family (for example Gaussian, Student-$t$,
Poisson, negative binomial, Bernoulli, beta, or ordinal). When a family includes
additional parameters (for example $\sigma$, $\nu$, or $\phi$), those parameters
can be modelled through `formulaDist`. This allows heteroscedasticity or
covariate-dependent dispersion while keeping the mean structure aligned across
markers.

### Association transformations

The association terms can be transformed using one of four modes: identity,
functional bytecode expressions, monotone I-splines, or ordered piecewise-linear
interpolation. These transformations are specified in R and passed to Stan as
data so that the likelihood remains deterministic and reproducible.

## Marker-only random effects (no inner id term)

In some studies, we may want **marker-level random effects only**, without subject-specific marker deviations. This is done by omitting the inner `( ... | id )` term inside the marker block. In this configuration, marker-by-id random effects are disabled (internal standata `Q_idm = 0`), and covariance-style associations that require those effects (`corr`, `vcov`) are not permitted. Note: assoc = c("corr") or assoc = c("vcov") is not allowed when Q_idm = 0.

```{r marker-only-example, eval=FALSE}
# Marker-only random effects: no inner ( ... | id )
formulaLong_marker_only <- y ~ time + x1 +
  (1 + time || id) +
  (0 + x1 || marker)
```

Scales of the parameters can be set via `joinme_priors()`, which allows explicit control over the prior distributions for fixed effects, association coefficients, and correlation matrices. Stricter shrinkage can be set via `shrinkage` params.
```{r priors}
# Define explicit priors for transparency (optional but recommended)
# Note: JoiNMe exposes beta (fixed effects), alpha (associations), and lkj
priors <- joinme_priors(
  beta = list(scale = 2.5),       # Fixed effects (Student-t, df = 6)
  alpha = list(scale = 1.0),      # Association coefficients
  lkj = 2.0                       # Correlation matrices (regularizing)
)
```

# Model Fitting

We fit the model using Hamiltonian Monte Carlo (HMC) via Stan. By default, `joinme` prefers `cmdstanr` if compiled, otherwise it falls back to `rstan`. The control parameters can be adjusted using `cmdstanr` syntax.

```{r fit}
has_cmdstan <- requireNamespace("cmdstanr", quietly = TRUE) &&
  !is.null(tryCatch(cmdstanr::cmdstan_version(error_on_NA = FALSE), error = function(e) NULL))
engine <- if (has_cmdstan) "cmdstanr" else "rstan"

control <- list(
  engine = engine,
  chains = 1,           # Use 4 chains for publication
  iter_warmup = 200,    # Typically 1000+
  iter_sampling = 200,  # Typically 1000+
  parallel_chains = 1,
  adapt_delta = 0.90,   # Increase if divergences occur
  max_treedepth = 12
)


fit <- joinme(
  dataLong = dataLong,
  dataEvent = dataEvent,
  # Longitudinal formula: Fixed effects + Random effects
  formulaLong = y ~ time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  # Survival formula: Baseline covariates
  formulaEvent = survival::Surv(time, event) ~ x1 + x2,
  # Association structure: Link hazard to the total current value of markers
  assoc = c("cv_total"),
  transforms = joinme_tf(cv_total = "identity"),
  priors = priors,
  control = control
)

# Counting-process fit with left truncation/time-split covariates:
fit_split <- joinme(
  dataLong = sim_split$dataLong,
  dataEvent = dataEvent_split,
  formulaLong = y ~ time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x1 + x2,
  assoc = c("cv_total"),
  control = control
)
```

# Convergence Diagnostics

We use the `posterior` and `bayesplot` packages to inspect the MCMC chains.

## R-hat and Effective Sample Size

Values of $\hat{R} < 1.01$ and ESS > 400 indicate reasonable convergence for inference, as recommended by Stan developers. 

```{r diagnostics}
# robust summary of regression coefficients on the renamed user-facing scale
draws_obj <- draws(fit, format = "draws_array")

draws_sub <- posterior::subset_draws(
  draws_obj,
  variable = c("^\\(Intercept\\)$", "^time$", "^event", "^cv_total"),
  regex = TRUE
)

posterior::summarise_draws(
  draws_sub,
  default_summary_measures(),
  default_convergence_measures()
)
```

## Traceplots

Visual inspection of traceplots helps identify mixing issues or stuck chains. Note that the marker-specific association term for the first association (if marker weights are shared) is forced to be positive. If `shared_marker_weights == FALSE`, then the marker-specific association terms are all constrained to be positive.

```{r trace}
# Plot traces for the association parameter using the helper API
mcmc_plot(fit, variable = "cv_total", type = "trace")
```

# Dynamic Prediction

One of the most powerful features of joint models is **dynamic prediction**: updating survival probabilities as new biomarker data becomes available.

If competing risks are present, the dynamic survival and cumulative hazard
outputs correspond to the overall survival that integrates the sum of
cause-specific hazards. Cause-specific survival curves are not returned by the
default prediction outputs.

We select a subject who had an event at a later time and predict their survival probability conditionally on data observed up to an earlier time point.

```{r predict-setup}
# Select a subject with an event > 5
target_id <- dataEvent$id[which(dataEvent$time > 5 & dataEvent$event == 1)[1]]
if (is.na(target_id)) target_id <- dataEvent$id[1]

# Define the conditioning time (time_start)
t_cond <- 4

# You can also pass a column name in newdataEvent, e.g. time_start = "time_start".

# Filter history up to t_cond
ndLong <- dataLong |> filter(id == target_id, time <= t_cond)
ndEvent <- dataEvent |> filter(id == target_id)

print(paste("Predicting for ID:", target_id, "conditioned on history up to t =", t_cond))
```

Run the prediction:

```{r predict-run}
# Predict conditional survival from t_cond onwards
preds <- tryCatch(
  predict(
    fit,
    newdataLong = ndLong,
    newdataEvent = ndEvent,
    time_start = t_cond,
    times = seq(t_cond, 8, length.out = 30),
    process = "event", # Focus on survival and cumhaz
    control = list(progress = FALSE) # just to silence the progress bar
  ),
  error = function(e) {
    message("Prediction skipped: ", conditionMessage(e))
    NULL
  }
)

head(preds$predictions$survival)
```

`concordance(fit)` is a follow-up-wide survival-curve concordance rather than an
AUC at a selected horizon. By default, residual follow-up for subject \(i\)
starts at their final longitudinal measurement strictly before the observed
event or censoring time, \(T_{0i}\). For an event at residual time
\(r_i=T_i-T_{0i}\), the event subject is compared with every subject known to
survive longer. The pair is concordant when
\(\widehat S_i(r_i)<\widehat S_j(r_i)\): both conditional survival curves are
evaluated at the same earlier event time. This is the survival-curve
concordance proposed by Antolini and colleagues and does not use the outcome
time to construct a fixed subject score.

Premature censoring is handled through observability. A subject censored before
\(r_i\) is not comparable with that event; censoring at \(r_i\) establishes
survival through that time and remains comparable. `type_weights = "n/G2"`
requests Uno's inverse-censoring weighting through `survival::concordance()`.
Fitted monotone splines and ordered piecewise-linear associations enter every
conditional curve through their posterior ordinates, exactly as in
`predict()`.

`auc(fit, ...)` retains the horizon-specific estimand. At landmark \(s\) and
horizon \(t\), it compares cumulative risk
\(1-S_i(t\mid s)\) between cases observed by \(t\) and controls known to remain
event-free beyond \(t\). Subjects censored earlier have unknown case/control
status and are omitted. Dynamic prediction for both methods reuses the exact
fitted B-spline, natural-spline, or formula baseline-hazard basis, including
its original centring constants.

The estimands, censoring rules, event-time weighting, computational steps, and
reporting recommendations are derived in the
[dedicated discrimination vignette](joinme-discrimination.html).

```{r discrimination-example, eval=FALSE}
concordance(fit)
concordance(fit, time_start = 2, type_weights = "n/G2")
auc(fit, time_start = 2, time_horizon = 5)
```

Finally, we visualise the predicted survival curve. The shaded area represents the 95% credible interval, accounting for uncertainty in both the parameters and the random effects.

```{r predict-plot}
if (!is.null(preds)) {
  survival_plot(preds) +
    ggtitle(paste("Dynamic Survival Prediction for Subject", target_id)) +
    geom_vline(xintercept = dataEvent$time[dataEvent$id == target_id], linetype = "dashed", color = "red") +
    annotate("text", x = dataEvent$time[dataEvent$id == target_id], y = 0.1, label = "Observed event time", color = "red", hjust = 1.1)
}
```

# Posterior reporting helpers

The scientific reporting layer should use the same user-facing parameter names as the tables returned by `summary(fit)`. The `draws()` and `mcmc_plot()` helpers make that possible without directly handling raw Stan variables.

```{r posterior-reporting}
renamed_draws <- draws(fit, variables = c("time", "cv_total"), format = "draws_df")
head(renamed_draws)

mcmc_plot(fit, variable = c("time", "cv_total"), type = "intervals")
```

The plotting wrappers expose the same fitted and predicted views as the main `plot()` methods while making the intent explicit in scripts and reports:

```{r plot-wrappers, eval=FALSE}
longitudinal_plot(fit, subject = 1, scale = "epred")
association_plot(fit)
diagnostic_plot(fit, type = "ess_bulk")

if (!is.null(preds)) {
  longitudinal_plot(preds)
  survival_plot(preds)
  cumhaz_plot(preds)
}
```

For fitted longitudinal displays, the heatmap is now an alternative to the
usual separated marker curves:

```{r predict-heatmap, eval=FALSE}
plot(
  fit,
  type = "longitudinal",
  longitudinal_style = "heatmap",
  scale = "epred",
  threshold = 0.05
)
```

`plot(fit, ...)` uses fitted posterior samples directly and does not accept a conditioning argument. Named covariate profiles belong to the conditional- effects estimand. Build those profiles with `make_conditions()` and pass them to `conditional_effects()`:

```{r predict-conditions, eval=FALSE}
profiles <- make_conditions(dataEvent, vars = c("x1", "x2"))

effects_data <- conditional_effects(
  fit,
  effects = list(
    longitudinal = "time",
    event = c("x1", "x2")
  ),
  conditions = profiles,
  process = c("longitudinal", "event"),
  longitudinal_estimand = "population",
  plot = FALSE
)

plot(effects_data, ask = FALSE)
```

The longitudinal calculation has three estimands. `"population"` uses only the fixed-effect contribution. `"marker"` adds the fitted marker-level deviation while still excluding subject and marker-by-subject deviations.
`"marginal_marker"` calculates each selected marker response first and averages those responses within every posterior draw. The last ordering is important for nonlinear links: it estimates the average expected response rather than the expected response at an average marker effect. The `markers` argument restricts the marker-specific panels or the equally weighted set entering that average.

```{r predict-conditions-marker-average, eval=FALSE}
average_marker_effect <- conditional_effects(
  fit,
  effects = "time",
  process = "longitudinal",
  longitudinal_estimand = "marginal_marker",
  markers = c("m1", "m2"),
  plot = FALSE
)
```

The event tables contain the covariate-specific relative hazard
`exp(W * gamma)`, holding the baseline hazard and longitudinal association contribution fixed.
