---
title: "JoiNMe: Association Construction"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: false
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
options(width = 80)
```

# Purpose

This vignette provides a brief explanation of **association term construction**, including transformation selection and interpretability.

Association transforms apply to all supported longitudinal families (including double_exponential, skew_double_exponential, beta, and cumulative_logit) and to competing-risk survival models.

## Association term decomposition in Joint Nested Mixed-effects models

At the event time and at the default 15 Gauss-Kronrod nodes used for integration, the longitudinal linear predictor is decomposed into a mean component and a markercomponent. From these components, `joinme` (the package) constructs:

- **CV_total**: total current value (mean + marker components).
- **CV_mean**: current value of the mean component only.
- **CV_marker**: current value of the marker component only.
- **CS_total**: total current slope, approximated by a forward finite difference
  and rescaled to the original time units.
- **CS_mean**: slope of the mean component.
- **CS_marker**: slope of the marker component.
- **Corr**: a subject-specific lower-triangular matrix of the correlation of the marker-by-id covariance structure, constant across time.
- **vcov**: a subject-specific lower-triangular matrix of the marker-by-id Cholesky factor `L`, also constant across time.

# Marker-weighted CV and CS

When multiple biomarkers are modelled, `joinme` constructs marker-average association components using weights $\omega_d$ (one per marker). For marker-aggregated terms (`cv_total`, `cv_marker`, `cs_total`, `cs_marker`), transforms are applied at the marker level first and then averaged with weights. This matters whenever transforms are nonlinear.

For example, with marker-specific total current values $CV_{id}(t)$ and transform $f_{cv}$:

$$
\frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cv}\!\left(CV_{id}(t)\right),
$$

which differs from $f_{cv}(\frac{1}{D}\sum_d \omega_d CV_{id}(t))$ unless $f_{cv}$
is linear.

For slopes, let $CS_{id}(t)$ be a marker-specific slope feature and $f_{cs}$ the vonfigured CS transform. Then the association is

$$
\frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cs}\!\left(CS_{id}(t)\right),
$$

and analogously for `cs_marker`. This is intentionally not equal to $f_{cs}(\frac{1}{D}\sum_d \omega_d CS_{id}(t))$ unless $f_{cs}$ is linear.

In practice, place the entire declaration in
`jm_prior(marker_weights = ...)`. Use `family = "constant"` or `"none"` for fixed weights, or a stochastic family to fit a hierarchical marker-weight decomposition. For marker (d) in weight set (s),

$$
\omega_{sd}=\omega^{(0)}_{sd}+\mu_{\omega,s}+z_{sd}.
$$

Here $\omega^{(0)}_{sd}$ is the offset in `marker_weights$offset`, $\mu_{\omega,s}$ is a fitted common location, and $z_{sd}$ is a marker departure drawn directly from the selected centred unit-scale family.

Every marker attached to set (s) contributes likelihood information about $\mu_{\omega,s}$, so the common location is the marker-weight analogue of a population intercept. The common location uses `jm_prior(marker_weights = list(intercept = ...))`. The departures use `marker_weights$family`; their location and scale are fixed at zero and one.

A shared weight structure has one
(\mu_{\omega}) and one departure vector. With `marker_weights$shared = FALSE`, every active weighted association term has its own mean and departures.

With estimated weights and no supplied base, (\omega^{(0)}=0). In `simulate_joinme()` and `simulate_joinme_mix()`, a numeric `truth$marker_weights$intercept` supplies the true (\mu_{\omega,s}); a `prior_*()` declaration inside `jm_truth()` draws it once for each fitted set and then holds it fixed. The simulator stores the base, common mean, marker-specific departure, and effective truths in `truth$marker_weights_offset`, `truth$marker_weight_mean`, `truth$marker_weights_latent`, and `truth$marker_weights`. The final quantity is the effective weight used to generate survival times. The recovery call does not pass the true common mean to fitting, because it is an estimand.

In the log hazard, this feature is multiplied by the association slope (declared through `assoc$slope`). The `iota` intercept and slope belong to an optional affine transformation inside `tf()`; they are not the association multiplier.

## Plotting fitted association curves

```{r association-plot-helper, eval=FALSE}
association_plot(fit)
```

## Posterior association extraction and covariance-style labels

Use `posterior_assoc(fit, summary = TRUE)` (or `assoc(fit, summary = TRUE)`) to extract association-specific posterior summaries without requiring the full model summary. Use `summary = FALSE` to obtain posterior draws per association term.

## Renamed posterior draws for association terms

Association terms can be extracted directly on the reporting scale with `posterior_draws()`. The canonical posterior variables are `alpha_cv_total`, `alpha_cs_total`, `alpha_cv_mean`, `alpha_cs_mean`, `alpha_corr[...]`, and `alpha_vcov[...]`. Each is the coefficient actually used in the event hazard, which is also the scale reported by `summary()` and used by prediction andplotting methods.

```{r association-renamed-draws, eval=FALSE}
# Extracting association terms
assoc_draws <- posterior_draws(fit, variables = "^alpha_", regex = TRUE, format = "draws_df")
head(assoc_draws)

# Bayesplot visualisation with the same labels used by summary(fit)
mcmc_plot(fit, variable = "alpha_cv_total", type = "dens_overlay")
```

## Conditional effects in the two joint-model processes

Conditional effects are distinct from subject-specific dynamic prediction. `conditional_effects()` varies named predictors while holding other predictors at explicit profile values and can return the longitudinal process, the event process, or both. The longitudinal estimand is the population-level expected marker response by default. Set `longitudinal_estimand = "marker"` to retain the fitted marker-level deviation, or `"marginal_marker"` to average the marker-specific predictions within each posterior draw. Because marker averaging occurs after each inverse-link transformation, it targets an average expected response rather than an expected response at an average marker effect.

The event estimand is the direct relative-hazard component `exp(W * gamma)`; the baseline hazard and longitudinal association contribution are held fixed when making event-covariate comparisons.

```{r association-conditional-effects, eval=FALSE}
profiles <- make_conditions(fit$dataEvent, vars = "treatment")

ce <- conditional_effects(
  fit,
  effects = list(
    longitudinal = "time:treatment",
    event = "treatment"
  ),
  conditions = profiles,
  process = c("longitudinal", "event"),
  longitudinal_estimand = "marker",
  plot = FALSE
)

# The nested tables use the same estimate__, se__, lower__, upper__, and cond__
# columns as brms conditional effects.
head(ce$longitudinal[[1]])
head(ce$event[[1]])
plot(ce, ask = FALSE)
```

# Association coefficients and naming

Association coefficients are denoted by $\alpha$. This follows joint-model conventions and avoids confusion with the linear predictor $\eta$. Total, mean, marker, correlation, and covariance associations use the slope prior declared by `jm_prior(assoc = list(slope = ...))`. Its location and scale describe the effective coefficient entering the log hazard; there is no additional random association-scale multiplier. Marker-weight sign conventions are applied after this prior transformation where required for identification.

Summaries report the selected association components so that interpretation remains focused on the fitted association structure.

# Transform modes

`joinme` supports the following transformation families for association features:

- **Identity** (`type = "identity"`): no transformation.
- **Functional** (`type = "functional"`): a user-defined mathematical
  expression.
- **Monotone I-spline** (`type = "ispline"` or `"ispline_penalised"`): a smooth
  monotone mapping.
- **Bounded-domain monotone I-spline** (`type = "ispline_expit"` or `"ispline_expit_penalised"`): the same monotone spline machinery, but applied to `plogis(x)` rather than directly to `x`.
- **Ordered piecewise linear** (`type = "pwlin"`): posterior interpolation across fixed knots with estimated monotone knot ordinates.

## Default values and minimal parameterisations

- **Identity**:
  - Syntax: `list(type = "identity")`
  - Default when a transform term is omitted or set to `NULL`.
  - No other parameters are used.
- **Functional**:
  - Syntax: `list(type = "functional", expr = ~ log1p(x))`
  - Required field: `expr`.
  - No other defaults beyond the supplied expression.
- **I-spline**:
  - Syntax: `list(type = "ispline", knots = c(-1, 0, 1), coeff = c(0, 0.3, 0.8, 1.1, 1.3))`
  - Required fields: `knots`, `coeff`.
  - Default: `degree = 3` if omitted.
- **Penalised I-spline**:
  - Syntax:
    `list(type = "ispline_penalised", x = seq(-2, 2, length.out = 50), n_knots = 6, degree = 3, lambda = 1)`
  - Defaults: `n_knots = 6`, `degree = 3`, `lambda = 1`.
  - If `knots` are omitted when the curve is estimated, they are derived from `x` quantiles.
- **Expit-based penalised I-spline**:
  - Syntax:
    `list(type = "ispline_expit_penalised", x = seq(0.02, 0.98, length.out = 50), n_knots = 6, degree = 3, lambda = 1)`
  - Defaults: `n_knots = 6`, `degree = 3`, `lambda = 1`.
  - The raw association feature is mapped through `plogis(x)` at runtime, and the spline basis is built on that bounded domain. Supply training `x` and explicit `knots` directly on the expit scale in `[0, 1]` so the fitted spline domain matches the runtime basis.
- **Ordered piecewise linear**:
  - Syntax: `list(type = "pwlin", knots = c(-2, -1, 0, 1, 2), direction = "increasing")`.
  - Required field: `knots` (aliases: `cutpoints` and the earlier `x`).
  - Default: `direction = "increasing"`; use `"decreasing"` for a falling transform. Note that this is not important from the perspective of modelling.
  - An optional non-negative `lambda` penalises second differences of the ordered knot ordinates; it defaults to zero.

# Functional Transform

```{r functional-transform, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "functional",
    expr = ~ sinh(x) + tanh(x)
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```

The functional evaluator supports common arithmetic and functions such as `log`, `exp`, `sqrt`, `sin`, `cos`, `tanh`, plus aliases `sigmoid`/`expit`, `softplus`/`log1p_exp`, and helpers like `cbrt` and `power`.

```{r functional-transform-mixed-demo, eval=FALSE}
transforms <- list(
  cv_total = list(type = "functional", expr = ~ inv_logit(x)),
  cv_mean = list(type = "functional", expr = ~ exp(0.2 * x)),
  cv_marker = list(type = "functional", expr = ~ power(x, 2)),
  cs_total = list(type = "functional", expr = ~ sqrt(abs(x) + 1e-6)),
  corr = list(type = "functional", expr = ~ cbrt(x))
)
```

# Monotone I-Spline

For `type = "ispline"`, we specify the transform shape following `splines2::iSpline()` convention:

- `knots`: knot locations (first and last are boundary knots),
- `coeff`: non-negative I-spline basis coefficients,
- `degree`: spline degree (usually `3`).

Defaults:

- `degree = 3` if omitted.
- `knots` and `coeff` must be supplied explicitly.

For `type = "ispline_expit"`, the same direct specification applies, except the I-spline basis is evaluated on `plogis(x)`. The runtime association feature is still shown on its raw scale, but the transform's `knots` must be supplied on the expit scale because that is the spline domain.

This is highly expremental and meant for stabilising the boundary knots (with is usually unknown). You may want to standardise the longitudinal outcome so that most association would not squashed to 0 or 1, as `plogis(x)` is very flat outside of the range `[-5, 5]`. You may also want to put more knots as the extreme values of `plogis(x)` to overcome the flatness of the transform. 

```{r monotone-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.5, 1, 1.2, 1.5, 1.6),
    degree = 3
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```


# Penalised Monotone I-Spline

For `type = "ispline_penalised"`, there are two supported modes:

- **Simulation mode**: provide `x` and `y` to define a fixed monotone curve.
- **Modelling mode**: omit `y`, retain `knots` (or `n_knots`) plus `lambda`,
  and estimate the monotone spline jointly with the rest of the model.

Meaning of the arguments:

- `x`: raw association-feature values, used either as training inputs or to help choose knots.
- `y`: optional target transformed values at those `x` points. If present, you are telling the package the shape you want to approximate.
- `lambda`: smoothness penalty weight (higher = smoother, lower = more flexible).

Defaults:

- `n_knots = 6` if `knots` are omitted.
- `degree = 3` if omitted.
- `lambda = 1.0` if omitted.
- In helper-based plug-in fitting, `weights = NULL` means equal weights and
  `diff_order = 2` is the default penalty order.

```{r penalised-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "ispline_penalised",
    x = seq(-2, 2, length.out = 40),
    y = exp(seq(-2, 2, length.out = 40)),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

```{r penalised-ispline-stan, eval=FALSE}
transforms <- list(
  corr = list(
    type = "ispline_penalised",
    x = seq(-2, 2, length.out = 40),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

# Ordered Piecewise-Linear Association

For model fitting, the raw association feature is divided by fixed knots, but the association values at those knots are not supplied as an outcome-like `y` vector. With $K$ knots, the model estimates a $K-1$ simplex $\boldsymbol{\zeta}$. The relative transform ordinate at knot $j$ is

$$
f_j = s \sum_{r=1}^{j-1} \zeta_r, \qquad
s =
\begin{cases}
  1, & \text{increasing}, \\
  -1, & \text{decreasing}.
\end{cases}
$$

Thus $f_1=0$, $f_K=s$, and every adjacent difference has the requested sign. Linear interpolation is used between knots and the boundary ordinates are held constant outside the knot range. The survival contribution is $\alpha f(x)$: $\alpha$ estimates the total log-hazard span and the simplex estimates how that span is distributed across intervals. 

```{r  piecewise-linear, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "pwlin",
    knots = c(-2, -1, 0, 1, 2),
    direction = "increasing"
  )
)
```

# Association Diagram

```{mermaid}
%%| fig-cap: "Association components mapped into the survival predictor."
graph TD
  CV[CV total] --> FCV[f_cv]
  CS[CS total] --> FCS[f_cs]
  V[Corr] --> FCORR[f_corr]
  FCV --> ETA[eta assoc]
  FCS --> ETA
  FCORR --> ETA
```

# Hierarchical Covariance Regression

```{mermaid}
%%| fig-cap: "Covariance regression structure for marker-by-id effects."
graph TD
  X[Covariates] --> LP[Linear predictor]
  UL[Latent effect] --> LP
  LP --> L[Cholesky factor]
  L --> W[Marker-by-id effects]
  Z[Latent z] --> W
```

# Fit with Custom Association

```{r fit-custom-association, eval=FALSE}
fit <- joinme(
  formulaLong = y ~ 1 + time + x1 +
    (1 + time || id) +
    (0 + x1 + (1 + time || id) || marker),
  dataLong = sim$dataLong,
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  dataEvent = sim$dataEvent,
  transforms = transforms,
  assoc = c("cv_total", "corr", "vcov")
)
```
