---
title: "Independent standard-deviation and correlation regression"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

# Scientific purpose

Marker-by-subject random effects describe departures specific to one subject
and one longitudinal marker. Their covariance need not be constant across
subjects. A treatment may alter the variability of these departures without
altering their dependence, whilst a baseline physiological measurement may
alter dependence without changing marginal variability. A single covariance
formula cannot express these hypotheses independently.

JoiNMe therefore separates covariance regression into:

1. a standard-deviation regression, declared by `formulaVCov$sd`; and
2. an off-diagonal partial-correlation regression, declared by
   `formulaVCov$corr`.

The same declaration is consumed by `joinme()`, `joinme_mix()`,
`simulate_joinme()`, and `simulate_joinme_mix()`. The four entry points share
one parser, coefficient order, and covariance reconstruction.

# Formula interface

The explicit form is:

```{r}
covariance_formula <- list(
  sd = ~ treatment + age,
  corr = ~ baseline_score
)
```

A single formula is an exact shorthand for two identical formulae:

```{r}
formulaVCov = ~ treatment + age

# exactly equivalent to
formulaVCov = list(
  sd = ~ treatment + age,
  corr = ~ treatment + age
)
```

Both named components must be present when a list is used. Random-effect bars
and the longitudinal time variable are excluded because covariance predictors
are evaluated once per subject. Formula intercept columns are removed: every
covariance coordinate already has its own explicit intercept.

# Statistical model

Let $Q$ be the dimension of the marker-by-subject random-effect basis. For
subject $i$, JoiNMe reconstructs

$$
L_i=\operatorname{diag}(\boldsymbol\sigma_i)K_i,
$$

where $\boldsymbol\sigma_i=(\sigma_{i1},\ldots,\sigma_{iQ})^\top$ contains
marginal standard deviations and $K_i$ is a Cholesky correlation factor. The
marker-by-subject effect for marker $d$ is

$$
\boldsymbol w_{id}=L_i\boldsymbol z_{w,id},
\qquad
\boldsymbol z_{w,id}\sim N_Q(\boldsymbol0,I_Q).
$$

## Standard deviations

Let $\boldsymbol x_{sd,i}$ be the design row produced from `formulaVCov$sd`.
For coordinate $r=1,\ldots,Q$,

$$
\eta^{sd}_{ir}
=
\alpha^{sd}_r
+\boldsymbol x_{sd,i}^{\top}\boldsymbol\beta^{sd}_r
+\lambda_{m(r)}z_{L,i,m(r)}.
$$

The marginal standard deviation is

$$
\sigma_{ir}=g_{sd}(\eta^{sd}_{ir}),
$$

where $g_{sd}$ is softplus by default and may be changed to the exponential
link with `control = list(vcov_diag_link = "exp")`.

## Off-diagonal dependence

Let $\boldsymbol x_{corr,i}$ be the independent design row produced from
`formulaVCov$corr`. For row-major off-diagonal coordinate $h=(r,c)$, $c<r$,

$$
\eta^{corr}_{ih}
=
\alpha^{corr}_h
+\boldsymbol x_{corr,i}^{\top}\boldsymbol\beta^{corr}_h
+\lambda_{m(h)}z_{L,i,m(h)},
\qquad
a_{ih}=\tanh(\eta^{corr}_{ih}).
$$

The $a_{ih}$ are row-wise partial correlations. They are assembled as

$$
K_{i,rc}
=
\left\{\prod_{j<c}\sqrt{1-a_{i,(r,j)}^2}\right\}a_{i,(r,c)},
\qquad
K_{i,rr}
=
\prod_{j<r}\sqrt{1-a_{i,(r,j)}^2}.
$$

This guarantees a valid correlation matrix for every covariate profile and
posterior draw. The off-diagonal regression is absent when the nested
marker-by-subject term uses `||` and requests a diagonal covariance.

## Meaning of the latent loading

There is one non-negative scalar $\lambda_m$ for each packed covariance
coordinate. Collectively, $\boldsymbol\lambda_L$ is a vector, not a dense
matrix. Its action is

$$
\operatorname{diag}(\boldsymbol\lambda_L)\boldsymbol z_{L,i},
$$

so latent coordinate $m$ affects only predictor coordinate $m$. In an
ordinary model, $\lambda_m^2$ is the conditional residual variance of that
untransformed predictor. It is not itself a final standard deviation,
correlation, or covariance.

# Parameter order and priors

The lower-triangular latent vector remains packed row by row:

$$
(1,1),(2,1),(2,2),(3,1),(3,2),(3,3),\ldots.
$$

This stable order is used by `z_L`, `lambda_L`, `class_type = "vcov"`, and
covariance associations. The observed-covariate blocks have independent
orders:

- `vcov_sd`: all $Q$ intercepts, followed by the $Q\times K_{sd}$ slope
  matrix in row-major order;
- `vcov_corr`: all $Q(Q-1)/2$ intercepts in `(2,1), (3,1), (3,2), ...`
  order, followed by the correlation slope matrix in row-major order.

Different formula dimensions therefore create no unused or zero-padded
coefficients. Priors are declared independently:

```{r}
covariance_priors <- jm_prior(
  vcov_sd = prior_normal(mu = 0, scale = 1),
  vcov_corr = prior_horseshoe(
    global_scale = 0.25,
    slab_df = 4,
    slab_scale = 1
  )
)
```

Normal, fixed-degrees-of-freedom Student-$t$, Laplace, and regularised
horseshoe families are available. A scalar location or scale is repeated only
after the formula dimensions are known; a vector must match its packed block
exactly. The positive `lambda_L` loadings retain a half-Student-$t_6$ prior
because they describe unexplained heterogeneity rather than observed-
covariate regression coefficients.

# Fitting

```{r}
fitted <- joinme(
  formulaLong = y ~ treatment * time +
    (1 + time | id) +
    (1 + time + (1 + time | id) | marker),
  dataLong = data_long,
  formulaEvent = survival::Surv(event_time, event) ~ treatment,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  priors = covariance_priors,
  families = marker_families,
  assoc = "vcov"
)
```

`summary(fitted)` labels SD and correlation slopes separately using their own
model-matrix column names. Dynamic prediction evaluates both formulae for the
new subject and applies the same row-wise reconstruction.

The latent-progress entry point uses precisely the same covariance syntax:

```{r}
mixture_fit <- joinme_mix(
  formulaLong = y ~ treatment * time +
    (1 + time | id) +
    (1 + time + (1 + time | id) | marker),
  dataLong = data_long,
  formulaEvent = survival::Surv(event_time, event) ~ treatment,
  dataEvent = data_event,
  formulaVCov = covariance_formula,
  priors = covariance_priors,
  n_classes = 3,
  formulaClass = ~ treatment,
  class_type = "vcov"
)
```

The mixture acts on selected standardised `z_L` coordinates before the SD or
correlation link. It does not form a Cartesian product of class allocations
for the two regressions. `class_type = "corr"` selects off-diagonal
coordinates; `class_type = "vcov"` may select diagonal and off-diagonal
coordinates.

# Exact simulation-to-fit translation

The simulator uses the same formula object and component names:

```{r}
simulation <- simulate_joinme(
  formulaLong = y ~ treatment * time +
    (1 + time | id) +
    (1 + time + (1 + time | id) | marker),
  formulaEvent = survival::Surv(event_time, event) ~ treatment,
  formulaVCov = covariance_formula,
  priors = covariance_priors,
  re_params = list(
    id = list(sd = c(0.6, 0.2), corr = diag(2)),
    marker = list(sd = c(0.4, 0.15), corr = diag(2)),
    id_marker_cov = list(
      sd = list(
        alpha = c(-0.2, -0.4),
        beta = matrix(c(0.35, -0.10, 0.20, 0.05), nrow = 2),
        lambda = c(0.25, 0.20)
      ),
      corr = list(
        alpha = 0.15,
        beta = matrix(-0.30, nrow = 1),
        lambda = 0.20
      ),
      diag_link = "exp"
    )
  )
)

# Every fitting argument, including data, formulae and priors, is retained.
recovery_fit <- do.call(
  joinme,
  c(simulation$truth$recovery$arguments, list(seed = 401))
)
```

The generating values are recorded separately in
`truth$id_marker_cov_effective$sd` and
`truth$id_marker_cov_effective$corr`. Each record contains `alpha`, `beta`,
`lambda`, `z`, and its formula. `simulate_joinme_mix()` uses the same ordinary
generator and adds only the class distribution; its recovery record names
`joinme_mix` as the paired fitting entry point.

For automated coefficient recovery, the same values also appear under their
fitted names: `truth$stan_fit$alpha_L`, `beta_L_sd`, `beta_L_corr`, `lambda_L`,
and `z_L`. Thus a study can join generating values to posterior variable names
without reconstructing the lower-triangle or formula-specific packs itself.

# Recovery assessment

Recovery studies should compare like with like:

1. compare SD intercepts and slopes with the `SD[id:marker]` summary rows;
2. compare correlation intercepts and slopes with `K[id:marker]` rows;
3. compare `lambda_L` on the unlinked predictor scale;
4. compare reconstructed standard deviations or correlations only after
   applying the configured link or partial-correlation recursion; and
5. for mixture fits, condition comparisons on a resolved class-labelling rule.

Exact numerical equality is expected between simulation and fitting design
matrices and parameter order. Posterior estimates need not equal generating
values in a finite sample; that is an inferential property rather than an
interface translation discrepancy.
