---
title: "Guide to simulating latent-progress JoiNMe models"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  eval: false
  warning: false
  message: false
---

# Overview

This guide describes a simulation-to-fit workflow using
`simulate_joinme_mix()` and `joinme_mix()`. The formula, family, association
and clustering arguments retain the fitting meanings. Additional generative
values are collected in `class_parameters`, in the same way that
`beta_long`, `re_params` and `assoc_coefs` provide truth for the ordinary
simulator.

# A minimal subject-level mixture

```{r}
library(joinme)

simulation <- simulate_joinme_mix(
  formulaLong = y ~ 1 + time + x1 +
    (1 + time | id) +
    (0 + x1 + (1 + time | id) | marker),
  formulaEvent = survival::Surv(time, event) ~ x1 + x2,
  n_id = 100,
  families = c("gaussian", "student_t", "gaussian"),
  times_obs = seq(0, 5, length.out = 8),
  n_clusters = 2,
  cluster_type = "subject",
  seed = 4101
)
```

By default, the first two available subject coordinates are clustered.
The two component centres are placed at \(-1.25\) and \(1.25\), and the
component scales are 0.65.

```{r}
simulation$truth$mixture$dimensions
simulation$truth$mixture$location
simulation$truth$mixture$scale
table(simulation$truth$mixture$allocation$subject)
```

# Matching simulation and fitting syntax

The following arguments have the same meaning in both entry points:

```text
formulaLong
formulaEvent
formulaVCov
formulaDist
transforms
families
fixed_marker_weights
shared_marker_weights
n_clusters
formulaCluster
cluster
cluster_dimensions
cluster_ordering
seed
```

The simulation produces `dataLong` and `dataEvent`; the fitting function
consumes them:

```{r}
fit <- joinme_mix(
  formulaLong = simulation$truth$formulaLong,
  dataLong = simulation$dataLong,
  formulaEvent = simulation$truth$formulaEvent,
  dataEvent = simulation$dataEvent,
  formulaVCov = simulation$truth$formulaVCov,
  families = simulation$marker_info$families,
  n_clusters = 2,
  cluster_type = "subject",
  seed = 4101
)
```

# Selecting progress coordinates

`cluster_dimensions` refers to standardised coordinates at the selected level.

```{r}
simulation <- simulate_joinme_mix(
  n_id = 120,
  families = rep("gaussian", 3),
  n_clusters = 3,
  cluster_type = c("subject", "vcov"),
  cluster_dimensions = list(
    subject = c(1, 2),
    vcov = c(1, 3)
  ),
  seed = 4102
)
```

The packed location matrix has four columns in the order subject
coordinates 1 and 2, then covariance-regression coordinates 1 and 3.

```{r}
simulation$truth$mixture$starts
simulation$truth$mixture$dimensions
```

# Supplying class probabilities

```{r}
simulation <- simulate_joinme_mix(
  n_id = 200,
  families = rep("gaussian", 3),
  n_clusters = 3,
  cluster_type = "subject",
  class_parameters = list(
    probability = c(0.20, 0.55, 0.25)
  ),
  seed = 4103
)
```

These are baseline probabilities. Realised frequencies vary by Monte Carlo
sampling, and may also vary systematically when `formulaCluster` is active.

# Class-probability regression

For a shared formula and three classes, coefficients are stored against the
design labels shown below:

```{r}
pilot <- simulate_joinme_mix(
  n_id = 20,
  families = rep("gaussian", 3),
  n_clusters = 3,
  formulaCluster = ~ treatment + age,
  covariate_formulas = list(
    treatment ~ rbinom(n_id, 1, 0.5),
    age ~ rnorm(n_id, 65, 8),
    x1 ~ rnorm(n_id),
    x2 ~ rnorm(n_id)
  ),
  class_parameters = list(probability = c(0.25, 0.50, 0.25)),
  seed = 4104
)

pilot$truth$mixture$class_design$subject$columns
```

Use those labels for robust coefficient matching:

```{r}
simulation <- simulate_joinme_mix(
  n_id = 250,
  families = rep("gaussian", 3),
  n_clusters = 3,
  formulaCluster = ~ treatment + age,
  covariate_formulas = list(
    treatment ~ rbinom(n_id, 1, 0.5),
    age ~ rnorm(n_id, 65, 8),
    x1 ~ rnorm(n_id),
    x2 ~ rnorm(n_id)
  ),
  class_parameters = list(
    probability = c(0.25, 0.50, 0.25),
    coefficient = list(
      subject = c(
        "class_1:treatment" = -0.7,
        "class_1:age" = -0.02,
        "class_2:treatment" = 0.5,
        "class_2:age" = 0.01
      )
    )
  ),
  seed = 4105
)
```

Inspect the realised unit probabilities rather than assuming they equal the
baseline simplex:

```{r}
head(simulation$truth$mixture$probability_by_unit$subject)
```

# Class-specific formulae

```{r}
simulation <- simulate_joinme_mix(
  n_id = 180,
  families = rep("gaussian", 3),
  n_clusters = 3,
  formulaCluster = list(
    ~ treatment,
    ~ age,
    ~ 1
  ),
  cluster_ordering = "none",
  covariate_formulas = list(
    treatment ~ rbinom(n_id, 1, 0.5),
    age ~ rnorm(n_id, 65, 8),
    x1 ~ rnorm(n_id),
    x2 ~ rnorm(n_id)
  ),
  seed = 4106
)
```

A class-specific formula list defaults to no ordering. Supply an ordering
explicitly only when the chosen truth satisfies it.

# Locations and scales

For two selected coordinates and three classes:

```{r}
location <- rbind(
  c(-1.6, -0.8),
  c( 0.0,  0.2),
  c( 1.7,  1.0)
)

scale <- rbind(
  c(0.45, 0.55),
  c(0.70, 0.65),
  c(0.50, 0.60)
)

simulation <- simulate_joinme_mix(
  n_id = 180,
  families = rep("gaussian", 3),
  n_clusters = 3,
  cluster_type = "subject",
  cluster_dimensions = c(1, 2),
  class_parameters = list(
    probability = c(0.3, 0.4, 0.3),
    location = location,
    scale = scale
  ),
  seed = 4107
)
```

Small within-class scales and distant locations create an easy classification
problem. Recovery studies should vary both quantities.

# Combined subject and covariance clustering

```{r}
simulation <- simulate_joinme_mix(
  formulaVCov = ~ treatment + age,
  n_id = 180,
  families = rep("gaussian", 4),
  n_clusters = 3,
  cluster_type = c("subject", "vcov"),
  cluster_dimensions = list(
    subject = c(1, 2),
    vcov = c(1, 2)
  ),
  class_parameters = list(
    location = list(
      subject = rbind(
        c(-1.4, -0.7),
        c( 0.0,  0.0),
        c( 1.4,  0.7)
      ),
      vcov = rbind(
        c(-0.8, -0.3),
        c( 0.0,  0.0),
        c( 0.8,  0.3)
      )
    ),
    scale = 0.6
  ),
  covariate_formulas = list(
    treatment ~ rbinom(n_id, 1, 0.5),
    age ~ rnorm(n_id, 65, 8),
    x1 ~ rnorm(n_id),
    x2 ~ rnorm(n_id)
  ),
  seed = 4108
)
```

There is one subject allocation vector:

```{r}
simulation$truth$mixture$allocation_domains$subject
simulation$truth$mixture$allocation$subject
```

# Marker clustering with ordinary marker weights

```{r}
simulation <- simulate_joinme_mix(
  n_id = 100,
  families = rep("gaussian", 6),
  assoc = c("cv_total", "cs_marker"),
  shared_marker_weights = FALSE,
  fixed_marker_weights = FALSE,
  n_clusters = 2,
  formulaCluster = ~ marker,
  cluster_type = "marker",
  class_parameters = list(
    probability = c(0.5, 0.5),
    location = list(
      marker = matrix(c(-1.2, 1.2), nrow = 2)
    ),
    scale = 0.5
  ),
  seed = 4109
)
```

The unshared weighted associations remain part of the survival model, whilst
the latent classes act only on the marker random-effect block.

# Alternative component families

```{r}
student_t_simulation <- simulate_joinme_mix(
  shrinkage = 0,
  n_clusters = 2,
  seed = 4110
)

laplace_simulation <- simulate_joinme_mix(
  shrinkage = 1,
  n_clusters = 2,
  seed = 4111
)

normal_simulation <- simulate_joinme_mix(
  shrinkage = 2,
  n_clusters = 2,
  seed = 4112
)
```

The fitting call must use the same `shrinkage` value through `...`.

# Longitudinal-only simulation

Set `formulaEvent = NULL` in both simulation and fitting:

```{r}
simulation <- simulate_joinme_mix(
  formulaEvent = NULL,
  n_id = 120,
  families = rep("gaussian", 3),
  n_clusters = 2,
  cluster_type = "subject",
  seed = 4113
)

fit <- joinme_mix(
  formulaLong = simulation$truth$formulaLong,
  dataLong = simulation$dataLong,
  n_clusters = 2,
  cluster_type = "subject"
)
```

`simulation$dataEvent` and `simulation$truth$formulaEvent` are `NULL`.
Association arguments are invalid because no event process exists.

# Ordering

Probability ordering requires increasing baseline probabilities:

```{r}
simulation <- simulate_joinme_mix(
  n_clusters = 3,
  cluster_ordering = "probability",
  class_parameters = list(
    probability = c(0.15, 0.30, 0.55)
  ),
  seed = 4114
)
```

Intercept ordering checks only the packed selected random-intercept location.
Slope locations need not be ordered.

# Recommended simulation study design

Vary at least:

- sample size and marker count;
- `n_clusters`;
- minimum class prevalence;
- location separation;
- component scales;
- class-regression strength;
- shrinkage family;
- association magnitude;
- observation frequency;
- response-family mixture;
- censoring and delayed entry; and
- fitted class-count misspecification.

For every replicate, store convergence diagnostics, component parameter bias,
interval coverage, posterior allocation probabilities, entropy, proper class
scores, predictive performance and event-model recovery.

# Common mistakes

1. Do not append true labels to the fitted data.
2. Do not compare unordered labels without permutation alignment.
3. Do not interpret `class_parameters$probability` as realised frequency.
4. Do not treat covariance-regression coordinates as realised covariance
   entries; they are standardised latent predictors entering those entries.
5. Do not use only extremely separated components when evaluating practical
   performance.
