---
title: "JoiNMe: Association Construction"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
options(width = 80)
```

# Purpose

This vignette focuses on **association term construction**, including transformation selection, interpretability, and practical trade-offs.

It also explains how those association terms propagate into the public summary,
draw-extraction, and plotting APIs so the same labels are used consistently in
tables, MCMC plots, and fitted association curves.

Association transforms apply to all supported longitudinal families (including
double_exponential, skew_double_exponential, beta, and cumulative_logit) and to
competing-risk survival models.

## Step-by-step: how association summaries are constructed

At the event time and at the default 15 Gauss-Kronrod nodes used for integration
(configurable), the
longitudinal linear predictor is decomposed into a mean component and a marker
component. From these components, `JoiNMe` constructs:

- **CV_total**: total current value (mean + marker components).
- **CV_mean**: current value of the mean component only.
- **CV_marker**: current value of the marker component only.
- **CS_total**: total current slope, approximated by a forward finite difference
  and rescaled to the original time units.
- **CS_mean**: slope of the mean component.
- **CS_marker**: slope of the marker component.
- **CORR**: a subject-specific variance summary of the marker-by-id covariance
  structure, constant across time.
- **VCOV**: a subject-specific lower-triangular summary of the marker-by-id
  Cholesky factor `L`, also constant across time. The channel uses the entries
  of `L` directly rather than the reconstructed covariance matrix `L L'`.

Each component can be included or excluded using the `assoc` argument. If
`assoc` does not include a component, the corresponding coefficient is prior-only
and does not enter the likelihood.

### Scale recovery and exponentiation in summaries

`JoiNMe` reports association and survival-process coefficients on the linear
predictor scale first. Any internal time normalisation used by the Stan model
is recovered on that linear scale before exponentiation.

For multiplicative interpretations:

- `summary(fit)$tables$baseline_hazard` reports baseline-hazard coefficients on
  the log scale and hazard-ratio scale. Only the intercept is adjusted by
  `-log(tmax)`; non-intercept basis coefficients are unchanged.
- `summary(fit)$tables$survival_process` reports baseline covariate coefficients
  and `Hazard.Ratio` interval columns from exponentiated log-scale summaries.
- `summary(fit)$tables$assoc` reports association coefficients on their native
  linear scale (including marker-weight terms when active).

The intercept-only baseline adjustment follows the proportional-hazards identity
that internal time scaling contributes a constant offset on the log-hazard
scale. Therefore only the intercept is shifted; spline/formula basis terms are
kept as fitted.

The CORR and VCOV components require marker-by-id random effects. If the marker
block does not include an inner `( ... | id )` term, then `Q_idm = 0` and these
covariance-style associations are not available.

# Marker-weighted CV and CS

When multiple biomarkers are modelled, `JoiNMe` constructs marker-average association
components using weights $\omega_d$ (one per marker). For marker-aggregated
terms (`cv_total`, `cv_marker`, `cs_total`, `cs_marker`), transforms are applied
at the marker level first and then averaged with weights. This matters whenever
transforms are nonlinear.

For example, with marker-specific total current values $CV_{id}(t)$ and transform
$f_{cv}$:

$$
  	ext{cv\_total}(t) = \frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cv}\!\left(CV_{id}(t)\right),
$$

which differs from $f_{cv}(\frac{1}{D}\sum_d \omega_d CV_{id}(t))$ unless $f_{cv}$
is linear.

For slopes, let $CS_{id}(t)$ be a marker-specific slope feature and $f_{cs}$ the
configured CS transform. Then

$$
  	ext{cs\_total}(t) = \frac{1}{D}\sum_{d=1}^{D} \omega_d\, f_{cs}\!\left(CS_{id}(t)\right),
$$

and analogously for `cs_marker`. This is intentionally not equal to
$f_{cs}(\frac{1}{D}\sum_d \omega_d CS_{id}(t))$ unless $f_{cs}$ is linear.

In practice, provide fixed weights with `marker_weights` and
`fixed_marker_weights = TRUE`, or set `fixed_marker_weights = FALSE` to use
signed perturbations around base weights,
$\omega_d = \omega_d^{(0)} + z_d$. The `shrinkage` switch controls the
standardized latent family in both fitting and simulation: `0` is
Student-$t_6(0,1)$, `1` is Laplace$(0,1)$, and `2` is Normal$(0,1)$. With
estimated weights and no supplied base, $\omega^{(0)}=0$. The simulator stores
the base, latent, and effective truths in `truth$marker_weights_base`,
`truth$marker_weights_latent`, and `truth$marker_weights`; the last of these is
the effective weight used to generate survival times.

## Plotting fitted association curves

```{r association-plot-helper, eval=FALSE}
association_plot(fit)
```

## Posterior association extraction and covariance-style labels

Use `posterior_assoc(fit, summary = TRUE)` (or `assoc(fit, summary = TRUE)`) to
extract association-specific posterior summaries without requiring the full model
summary. Use `summary = FALSE` to obtain posterior draws per association term.

## Renamed posterior draws for association terms

Association terms can be extracted directly on the reporting scale with
`draws()`. The canonical posterior variables are `alpha_cv_total`,
`alpha_cs_total`, `alpha_cv_mean`, `alpha_cs_mean`, `alpha_corr[...]`, and
`alpha_vcov[...]`. Each is the coefficient actually used in the event hazard,
which is also the scale reported by `summary()` and used by prediction and
plotting methods.

```{r association-renamed-draws, eval=FALSE}
# All association terms with reporting-friendly names
assoc_draws <- draws(fit, variables = "^alpha_", regex = TRUE, format = "draws_df")
head(assoc_draws)

# Bayesplot visualisation with the same labels used by summary(fit)
mcmc_plot(fit, variable = "alpha_cv_total", type = "dens_overlay")
```

## Conditional effects in the two joint-model processes

Conditional effects are distinct from subject-specific dynamic prediction. `conditional_effects()` varies named predictors while holding other predictors at explicit profile values and can return the longitudinal process, the event
process, or both. The longitudinal estimand is the population-level expected marker response by default. Set `longitudinal_estimand = "marker"` to retain the fitted marker-level deviation, or `"marginal_marker"` to average the marker-specific predictions within each posterior draw. Because marker averaging occurs after each inverse-link transformation, it targets an average expected response rather than an expected response at an average marker effect.

The event estimand is the direct relative-hazard component `exp(W * gamma)`; the baseline hazard and longitudinal association contribution are held fixed when making event-covariate comparisons.

```{r association-conditional-effects, eval=FALSE}
profiles <- make_conditions(fit$dataEvent, vars = "treatment")

ce <- conditional_effects(
  fit,
  effects = list(
    longitudinal = "time:treatment",
    event = "treatment"
  ),
  conditions = profiles,
  process = c("longitudinal", "event"),
  longitudinal_estimand = "marker",
  plot = FALSE
)

# The nested tables use the same estimate__, se__, lower__, upper__, and cond__
# columns as brms conditional effects.
head(ce$longitudinal[[1]])
head(ce$event[[1]])
plot(ce, ask = FALSE)
```

# Association coefficients and naming

Internally, association coefficients are denoted with the $\alpha$ prefix (e.g., `alpha_cv_total`). 
This follows joint-model conventions and avoids confusion with the linear predictor
$\eta$ used throughout the longitudinal and survival submodels.
Total, mean, and marker associations use a positive non-centred parameterisation
$\alpha = z_{\alpha} \cdot sd_{\alpha}$ for stable sampling.

When an association component is not selected in `assoc`, its coefficient is prior-only and does not enter the likelihood. Summaries only report active components to keep interpretation focused on the fitted association structure.

# Transform modes

`JoiNMe` supports four transformation modes for association features:

- **Identity** (`type = "identity"`): no transformation.
- **Functional** (`type = "functional"`): a user-defined expression converted
  into a fixed bytecode program.
- **Monotone I-spline** (`type = "ispline"` or `"ispline_penalised"`): a smooth
  monotone mapping.
- **Bounded-domain monotone I-spline** (`type = "ispline_expit"` or
  `"ispline_expit_penalised"`): the same monotone spline machinery, but applied
  to `plogis(x)` rather than directly to `x`.
- **Piecewise linear** (`type = "pwlin"`): linear interpolation across knots.

Interpretation note for recovery experiments:

- If you use a fixed spline (`type = "ispline"` or an expit variant with
  explicit `coeff`), the association coefficient is interpretable on that fixed
  transformed scale and can be checked directly in simulate-fit recovery tests.
- If you use a jointly estimated penalised spline (`*_penalised`),
  the transform shape and the association coefficient are learned together. The
  fitted alpha is still meaningful inside the model, but coefficient recovery is
  typically weaker than in the fixed-transform case because the data must learn
  both the transformed scale and the hazard weight simultaneously.

## Default values and minimal parameterisations

- **Identity**:
  - Syntax: `list(type = "identity")`
  - Default when a transform term is omitted or set to `NULL`.
  - No other parameters are used.
- **Functional**:
  - Syntax: `list(type = "functional", expr = ~ log1p(x))`
  - Required field: `expr`.
  - No other defaults beyond the supplied expression.
- **I-spline**:
  - Syntax: `list(type = "ispline", knots = c(-1, 0, 1), coeff = c(0, 0.3, 0.8, 1.1, 1.3))`
  - Required fields: `knots`, `coeff`.
  - Default: `degree = 3` if omitted.
- **Penalised I-spline**:
  - Syntax:
    `list(type = "ispline_penalised", x = seq(-2, 2, length.out = 50), n_knots = 6, degree = 3, lambda = 1)`
  - Defaults: `n_knots = 6`, `degree = 3`, `lambda = 1`.
  - If `knots` are omitted in Stan-estimated mode, they are derived from `x` quantiles.
- **Expit-based penalised I-spline**:
  - Syntax:
    `list(type = "ispline_expit_penalised", x = seq(0.02, 0.98, length.out = 50), n_knots = 6, degree = 3, lambda = 1)`
  - Defaults: `n_knots = 6`, `degree = 3`, `lambda = 1`.
  - The raw association feature is mapped through `plogis(x)` at runtime, and
    the spline basis is built on that bounded domain. Supply training `x` and
    explicit `knots` directly on the expit scale in `[0, 1]` so the fitted
    spline domain matches the runtime basis.
- **Piecewise linear**:
  - Syntax: `list(type = "pwlin", x = c(-2, -1, 0, 1, 2), y = c(0.2, 0.5, 1, 0.5, 0.2))`
  - Required fields: `x`, `y`.
  - No additional defaults.

# Functional Transform

```{r functional-transform, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "functional",
    expr = ~ sinh(x) + tanh(x)
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```

The functional evaluator supports common arithmetic and functions such as
`log`, `exp`, `sqrt`, `sin`, `cos`, `tanh`, plus aliases `sigmoid`/`expit`,
`softplus`/`log1p_exp`, and helpers like `cbrt` and `power`.

```{r functional-transform-mixed-demo, eval=FALSE}
transforms <- list(
  cv_total = list(type = "functional", expr = ~ inv_logit(x)),
  cv_mean = list(type = "functional", expr = ~ exp(0.2 * x)),
  cv_marker = list(type = "functional", expr = ~ power(x, 2)),
  cs_total = list(type = "functional", expr = ~ sqrt(abs(x) + 1e-6)),
  corr = list(type = "functional", expr = ~ cbrt(x))
)
```

# Monotone I-Spline

For `type = "ispline"`, we specify the transform shape following `splines2::iSpline()` convention:

- `knots`: knot locations (first and last are boundary knots),
- `coeff`: non-negative I-spline basis coefficients,
- `degree`: spline degree (usually `3`).

Defaults:

- `degree = 3` if omitted.
- `knots` and `coeff` must be supplied explicitly.

For `type = "ispline_expit"`, the same direct specification applies, except the I-spline basis is evaluated on `plogis(x)`. The runtime association feature is still shown on its raw scale, but the transform's `knots` must be supplied on the expit scale because that is the spline domain.

This is highly expremental and meant for stabilising the boundary knots (with is usually unknown). You may want to standardise the longitudinal outcome so that most association would not squashed to 0 or 1, as `plogis(x)` is very flat outside of the range `[-5, 5]`. You may also want to put more knots as the extreme values of `plogis(x)` to overcome the flatness of the transform. 

```{r monotone-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.5, 1, 1.2, 1.5, 1.6),
    degree = 3
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "identity"
  )
)
```


# Penalised Monotone I-Spline

For `type = "ispline_penalised"`, there are now two supported modes:

- **Simulation mode**: provide `x` and `y`, and the package first learns a monotone spline in R before passing fixed coefficients into Stan. This is meant for the simulation but can be used in model fitting as a mean to provide some prior the association curve.
- **Modelling mode**: omit `y`, keep `knots` (or `n_knots`) plus `lambda`, and Stan estimates the monotone spline coefficients jointly with the rest of the model.

Meaning of the arguments:

- `x`: raw association-feature values, used either as training inputs or to help choose knots.
- `y`: optional target transformed values at those `x` points. If present, you are telling the package the shape you want to approximate.
- `lambda`: smoothness penalty weight (higher = smoother, lower = more flexible).

Defaults:

- `n_knots = 6` if `knots` are omitted.
- `degree = 3` if omitted.
- `lambda = 1.0` if omitted.
- In helper-based plug-in fitting, `weights = NULL` means equal weights and
  `diff_order = 2` is the default penalty order.

```{r penalised-ispline, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "ispline_penalised",
    x = seq(-2, 2, length.out = 40),
    y = exp(seq(-2, 2, length.out = 40)),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

```{r penalised-ispline-stan, eval=FALSE}
transforms <- list(
  corr = list(
    type = "ispline_penalised",
    x = seq(-2, 2, length.out = 40),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)
```

**When to use**

- You want a monotone transform with smoothing, either anchored by `(x, y)` orestimated jointly in Stan.
- Penalisation avoids overfitting when knot counts are moderate.

# Piecewise-Linear

This is mean for simulation but you can use for fiting as well as a mean to provide some prior the association curve. The transform is defined by a set of knots and corresponding values. The transform is linear between knots and constant outside the knot range. 

```{r  piecewise-linear, eval=FALSE}
transforms <- list(
  cv_total = list(
    type = "identity"
  ),
  cs_total = list(
    type = "identity"
  ),
  corr = list(
    type = "pwlin",
    x = c(-2, -1, 0, 1, 2),
    y = c(0.2, 0.5, 1, 0.5, 0.2)
  )
)
```


# Association Diagram

```{mermaid}
%%| fig-cap: "Association components mapped into the survival predictor."
graph TD
  CV[CV total] --> FCV[f_cv]
  CS[CS total] --> FCS[f_cs]
  V[Corr] --> FCORR[f_corr]
  FCV --> ETA[eta assoc]
  FCS --> ETA
  FCORR --> ETA
```

# Hierarchical Covariance Regression

```{mermaid}
%%| fig-cap: "Covariance regression structure for marker-by-id effects."
graph TD
  X[Covariates] --> LP[Linear predictor]
  UL[Latent effect] --> LP
  LP --> L[Cholesky factor]
  L --> W[Marker-by-id effects]
  Z[Latent z] --> W
```

# Fit with Custom Association

```{r fit-custom-association, eval=FALSE}
fit <- joinme(
  formulaLong = y ~ 1 + time + x1 +
    (1 + time || id) +
    (0 + x1 + (1 + time || id) || marker),
  dataLong = sim$dataLong,
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  dataEvent = sim$dataEvent,
  transforms = transforms,
  assoc = c("cv_total", "corr", "vcov")
)
```
