---
title: "Using latent-progress mixtures in JoiNMe"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(joinme)
library(survival)
```

# Purpose

This guide is a practical companion to the statistical theory vignette. It
shows how to choose a class type, how combined types share classes, how
to fit longitudinal-only and joint models, and how to use the inherited
summary, prediction and diagnostic methods.

# Minimal joint model

Start with the same nested longitudinal formula that would be supplied to
`joinme()`:

```{r}
longitudinal_formula <-
  y ~ 1 + time + treatment +
  (1 + time | id) +
  (0 + treatment + (1 + time | id) | marker)

event_formula <-
  Surv(time, event) ~ treatment + age
```

A two-class subject latent-progress model is:

```{r}
fit_subject <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  formulaEvent = event_formula,
  dataEvent = data_event,
  n_classes = 2,
  class_type = "subject",
  assoc = c("cv_mean", "cv_total"),
  seed = 2718
)
```

The class distribution replaces the ordinary distribution of the first two
standardised subject random-effect coordinates. If these are a random
intercept and time slope, the progress plane has the familiar interpretation
of class-specific level and rate of change. Confirm the actual coordinate
labels in `summary(fit_subject)`.

# Choosing a class type

## Subject

Use `"subject"` when the scientific question concerns shared progression
across all markers:

```{r}
fit_subject <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 3,
  class_type = "subject",
  assoc = "cv_mean"
)
```

The posterior contains one class-probability vector per subject.

## Marker

Use `"marker"` when markers themselves may form classes with different common
longitudinal deviations:

```{r}
fit_marker <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 2,
  class_type = "marker",
  assoc = "cv_total"
)
```

This changes only the longitudinal marker-effect distribution. It does not
create a new association coefficient by class. Any survival difference arises
through the class-specific longitudinal trajectory entering an already fitted
association.

## Marker-by-subject covariance regression

Use `"vcov"` when subjects may differ in the covariance structure of
their marker-specific departures:

```{r}
fit_covariance <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  formulaVCov = ~ treatment + age,
  n_classes = 3,
  class_type = "vcov",
  assoc = "vcov"
)
```

The nested marker term must contain an inner subject random-effect term.
The selected coordinates refer to the covariance-regression latent vector
`z_L`, not to `w_idm` innovations.

Each covariance coordinate \(m\) has one fitted non-negative scalar loading
`lambda_L[m]`. All loadings together form a vector of length
\(M_{\mathrm{cov}}\), not a dense matrix. `lambda_L[m]` multiplies only
`z_L[i,m]` and measures residual subject heterogeneity on that covariance
predictor before applying the diagonal link or off-diagonal `tanh`
transformation. In class \(g\), the corresponding shift and scale are
`lambda_L[m] * mix_location[g,m]` and
`lambda_L[m] * mix_scale[g,m]`, respectively.

## Estimated marker weights

Marker weights retain their ordinary JoiNMe meaning and may still be shared or
association-specific. They are not a latent-progress class type.

# Combining types without Cartesian expansion

Compatible subject-indexed blocks can be combined:

```{r}
fit_subject_process <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  formulaVCov = ~ treatment,
  n_classes = 3,
  class_type = c("subject", "vcov"),
  assoc = c("cv_mean", "vcov")
)
```

For subject \(i\), the subject effects and covariance-regression effects
are evaluated under the same class \(g\) inside one `log_sum_exp`. There are
three joint classes, not nine.

Marker effects may be combined with either subject or covariance classes:

```{r}
fit_marker_process <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 3,
  class_type = c("subject", "marker"),
  assoc = c("cv_total", "cs_total"),
  fixed_marker_weights = FALSE
)
```

It is also possible to select blocks from both natural domains:

```{r}
fit_both_domains <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 2,
  class_type = c("subject", "marker")
)
```

The two domains share the same two component labels and mixing proportions.
Class probabilities remain separate because subjects and markers are
different observational units.

# Selecting the two-dimensional progress plane

The default is the first two available coordinates per selected block. Make the
choice explicit when formula order is not the intended scientific order:

```{r}
fit_selected <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 3,
  class_type = c("subject", "vcov"),
  class_dimensions = list(
    subject = c(1, 2),
    vcov = c(1, 3)
  )
)
```

For one selected type, an integer vector is sufficient:

```{r}
fit_selected_marker <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 2,
  class_type = "marker",
  class_dimensions = c(1, 2)
)
```

Marker weights are the exception to the two-coordinate default. Every active
weight set is included unless `class_dimensions` explicitly selects a
subset. This protects the meaning of unshared association weights.

# Longitudinal-only latent progress

Omit both event arguments:

```{r}
fit_longitudinal <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = data_long,
  n_classes = 3,
  class_type = c("subject", "vcov"),
  formulaVCov = ~ treatment
)
```

No survival likelihood is introduced and `assoc` should not be supplied. The
following remain available:

```{r}
summary(fit_longitudinal)
posterior_class(fit_longitudinal)
pp_check(fit_longitudinal)
mcmc_plot(fit_longitudinal, type = "trace")
longitudinal_plot(
  fit_longitudinal,
  estimand = "mean_per_class"
)
```

Event-specific requests, such as `concordance(fit_longitudinal)`, return a clear
message because no event process was fitted.

# Priors and component family

The main controls are:

```{r}
fit_regularised <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 3,
  class_type = "subject",
  formulaClass = ~ treatment + age_at_entry,
  priors = jm_prior(
    class_probability = c(2, 2, 2),
    class_regression = 1
  ),
  shrinkage = 0
)
```

- `class_probability` is the Dirichlet concentration for zero-covariate class
  probabilities.
- `class_regression` is the Normal prior scale for coefficients introduced by
  `formulaClass`.
- `class_ordering = "intercept"` orders only the selected random-intercept
  location; slope locations remain unrestricted.
- `class_ordering = "probability"` orders baseline class probabilities,
  whilst `"none"` imposes no ordering.
- `shrinkage` selects Student-\(t_6\), Laplace, or Normal components.

Ordering identifies the labels; it does not guarantee that the data support
well-separated scientific classes. Fit sensitivity models and compare posterior
allocations and predictive adequacy.

Class-specific membership predictors are written as one formula per class:

```{r}
fit_class_specific <- joinme_mix(
  longitudinal_formula,
  data_long,
  event_formula,
  data_event,
  n_classes = 3,
  class_type = "subject",
  formulaClass = list(
    ~ treatment,
    ~ age_at_entry + biomarker_at_entry,
    ~ treatment + age_at_entry
  )
)
```

This form defaults to `class_ordering = "none"`. Set
`class_ordering = "intercept"` or `"probability"` explicitly when the
scientific interpretation supports that order.

# Summaries and class membership

```{r}
fit_summary <- summary(fit_subject_process)
fit_summary

class_probability <- posterior_class(
  fit_subject_process,
  draws = 1000,
  seed = 41
)
class_probability
plot(class_probability, domain = "subject")
```

The summary adds:

- zero-covariate class probabilities;
- separate component-location and within-class-scale tables, with original
  block labels;
- class-membership regressions, with the reference class stated explicitly;
- a compact posterior allocation-count distribution by active domain; and
- the ordinary JoiNMe coefficient, covariance and diagnostic sections.

The component locations and scales are on the standardised latent
random-effect scale. They describe the distribution used to generate the
selected coordinates; they are not response-scale trajectories or covariance
curves. Each table presents `Estimate`, `Est.Error`, `Q2.5`, `Q97.5`, `Rhat`,
`ess_bulk` and `ess_tail`. Here `Estimate` is the posterior mean and
`Est.Error` is the posterior standard deviation. For a common
`formulaClass`, class `n_classes` is the zero-coefficient reference and is
therefore intentionally absent from the regression-coefficient table.

The allocation-count table reports the posterior mean, posterior standard
deviation and interval of the draw-specific soft class count, plus the number
of units assigned by maximum posterior probability. The console summary does
not repeat every subject-by-class or marker-by-class probability.
`posterior_class()` returns that complete table, using the same compact
posterior-summary columns, when individual
allocation uncertainty is required. Its print method first shows compact
maximum-probability allocation counts and then a bounded preview of the full
unit-by-class table; the returned object is never truncated. `plot()` displays
the same means and intervals.

Raw probability draws are available with:

```{r}
class_probability_draws <- posterior_class(
  fit_subject_process,
  summary = FALSE
)
```

Use these draws when a downstream analysis needs to propagate classification
uncertainty.

# Longitudinal plots

At component centres:

```{r}
longitudinal_plot(
  fit_subject_process,
  estimand = "mean_per_class",
  longitudinal_points = 100
)
```

Marginal over the fitted within-class distribution before applying the inverse
link:

```{r}
longitudinal_plot(
  fit_subject_process,
  estimand = "marginal_per_class",
  draws = 800,
  seed = 92
)
```

For a model with classes defined only through covariance, centre trajectories may coincide.
This is informative: the class distinction lies in dispersion or covariance
rather than the longitudinal location.

# Covariance curves

```{r}
plot(
  fit_covariance,
  type = "covariance_class",
  ci_levels = c(0.5, 0.95)
)
```

The method returns a separate plot for every marker. Every class curve is drawn
in the same panel for a covariance component. This permits direct comparison
without creating one scale per class.

# Association trajectories

```{r}
association_plot(
  fit_subject_process,
  estimand = "mean_per_class",
  association_term = "cv_mean"
)
```

For a weighted total association:

```{r}
association_plot(
  fit_marker_process,
  estimand = "mean_per_class",
  association_term = "cv_total"
)
```

The transformation and marker weighting follow the fitted hazard order. The
same plot is available from posterior association output:

```{r}
association_posterior <- posterior_assoc(fit_marker_process)
association_posterior
plot(
  association_posterior,
  estimand = "mean_per_class",
  association_term = "cv_total"
)
```

The first part of `association_posterior` reports the global association
coefficient or marker-weighted effect. If the fitted class type changes the
longitudinal or covariance feature entering that association, a second part
reports the corresponding log-hazard contribution by class at compact
reference points. No class rows are manufactured when the fitted class block
is irrelevant to the association term.

Or request it directly:

```{r}
posterior_assoc(
  fit_marker_process,
  trajectory = TRUE,
  estimand = "marginal_per_class",
  association_term = "cv_total"
)
```

For `marginal_per_class`, every within-class random-effect realisation is
passed through the fitted association transform before averaging. This matters
for nonlinear transforms: transforming an average trajectory would answer a
different question.

# Prediction and diagnostics

Joint-model prediction retains the existing interface:

```{r}
dynamic_prediction <- predict(
  fit_subject_process,
  newdataLong = new_longitudinal_history,
  newdataEvent = new_event_data,
  process = c("longitudinal", "event")
)

summary(dynamic_prediction)
plot(dynamic_prediction)
```

The returned object inherits from `JoiNMeDynPred` and additionally retains the
parent mixture specification. For class-specific subject-level blocks, the
conditioning calculation also replaces the ordinary random-effect prior by the
component probability, location and scale from the corresponding fitted-model
draw. Thus the new subject's longitudinal history updates both their random
effects and their uncertain class allocation.

```{r}
conditional_class_probability <- posterior_class(dynamic_prediction)
conditional_class_probability

plot(
  dynamic_prediction,
  type = "class_membership"
)
```

When `subject` and `vcov` are combined, this output contains one probability
vector per new subject, not one vector per block. The probability vector should be reported in
preference to a hard label whenever classes overlap.

Established fit methods are inherited:

```{r}
pp_check(fit_subject_process)
diagnosis(fit_subject_process)
mcmc_plot(
  fit_subject_process,
  variable = "mix_location",
  type = "trace"
)
concordance(fit_subject_process)
tv_roc <- tvROC(fit_subject_process, Tstart = 1, Dt = 2)
tvAUC(tv_roc)
```

For a mixture fit, `diagnosis()` adds three deliberately distinct pieces of
evidence. `energy_by_chain` reports E-BFMI, energy persistence, tree depth,
leapfrog effort, acceptance and step size for each chain. `scale_tradeoff`
reports the posterior correlation between an ordinary random-effect scale and
the within-class component scale applied before it. `energy_association`
ranks low-dimensional posterior block summaries by their within-chain
correlation with energy. These correlations are diagnostic clues, not causal
tests. A strong negative scale correlation explains a compensating posterior
ridge; it does not establish an indexing error.

For mixture-specific assessment, include at least:

```{r}
plot(fit_subject_process, type = "class_membership")
posterior_class(fit_subject_process)
summary(fit_subject_process)
```

# Practical fitting sequence

A stable workflow is:

1. Fit and check the corresponding ordinary `joinme()` model.
2. Choose a scientifically interpretable two-coordinate plane.
3. Fit \(G=2\) with mildly regularising priors.
4. Inspect mixture trace plots, \(\widehat R\), effective sample sizes and
   divergent transitions.
5. Inspect posterior membership probabilities.
6. Compare centre and marginal class trajectories.
7. Fit plausible neighbouring values of \(G\).
8. Repeat with alternative credible class priors and component family.
9. Report uncertainty in proportions, allocations, locations and curves.

# Frequent misunderstandings

`n_classes = 3` always means three component labels. Combining two levels does not
mean \(3\times3\) classes.

Ordering identifies labels but is not a scientifically meaningful
minimum-distance constraint. A list of `n_classes` class-specific formulae
defaults to `class_ordering = "none"`.

A modal allocation is not observed truth. It is a posterior summary.

A covariance class can have the same conditional mean as another class.

A marker class is a class of markers, not a class of subjects.
