---
title: "A simple Normal-mixture simulation and recovery example"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
execute:
  eval: false
  warning: false
  message: false
  engine: knitr
---


# Purpose

This example is deliberately small and conservative. It checks whether
`simulate_joinme_mix()` and `joinme_mix()` agree for a two-class mixture of
Normal subject-intercept effects. It is intended for implementation debugging,
not as a stress test of heavy-tailed mixtures, covariance classes or elaborate
class-membership regressions.

The example therefore uses:

- two latent classes;
- one class-specific subject-intercept coordinate;
- an intercept-only class-membership model;
- Normal mixture innovations (`shrinkage = 2L`);
- Gaussian longitudinal markers; and
- no survival process.

Student-\(t_6\), Laplace, multivariate class blocks and covariate-dependent class
membership are documented in the latent-progress guide. They should be added
only after this elementary correspondence check succeeds.

# Model specification

```{r}
library(joinme)

set.seed(5201)

longitudinal_formula <-
  y ~ 1 + time + treatment +
  (1 + time | id) +
  (0 + treatment + (1 + time | id) | marker)

class_formula <- ~ 1
```

Only the first standardised subject random-effect coordinate is assigned a
latent class. The ordered class locations are well separated, while both
classes have the same generating within-class scale:

```{r}
subject_location <- matrix(
  c(-1.40, 1.40),
  nrow = 2,
  ncol = 1,
  dimnames = list(c("slower", "faster"), "subject_intercept")
)

subject_scale <- matrix(
  c(0.60, 0.60),
  nrow = 2,
  ncol = 1,
  dimnames = dimnames(subject_location)
)
```

These locations and scales apply to the standardised latent coordinate before
the ordinary subject random-effect Cholesky factor is applied. They are not
response-scale longitudinal intercepts.

# Simulate from the Normal mixture

```{r}
simulation <- simulate_joinme_mix(
  formulaLong = longitudinal_formula,
  formulaEvent = NULL,
  formulaVCov = ~ 1,
  n_id = 60,
  families = rep("gaussian", 3),
  marker_levels = c("inflammation", "renal", "functional"),
  times_obs = seq(0, 6, length.out = 8),
  covariate_formulas = list(
    treatment ~ rbinom(n_id, 1, 0.5)
  ),
  assoc = character(0),
  shrinkage = 2L,
  beta_long = c(
    "(Intercept)" = 1.0,
    "time" = 0.20,
    "treatment" = -0.15
  ),
  n_classes = 2L,
  formulaClass = class_formula,
  class_type = "subject",
  class_dimensions = list(subject = 1L),
  class_ordering = "intercept",
  class_parameters = list(
    probability = c(0.50, 0.50),
    location = list(subject = subject_location),
    scale = list(subject = subject_scale)
  ),
  seed = 5201
)
```

Here `shrinkage = 2L` is the important Normal-mixture choice. The three
Gaussian `families` separately specify the observation distributions of the
longitudinal markers. Because no event formula is supplied, both simulation
and fitting use the longitudinal-only likelihood.

# Inspect the generating truth

```{r}
mixture_truth <- simulation$truth$mixture

mixture_truth$distribution
mixture_truth$probability
mixture_truth$dimensions
mixture_truth$location
mixture_truth$scale
table(mixture_truth$allocation$subject)
```

The realised class labels are retained only under `truth$mixture`; they are not
added to the fitting data.

# Verify simulation-to-fit correspondence

Before sampling, prepare the fitting data and compare the exact mixture fields
recorded by the simulator:

```{r}
prepared_fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = simulation$dataLong,
  formulaEvent = NULL,
  formulaVCov = ~ 1,
  families = simulation$marker_info$families,
  transforms = simulation$truth$transforms,
  n_classes = 2L,
  formulaClass = class_formula,
  class_type = "subject",
  class_dimensions = list(subject = 1L),
  class_ordering = "intercept",
  priors = jm_prior(class_probability = c(2, 2)),
  shrinkage = 2L,
  fit = FALSE,
  seed = 5201
)

simulation_contract <- mixture_truth$stan_data_contract
contract_agreement <- vapply(
  names(simulation_contract),
  function(field_name) {
    isTRUE(all.equal(
      prepared_fit$stan_data[[field_name]],
      simulation_contract[[field_name]],
      check.attributes = FALSE
    ))
  },
  logical(1)
)

data.frame(
  Stan_field = names(contract_agreement),
  agrees = unname(contract_agreement)
)
stopifnot(all(contract_agreement))
```

This audit checks the class count, selected coordinate, packed component
position, ordering rule, class-design segment and Normal family code. It does
not compare inferential priors with fixed simulation parameters.

# Fit without using the true classes

```{r}
fit <- joinme_mix(
  formulaLong = longitudinal_formula,
  dataLong = simulation$dataLong,
  formulaEvent = NULL,
  formulaVCov = ~ 1,
  families = simulation$marker_info$families,
  transforms = simulation$truth$transforms,
  n_classes = 2L,
  formulaClass = class_formula,
  class_type = "subject",
  class_dimensions = list(subject = 1L),
  class_ordering = "intercept",
  priors = jm_prior(class_probability = c(2, 2)),
  shrinkage = 2L,
  seed = 5201,
  control = list(
    chains = 4,
    parallel_chains = 4,
    threads_per_chain = 1,
    iter_warmup = 1000,
    iter_sampling = 500,
    max_treedepth = 12,
    adapt_delta = 0.95,
    init = 0.5
  )
)
```

Ordering the first class location identifies the labels. It does not fix or
anchor either scale parameter; `mix_scale` retains its

\[
  \operatorname{Exponential}(1)
\]

prior.

# Summaries and computational checks

```{r}
fit_summary <- summary(fit)
fit_summary

fit_diagnosis <- diagnosis(fit)
fit_diagnosis
fit_diagnosis$energy_by_chain
fit_diagnosis$scale_tradeoff
head(fit_diagnosis$energy_association)

mcmc_plot(fit, variable = "mix_probability", type = "trace")
mcmc_plot(fit, variable = "mix_location", type = "trace")
mcmc_plot(fit, variable = "mix_scale", type = "trace")
```

Posterior tables use one common numerical schema:

`Estimate`, `Est.Error`, `Q2.5`, `Q97.5`, `Rhat`, `ess_bulk`, and `ess_tail`.

`Estimate` is the posterior mean and `Est.Error` is the posterior standard
deviation. Median and duplicate `Mean` or `SD` aliases are intentionally not
printed.

Even this simple example contains two scale layers: the ordinary subject
random-effect scale and the within-class `mix_scale`. The likelihood can learn
their product more readily than either factor. E-BFMI below 0.3, persistently
high R-hat, very small step sizes or trajectories reaching the maximum tree
depth should therefore be treated as evidence that the fitted posterior is not
yet reliable. Longer sampling must not be used to conceal a persistent scale
ridge.

# Class membership

```{r}
posterior_membership <- posterior_class(fit)
head(posterior_membership$subject)
plot(posterior_membership, domain = "subject")
```

The returned table reports the posterior mean class probability as `Estimate`,
its posterior standard deviation as `Est.Error`, its central interval and its
chain diagnostics. `assigned_class` is the class with the greatest posterior
mean probability for that subject.

The simulated labels may be joined only for recovery assessment:

```{r}
allocation_assessment <- transform(
  posterior_membership$subject,
  true_class = mixture_truth$allocation$subject[
    match(unit, names(mixture_truth$allocation$subject))
  ]
)

mean(
  allocation_assessment$class ==
    paste0("class_", allocation_assessment$true_class)
)
```

Hard allocation accuracy should be accompanied by posterior probability or a
proper multiclass score.

# Class-specific trajectories

```{r}
longitudinal_plot(
  fit,
  estimand = "mean_per_class"
)

longitudinal_plot(
  fit,
  estimand = "marginal_per_class"
)
```

`mean_per_class` inserts the ordered class centre into the selected subject
intercept coordinate. `marginal_per_class` integrates the inverse-link
trajectory over the within-class distribution and remaining random effects.

# Posterior predictive assessment

```{r}
pp_check(fit)
```

This single simulation is a software correspondence example. Claims about
class recovery or predictive performance require repeated independently
simulated datasets.
