---
title: "joinme: Usage Demo"
output-dir: ../docs/articles
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  cache = FALSE
)
options(width = 80)

library(dplyr)
repo_root <- normalizePath(if (dir.exists("R")) "." else "..", mustWork = FALSE)
if (requireNamespace("devtools", quietly = TRUE) && file.exists(file.path(repo_root, "DESCRIPTION"))) {
  devtools::load_all(repo_root, quiet = TRUE)
} else {
  library(joinme)
}

options(stan_preferred_engine = "cmdstanr")
options(stan_preferred_engine = 'rstan')
```

# Purpose and Reproducibility

This vignette demonstrates the usage of `joinme` with multiple scenarios,
including **data simulation**, **model fitting**, **diagnostic interpretation**,
**dynamic prediction**, **plotting**, and **threaded computation**. Each block is
designed to be reproducible and to illustrate a specific modelling choice.

For the methodological background, see the theory vignette in [joinme-theory.html](joinme-theory.html).

This demo covers mixed families (including double_exponential, skew_double_exponential,
beta, and cumulative_logit) and competing-risk survival via event type columns.

For skew_double_exponential, you can fix the quantile parameter with
`control = list(tau_sde_fixed = 0.2)` to model a specific quantile. If no
skew_double_exponential family is present, the control is ignored with a warning.

```{r seed}
set.seed(123)
```

# Quick model-spec checklist

Accepted specifications include:

- **Longitudinal structure** (`formulaLong`, `families`, optional `formulaDist`)
- **Survival structure** (`formulaEvent`, and optional baseline hazard controls like `basehaz`, `n_knots`)
- **Association structure** (`assoc` plus `transforms = joinme_tf(...)`)
- **Priors** (`priors = joinme_priors(beta = ..., alpha = ..., lkj = ...)`)
- **Computation** (`control` for sampling, and optionally threading)

## Family-specific links (new)

Longitudinal families can now carry per-marker inverse-links with `jm_family(...)`.
This affects both `epred` and `predict` scales in dynamic prediction.

```{r family-link-spec, eval=FALSE}
families_linked <- list(
  jm_family("student_t", link = "identity"),
  jm_family("bernoulli", link = "probit"),
  jm_family("poisson", inv_link = ~ exp(x))
)

fit_linked <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  families = families_linked,
  assoc = c("cv_total")
)
```

# Scenario A: Multivariate Continuous Markers with CV Association

## Data simulation

We simulate a multi-marker longitudinal dataset with a survival process linked through **current value (CV)**. The simulator also returns truth values for fixed effects and survival coefficients.

For marker-aggregated terms (`cv_total`, `cv_marker`, `cs_total`, `cs_marker`),
both simulation and Stan fit/prediction apply transforms at marker level before
weighted averaging. This keeps nonlinear association terms consistent end-to-end.

```{r scenario-a-sim}
sim <- simulate_joinme(
  n_id = 100,
  families = rep("student_t", 5),
  n_obs_per_marker_per_id = 10,
  times_obs = seq(0, 8, length.out = 14),
  quadrature_nodes = 31,
  seed = 123,
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.6)
)

# Gauss-Kronrod nodes/weights are fixed in Stan; `quadrature_nodes` selects
# the node count used to build the design matrices in R.

# Custom inverse-links can be supplied via jm_family(..., inv_link = ~ ...).

# When estimating marker weights, compare against sim$truth$marker_weights
# because weights are used directly in Stan.
```

## Model specification and fit

We start with identity transforms and a conservative sampling configuration to
keep runtime manageable. The nested marker block `( ... | marker )` supports an
inner `( ... | id )` term; this enables marker-by-id effects and the CORR
association.

Subject/group-weighted declarations can be made per grouping term with
`weighted(group, weights = <column>)`, e.g.
`(0 + x1 + (1 + time | weighted(id, weights = id_w)) | weighted(marker, weights = marker_w))`.

```{r scenario-a-fit}
formulaLong <- y ~ 1 + time + x1 +
  (1 + time || id) +
  (0 + x1 + (1 + time || id) || marker)

formulaEvent <- survival::Surv(time, event) ~ 1 + x1 + x2

fit <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  families = rep('student_t',5),
  transforms = joinme_tf(cv_total = "identity"),
  priors = joinme_priors(),
  control = list(
    parallel_chains = 2,
    iter_warmup = 300,
    iter_sampling = 700,
    seed = 123,
    refresh = 100,
    adapt_delta = 0.75,
    max_treedepth=12,
    init = 1
    # engine = 'rstan'
  )
)
```

To customise priors or the baseline hazard basis, supply `priors = joinme_priors(...)` and/or
forward standata arguments via `...`. Distributional regression for parameters
such as $\sigma$ or $\nu$ can be specified through `formulaDist`.

```{r scenario-a-fit2, eval=FALSE}
fit2 <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  transforms = joinme_tf(cv_total = "identity"),
  priors = joinme_priors(beta = list(scale = 1.5), alpha = list(scale = 1.5), lkj = 2),
  basehaz = "bs",
  n_knots = 5,
  basehaz_degree = 3,
  control = list(iter_warmup = 500, iter_sampling = 500, adapt_delta = 0.9)
)
```

## Diagnostics and parameter interpretation

We extract diagnostics (R-hat, ESS, divergences) and compare posterior summaries to known truth values. This yields **bias** and **coverage** diagnostics for fixed effects and survival coefficients.

```{r scenario-a-diagnostics}
sum_obj <- summary(fit)
print(sum_obj)

# Sampler diagnostics
diag_tbl <- as.data.frame(t(unlist(sum_obj$diagnostics)))
knitr::kable(diag_tbl, digits = 3, caption = "Sampler diagnostics (scenario A).")

# Fixed effects (beta)
fixef_tbl <- sum_obj$tables$fixef
truth_beta <- sim$truth$beta

fixef_tbl$truth <- truth_beta[fixef_tbl$term]
fixef_tbl$bias <- fixef_tbl$Estimate - fixef_tbl$truth
fixef_tbl$covered <- fixef_tbl$truth >= fixef_tbl$Q2.5 &
  fixef_tbl$truth <= fixef_tbl$Q97.5

fixef_view <- fixef_tbl[, c("term", "Estimate", "Q2.5", "Q97.5", "truth", "bias", "covered")]
knitr::kable(fixef_view, digits = 3, caption = "Fixed effects summary (scenario A).")

# Survival covariates (gamma_w)
if (!is.null(sum_obj$tables$gamma_w)) {
  gamma_tbl <- sum_obj$tables$gamma_w
  truth_gamma <- sim$truth$gamma_w

  gamma_tbl$truth <- truth_gamma[gamma_tbl$term]
  gamma_tbl$bias <- gamma_tbl$Estimate - gamma_tbl$truth
  gamma_tbl$covered <- gamma_tbl$truth >= gamma_tbl$Q2.5 &
    gamma_tbl$truth <= gamma_tbl$Q97.5

  gamma_view <- gamma_tbl[, c("term", "Estimate", "Q2.5", "Q97.5", "truth", "bias", "covered")]
  knitr::kable(gamma_view, digits = 3, caption = "Survival covariate summary (scenario A).")
}

# Association (aggregate CV total)
if (!is.null(sum_obj$tables$assoc)) {
  assoc_tbl <- sum_obj$tables$assoc
  alpha_total_hat <- assoc_tbl$Estimate[assoc_tbl$term %in% c("cv_total")]
  alpha_total_bias <- alpha_total_hat - sim$truth$alpha_cv_total
  list(alpha_total_hat = alpha_total_hat, alpha_total_bias = alpha_total_bias)
}
```

## RSS for fitted trajectories

The `JoinMeDynPred` object includes fitted values for observed history. We compute a **residual sum of squares** (RSS) on the expected-response scale (`epred`).

```{r scenario-a-rss}
pred_fit <- rstantools::posterior_epred(
  fit,
  newdataLong = sim$dataLong,
  newdataEvent = sim$dataEvent,
  time_start = max(sim$dataLong$time),
  times = seq(0, max(sim$dataLong$time) + 1, length.out = 60),
  control = list(
    n_samples = 100,
    iter_sampling = 500,
    iter_warmup = 400,
    chains = 3,
    threads_per_chain = 5
  )
)

fitted_ep <- pred_fit$predictions$longitudinal_fitted

if (!requireNamespace("dplyr", quietly = TRUE)) {
  stop("Install dplyr to run RSS calculation.")
}

library(dplyr)

rss_tbl <- fitted_ep %>%
  filter(scale == "epred") %>%
  inner_join(
    sim$dataLong,
    by = c("id", "marker", "time")
  ) %>%
  summarise(RSS = sum((y - Estimate)^2, na.rm = TRUE))

knitr::kable(rss_tbl, digits = 3, caption = "RSS for fitted trajectories (scenario A).")
```

## Dynamic prediction and plotting

We generate a future prediction for one subject and plot longitudinal and survival outputs.

```{r scenario-a-predict-plot}
#| fig-cap: "Dynamic prediction (scenario A)."
id <- 1
ndL <- sim$dataLong[sim$dataLong$id == id, ]
ndE <- sim$dataEvent[sim$dataEvent$id == id, ]

time_start <- max(ndL$time)
times <- seq(time_start, time_start + 2, length.out = 50)

pred_ep <- rstantools::posterior_epred(
  fit,
  newdataLong = ndL,
  newdataEvent = ndE,
  time_start = time_start,
  times = times,
  control = list(
    n_samples = 200,
    n_pred_draws = 400,
    iter_sampling = 500,
    iter_warmup = 400,
    threads_per_chain = 5,
    chains = 2
  )
)

plot(pred_ep, type = "longitudinal")
plot(pred_ep, type = "survival")
```

`predict()` now accepts `scale` as a character vector. If omitted, all
longitudinal scales are produced: `c("epred", "linpred", "predict")`.
Use `plot(..., scale = "epred")` (or another available scale) to choose which
trajectory scale to visualise.

Dynamic prediction is conditional on observed history up to `time_start` and (by construction) survival up to that landmark; the survival plot corresponds to $\Pr(T > t \mid T > T_{start}, \mathcal{H}(T_{start}))$.

`control$n_samples` controls how many posterior parameter draws are extracted
from the fitted model, while `control$n_pred_draws` controls how many draws are produced
in the dynpred output object. This decoupling allows lightweight posterior
extraction with a larger prediction Monte Carlo layer when needed.

When `times = NULL`, you can now control the default forecast window with
`time_horizon` (defaulting to `tmax` from the fitted model):

```{r scenario-a-predict-horizon}
pred_ep_horizon <- rstantools::posterior_epred(
  fit,
  newdataLong = ndL,
  newdataEvent = ndE,
  time_start = time_start,
  times = NULL,
  time_horizon = 2,
  control = list(
    n_samples = 100,
    n_pred_draws = 150,
    iter_sampling = 200,
    iter_warmup = 200,
    threads_per_chain = 2,
    chains = 1
  )
)

head(pred_ep_horizon$predictions$survival)
```

Prediction enforces marker-level consistency: `newdataLong$marker` must use only levels seen during fitting. Internally, dynamic prediction outputs are stabilised by averaging over dynpred posterior rows for each indexed draw before constructing quantile summaries.

### Diagnostics and random-effects extraction

`joinme` now uses the same diagnostics table schema for fitted and predicted
objects.

```{r scenario-a-diagnosis}
diagnosis(fit)
diagnosis(pred_ep)
```

For fitted models, `ranef()` and `corr()` are nested by `formulaLong` and
`formulaDist`:

```{r scenario-a-ranef-corr-fit}
re_fit <- ranef(fit)
vc_fit <- corr(fit)

# Examples
re_fit$formulaLong$id
vc_fit$formulaLong$id
```

For dynamic prediction, marker-by-id random effects and covariance are exposed
only when marker covariance depends on id:

```{r scenario-a-ranef-corr-pred}
if (isTRUE(pred_ep$metadata$marker_corr_depends_on_id)) {
  ranef(pred_ep)
  corr(pred_ep)
}
```

# Scenario B: Competing-Risk Survival

This example adds a competing-risk indicator via `event_type`. The dynamic
survival output corresponds to overall survival across causes.

```{r scenario-b-competing-sim, eval=FALSE}
library(joinme)

sim_comp <- simulate_joinme(
  n_id = 40,
  families = rep("student_t", 3),
  n_obs_per_marker_per_id = 6,
  times_obs = seq(0, 8, length.out = 12),
  seed = 444,
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.6)
)

# Simulate cause-specific event times and derive event_type from the minimum time.
set.seed(444)
cause_levels <- c("cause1", "cause2")
time_cens <- 8.0

h0_list <- list(
  cause1 = weibull_h0(shape = 1.2, scale = 5.0),
  cause2 = weibull_h0(shape = 2.0, scale = 7.0)
)

gamma_w_mat <- rbind(
  cause1 = c("(Intercept)" = -0.2, "x1" = 0.4, "x2" = -0.3),
  cause2 = c("(Intercept)" = 0.1, "x1" = -0.2, "x2" = 0.5)
)

find_bracket <- function(f, lower = 0, upper = 1, upper_max = 50, expand = 1.7, max_expand = 60L) {
  f_lower <- f(lower)
  if (!is.finite(f_lower) || f_lower > 0) return(NULL)
  f_upper <- f(upper)
  if (!is.finite(f_upper)) return(NULL)
  it <- 0L
  while (f_upper < 0 && upper < upper_max && it < max_expand) {
    upper <- min(upper_max, upper * expand)
    f_upper <- f(upper)
    if (!is.finite(f_upper)) return(NULL)
    it <- it + 1L
  }
  if (f_upper >= 0) list(lower = lower, upper = upper) else NULL
}

draw_cause_time <- function(i, k) {
  haz_row <- c("(Intercept)" = 1, "x1" = sim_comp$dataEvent$x1[i], "x2" = sim_comp$dataEvent$x2[i])
  eta_w <- sum(haz_row * gamma_w_mat[k, ])
  alpha_cv <- sim_comp$truth$alpha_cv_total
  h0 <- h0_list[[k]]

  hazard_i <- function(t) h0(t) * exp(eta_w + alpha_cv * sim_comp$helpers$cv_total(i, t))
  cumhaz_i <- function(t) {
    if (t <= 0) return(0)
    out <- integrate(hazard_i, lower = 0, upper = t, rel.tol = 1e-6, subdivisions = 2000L)
    as.numeric(out$value)
  }

  U <- runif(1)
  target <- -log(U)
  f_root <- function(t) cumhaz_i(t) - target
  br <- find_bracket(f_root, lower = 0, upper = 1, upper_max = 50)
  if (is.null(br)) return(list(time = time_cens, event = 0L))
  T <- uniroot(f_root, lower = br$lower, upper = br$upper)$root
  if (T > time_cens) list(time = time_cens, event = 0L) else list(time = T, event = 1L)
}

event_time <- numeric(nrow(sim_comp$dataEvent))
event_type <- character(nrow(sim_comp$dataEvent))
event_flag <- integer(nrow(sim_comp$dataEvent))

for (i in seq_len(nrow(sim_comp$dataEvent))) {
  times_k <- vapply(cause_levels, function(k) draw_cause_time(i, k)$time, numeric(1))
  min_idx <- which.min(times_k)
  event_time[i] <- times_k[min_idx]
  event_flag[i] <- as.integer(event_time[i] < time_cens)
  event_type[i] <- cause_levels[min_idx]
  if (event_flag[i] == 0L) {
    event_type[i] <- cause_levels[1]
  }
}

sim_comp$dataEvent$time <- event_time
sim_comp$dataEvent$event <- event_flag
sim_comp$dataEvent$event_type <- factor(event_type, levels = cause_levels)

sim_comp$dataLong <- sim_comp$dataLong |>
  dplyr::inner_join(sim_comp$dataEvent[, c("id", "time")], by = "id", suffix = c("", "_event")) |>
  dplyr::filter(time <= time_event) |>
  dplyr::select(-time_event)
```

```{r scenario-b-competing-fit, eval=FALSE}
formulaLong_comp <- y ~ 1 + time + x1 +
  (1 + time || id) +
  (0 + x1 + (1 + time || id) || marker)

formulaEvent_comp <- survival::Surv(time, event) ~ 1 + x1 + x2

fit_comp <- joinme(
  formulaLong = formulaLong_comp,
  dataLong = sim_comp$dataLong,
  formulaEvent = formulaEvent_comp,
  dataEvent = sim_comp$dataEvent,
  assoc = c("cv_total"),
  families = rep("student_t", 3),
  transforms = joinme_tf(cv_total = "identity"),
  event_type_var = "event_type",
  control = list(
    engine = "cmdstanr",
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 150,
    iter_sampling = 150,
    seed = 444,
    refresh = 0,
    adapt_delta = 0.85,
    max_treedepth = 10
  )
)
```

Prediction is shown below but left unevaluated during rendering. Set
`eval = TRUE` once CmdStan is configured for dynamic prediction.

```{r scenario-b-competing-predict, eval=FALSE}
id_comp <- 1
ndL_comp <- sim_comp$dataLong[sim_comp$dataLong$id == id_comp, ]
ndE_comp <- sim_comp$dataEvent[sim_comp$dataEvent$id == id_comp, ]

time_start_comp <- max(ndL_comp$time)
times_comp <- seq(time_start_comp, time_start_comp + 2, length.out = 40)

pred_comp <- predict(
  fit_comp,
  newdataLong = ndL_comp,
  newdataEvent = ndE_comp,
  process = "event",
  time_start = time_start_comp,
  times = times_comp,
  control = list(
    n_samples = 100,
    engine = "cmdstanr",
    chains = 1,
    iter_warmup = 0,
    iter_sampling = 1,
    fixed_param = TRUE,
    refresh = 0
  )
)

plot(pred_comp, type = "survival")
```

# Scenario C: Mixed Families and Nonlinear Association

This example uses **mixed families** and a **nonlinear association transform**
to illustrate interpretability when markers live on distinct scales. The goal
is to show how different distributional families can coexist within a single
joint model.

```{r scenario-b-mixed}
#| eval: false

sim_mixed <- simulate_joinme(
  n_id = 20,
  n_obs_per_marker_per_id = 5,
  families = c("gaussian", "bernoulli", "poisson"),
  times_obs = seq(0, 4, length.out = 5),
  seed = 101
)

formulaLong_mixed <- y ~ 1 + time + x1 +
  (1 + time || id) +
  (0 + x1 + (1 + time || id) || marker)

formulaEvent_mixed <- survival::Surv(time, event) ~ 1 + x1 + x2

transforms <- joinme_tf(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.3, 0.7, 1.1, 1.3, 1.4),
    degree = 3
  )
)

# Notes on spline arguments:
# - type = "ispline": provide knots + coeff directly (no x/y/lambda fitting).
# - type = "ispline_penalised": provide knots (or n_knots) plus lambda.
#   Supplying y uses the legacy (x, y) plug-in fit; omitting y lets Stan
#   estimate the monotone spline jointly.
```

## Optional: distributional regression

If a marker family supports additional parameters (for example, Student-$t$
scale or degrees of freedom, negative binomial dispersion), you can model these
via `formulaDist`. Example (heteroscedastic scale):

```{r scenario-b-dist}
#| eval: false
fit_dist <- joinme(
  formulaLong = formulaLong,
  dataLong = sim$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  transforms = joinme_tf(cv_total = "identity"),
  families = rep("student_t", 4),
  formulaDist = list(sigma = ~ 1 + time),
  control = list(iter_warmup = 300, iter_sampling = 300, seed = 123, refresh = 0)
)

fit_mixed <- joinme(
  formulaLong = formulaLong_mixed,
  dataLong = sim_mixed$dataLong,
  formulaEvent = formulaEvent_mixed,
  dataEvent = sim_mixed$dataEvent,
  families = sim_mixed$marker_info$families,
  transforms = transforms,
  assoc = c("cv_total"),
  control = list(
    parallel_chains = 1,
    iter_warmup = 200,
    iter_sampling = 200,
    seed = 101,
    refresh = 0,
    adapt_delta = 0.95
  )
)

sum_mixed <- summary(fit_mixed)
diag_mixed <- as.data.frame(t(unlist(sum_mixed$diagnostics)))
knitr::kable(diag_mixed, digits = 3, caption = "Sampler diagnostics (scenario B).")

if (!is.null(sum_mixed$tables$fixef)) {
  fixef_mixed <- sum_mixed$tables$fixef
  fixef_mixed <- fixef_mixed[, c("term", "Estimate", "Q2.5", "Q97.5", "Rhat", "ESS")]
  knitr::kable(fixef_mixed, digits = 3, caption = "Fixed effects summary (scenario B).")
}
```

# Scenario D: Threaded Fit and Threaded Prediction

Threading is useful when the longitudinal dimension is large. This example shows a threaded fit and dynamic prediction using two threads.

```{r scenario-c-threaded}
#| eval: false
sim_thread <- simulate_joinme(
  n_id = 60,
  families = rep("student_t", 6),
  n_obs_per_marker_per_id = 6,
  times_obs = seq(0, 8, length.out = 12),
  seed = 202,
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.6)
)

fit_thread <- joinme(
  formulaLong = formulaLong,
  dataLong = sim_thread$dataLong,
  formulaEvent = formulaEvent,
  dataEvent = sim_thread$dataEvent,
  assoc = c("cv_total"),
  transforms = joinme_tf(cv_total = "identity"),
  control = list(
    parallel_chains = 1,
    iter_warmup = 200,
    iter_sampling = 200,
    seed = 202,
    refresh = 0,
    threads_per_chain = 2,
    adapt_delta = 0.9
  )
)

ndL_thread <- sim_thread$dataLong[sim_thread$dataLong$id %in% c(1, 2), ]
ndE_thread <- sim_thread$dataEvent[sim_thread$dataEvent$id %in% c(1, 2), ]

pred_thread <- rstantools::posterior_epred(
  fit_thread,
  newdataLong = ndL_thread,
  newdataEvent = ndE_thread,
  time_start = max(ndL_thread$time),
  times = seq(0, max(ndL_thread$time) + 2, length.out = 60),
  control = list(
    n_samples = 100,
    threads_per_chain = 2,
    chains = 1,
    iter_warmup = 100,
    iter_sampling = 1,
    seed = 202,
    refresh = 0
  )
)

plot(pred_thread, type = "longitudinal")
plot(pred_thread, type = "survival")
```

When requesting multiple outcomes with `combined = TRUE`, the method returns:

- one combined object for a single subject when `patchwork`/`cowplot` is available,
- a named list of combined objects (one per subject) for multiple subjects,
- and, if combiner backends are unavailable, a per-subject fallback to the
  regular nested outcome list.

# Discussion and Reporting Guidance

These scenarios illustrate how model assumptions and association choices influence interpretation. In a report or manuscript, we recommend:

- Reporting **R-hat**, **ESS**, and divergences as primary convergence diagnostics.
- Presenting **bias and coverage** for simulated truth when validation is possible.
- Summarizing **association parameters** alongside the transform used (functional, spline, piecewise).
- Including **dynamic prediction plots** to show how uncertainty evolves beyond the observed history.
- Stating whether **threading** was used and the number of threads per chain.

This structure provides a reproducible, interpretable workflow that can be adapted to substantive scientific questions.
