---
title: "simulate_JoiNMe Cookbook"
output-dir: ../docs/tutorials
format:
  html:
    toc: true
    number-sections: true
    code-fold: false
    embed-resources: true
    theme: cosmo
execute:
  eval: false
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
repo_root <- normalizePath(if (dir.exists("R")) "." else "..", mustWork = FALSE)
if (requireNamespace("devtools", quietly = TRUE) && file.exists(file.path(repo_root, "DESCRIPTION"))) {
  devtools::load_all(repo_root, quiet = TRUE)
} else {
  try(base::attachNamespace("JoiNMe"), silent = TRUE)
}
library(survival)
```

# Purpose

This cookbook gives practical, copy-paste patterns for `simulate_joinme()`.
It is organised by common simulation goals:

- selecting association channels (`assoc`, `formulaAssoc`, `assoc_coefs`),
- configuring random effects (`re_params`),
- controlling marker-by-id covariance regression (`formulaVCov` + `re_params$id_marker_cov`),
- matching fitted marker-weight shrinkage (`fixed_marker_weights`, `shrinkage`),
- defining distributional regressions (`formulaDist`) with family-scoped coefficients (`dist_coefs`),
- generating event-level covariates (`covariate_formulas`),
- checking outputs and truth objects.

The examples intentionally mirror the `joinme()` fitting syntax so your simulation and fitting pipelines stay aligned.

That alignment now extends to the posterior layer as well: the recommended
follow-up after fitting a simulated dataset is to use `summary()`, `draws()`,
and `mcmc_plot()` rather than inspecting raw Stan variables directly.

The fit class referenced throughout the examples below is `JoiNMeFit`.

# Quick mental model

At a high level, `simulate_joinme()` does this:

1. Parse formulas (`formulaLong`, `formulaEvent`, optional `formulaDist`, `formulaVCov`).
2. Simulate event-level covariates from `covariate_formulas`.
3. Draw random effects from `re_params`.
4. Build longitudinal trajectories.
5. Build hazard from baseline + association channels.
6. Draw event/censoring times.
7. Draw marker outcomes by `families` and distributional parameters.

Returned object:

- `sim$dataLong`: longitudinal table,
- `sim$dataEvent`: event table,
- `sim$truth`: parameters and configs used in generation,
- `sim$helpers`: helper functions (e.g., hazard and association component probes),
- `sim$tmax`: scaling horizon.

For recovery work, `sim$truth$baseline_hazard` records the resolved baseline
type, parameters, formula, and coefficients. A constant rate of `0.1`, for
example, is also exposed as the fitted log-hazard coefficient `log(0.1)` in
`sim$truth$stan_fit$bs_gamma_c`.

# Minimal baseline recipe

```{r}
set.seed(1)

sim <- simulate_joinme(
  n_id = 40,
  families = c("gaussian", "student_t", "skew_normal"),
  n_obs_per_marker_per_id = 6,
  times_obs = seq(0, 5, length.out = 8),
  assoc = c("cv_total"),
  assoc_coefs = c(cv_total = 0.5),
  seed = 1
)

str(sim$dataLong)
str(sim$dataEvent)
names(sim$truth)
```

## Match marker-weight generation to the fitted prior

Estimated marker weights use the same additive hierarchy in simulation and
Stan. The default is a zero base plus standardized Student-$t_6$ latents:

```{r}
sim_weights <- simulate_joinme(
  n_id = 50,
  families = rep("student_t", 3),
  assoc = "cv_total",
  fixed_marker_weights = FALSE,
  shrinkage = 0L,
  seed = 14
)

sim_weights$truth$marker_weights_base
sim_weights$truth$marker_weights_latent
sim_weights$truth$marker_weights       # effective weights used in the hazard
```

Pass the same switch to fitting:

```{r eval=FALSE}
fit_weights <- joinme(
  formulaLong = sim_weights$truth$formulaLong,
  dataLong = sim_weights$dataLong,
  formulaEvent = sim_weights$truth$formulaEvent,
  dataEvent = sim_weights$dataEvent,
  assoc = "cv_total",
  fixed_marker_weights = FALSE,
  shrinkage = 0L
)
```

The complete mapping is `0 = Student-t(6, 0, 1)`,
`1 = double_exponential(0, 1)`, and `2 = normal(0, 1)`. If you instead want
known fixed weights, supply them and set `fixed_marker_weights = TRUE`; then the
latent truth is exactly zero.

## Left truncation and interval-split event rows

`simulate_joinme()` can emit counting-process event data for
`Surv(time_start, time_stop, event)` workflows.

```{r}
sim_split <- simulate_joinme(
  n_id = 30,
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x1 + x2,
  left_truncation_max = 1.0,
  seed = 2
)

head(sim_split$dataEvent)
```

Use the same split `formulaEvent` when fitting:

```{r eval=FALSE}
fit_split <- joinme(
  formulaLong = y ~ 1 + time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  dataLong = sim_split$dataLong,
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x1 + x2,
  dataEvent = sim_split$dataEvent,
  assoc = c("cv_total")
)
```

## Time-varying covariates via `time_varyring(fun, n_step)`

Stepwise event-side covariates can now be declared directly in
`covariate_formulas`.

- If `n_step` has length 1, every subject gets that number of periods.
- If `n_step` has length 2, each subject draws its period count uniformly from
  `min(n_step):max(n_step)`.

When a time-varying covariate is present, `simulate_joinme()` emits split
counting-process rows automatically.

```{r}
sim_tv <- simulate_joinme(
  n_id = 25,
  formulaEvent = survival::Surv(time_start, time_stop, event) ~ x + x2,
  covariate_formulas = list(
    x ~ time_varyring(rnorm, c(3, 6), mean = 0, sd = 1),
    x2 ~ rnorm(n_id)
  ),
  left_truncation_max = 0.5,
  seed = 22
)

head(sim_tv$dataEvent)
```

# From simulation to reporting-scale posterior summaries

The main idea of a simulation study is not only to fit the model, but also to
recover known parameters using the same public summaries you would use in a real
analysis. The minimal reporting workflow looks like this:

```{r eval=FALSE}
fit <- joinme(
  formulaLong = y ~ 1 + time + x1 + (1 + time || id) + (0 + x1 + (1 + time || id) || marker),
  dataLong = sim$dataLong,
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  dataEvent = sim$dataEvent,
  assoc = c("cv_total"),
  transforms = joinme_tf(cv_total = "identity"),
  control = list(chains = 1, parallel_chains = 1, iter_warmup = 200, iter_sampling = 200)
)

summary(fit)
draws(fit, variables = c("time", "alpha_cv_total"), format = "draws_df")
mcmc_plot(fit, variable = c("time", "alpha_cv_total"), type = "intervals")
```

This workflow keeps simulation recovery tables, posterior graphics, and fitted
plots on the same user-facing parameter names.

# `draws()` versus `extract()`

These two functions both expose posterior information, but they are intended
for different jobs.

## Mental model

- `draws()` is the convenient, analysis-facing interface.
- `extract()` is the structured, component-facing interface.

In practice:

- use `draws()` when you want one posterior object you can pass straight to
  `posterior::summarise_draws()`, `posterior::subset_draws()`, or
  `mcmc_plot()`,
- use `extract()` when you want one semantic block of the model together with
  the metadata describing how that block was assembled.

## What `draws()` returns

`draws()` always returns a standard posterior object in one of three formats:

- `"draws_array"`
- `"draws_matrix"`
- `"draws_df"`

For fitted models, the variables are renamed onto the same user-facing scale as
`summary(fit)`. For example, raw Stan variables such as `beta[2]` or
`gamma_w[1,2]` are relabelled to readable terms such as `time` or
`event1: x1` when that mapping is available.

For prediction objects, `draws()` flattens stored prediction blocks into a
single posterior object with explicit labels containing id, scale, marker, and
time information.

```{r eval=FALSE}
# one posterior object, ready for posterior/bayesplot workflows
d_fit <- draws(fit, format = "draws_df")
d_assoc <- draws(fit, variables = "^alpha_", regex = TRUE, format = "draws_matrix")

posterior::summarise_draws(d_assoc)
mcmc_plot(fit, variable = "alpha_cv_total", type = "trace")
```

## What `extract()` returns

`extract()` returns a list, not a single posterior object. The exact structure
depends on the object and the requested component, but the core idea is the
same: keep the extraction tied to a model block.

For fitted models, `extract()` lets you request one component at a time:

- `what = "fixef"`
- `what = "gamma_w"`
- `what = "assoc"`
- `what = "distributional"`
- `what = "distributional_regression"`
- `what = "likelihood_scale"`
- `what = "raw"`
- `what = "association_plot"`

The returned list includes `draws` and `term_map`, and in the
`association_plot` case it also includes support ranges and transform payloads
used by fitted association plotting.

For prediction objects, `extract()` keeps the original stored block structure,
so you can request `longitudinal`, `survival`, `cumhaz`, or random-effect draw
blocks before flattening.

```{r eval=FALSE}
# component-aware extraction with explicit mapping metadata
ext_fixef <- extract(fit, what = "fixef")
ext_assoc <- extract(fit, what = "assoc", term = c("cv_total", "cs_total"))

str(ext_fixef$draws)
ext_fixef$term_map

# prediction blocks remain separated
ext_surv <- extract(pred, what = "survival", id = 1)
str(ext_surv$draws)
```

## Rule of thumb

Use `draws()` if your next step is one of these:

- posterior summarisation,
- MCMC diagnostics or plotting,
- regex-based variable selection,
- exporting a single posterior object for downstream tooling.

Use `extract()` if your next step is one of these:

- inspecting one model component in isolation,
- checking how summary labels map to underlying Stan variables,
- using the special `association_plot` payload,
- working with dynamic-prediction draw blocks before flattening.

The two functions are therefore complementary rather than redundant:

- `extract()` is the component-aware extraction layer,
- `draws()` is the standardised posterior-consumption layer built on top of the
  same naming logic.

# Association cookbook (`assoc`, `formulaAssoc`, `assoc_coefs`)

## Channel meaning

Available channels:

- `cv_mean`: mean current value across markers,
- `cv_marker`: marker-weighted marker-level current value,
- `cv_total`: marker-level total current value,
- `cs_mean`: mean current slope across markers,
- `cs_marker`: marker-weighted marker-level current slope,
- `cs_total`: marker-level total current slope,
- `corr`: off-diagonal entries of the subject-specific Cholesky-correlation
  factor `K` from marker-by-id random effects,
- `vcov`: those same off-diagonal `K` entries together with the
  subject-specific standard deviations.

## Named vector/list coefficient syntax

```{r}
sim <- simulate_joinme(
  n_id = 60,
  families = c("gaussian", "gaussian", "gaussian"),
  assoc = c("cv_total", "cs_mean", "corr"),
  assoc_coefs = list(
    cv_total = 0.4,
    cs_mean = -0.2,
    corr = c(0.15, -0.10, 0.05)
  seed = 2
)
```

## `formulaAssoc` overrides `assoc`

```{r}
sim <- simulate_joinme(
  n_id = 50,
  families = c("gaussian", "gaussian"),
  assoc = c("cv_total", "cs_total"),
  formulaAssoc = ~ cv_mean + vcov,
  assoc_coefs = c(cv_mean = 0.3, vcov = 0.2),
  seed = 3
)
```

## Functional transforms on association channels

```{r}
sim <- simulate_joinme(
  n_id = 30,
  families = c("gaussian", "gaussian"),
  assoc = c("cv_mean", "corr"),
  assoc_coefs = c(cv_mean = 0.2, corr = -0.15),
  transforms = joinme_tf(
    cv_mean = ~ log1p_exp(x),
    corr = ~ -x
  ),
  seed = 4
)
```

## I-spline and penalised I-spline arguments (`x`, `y`, `lambda`)

### Quick defaults by transformation type

- `identity`: `list(type = "identity")`, or omit the term entirely.
- `functional`: `list(type = "functional", expr = ~ log1p(x))`; `expr` is required.
- `ispline`: `list(type = "ispline", knots = ..., coeff = ..., degree = 3)`;
  default `degree = 3`.
- `ispline_penalised`: `list(type = "ispline_penalised", ..., n_knots = 6, degree = 3, lambda = 1)`;
  defaults are `n_knots = 6`, `degree = 3`, `lambda = 1`.
- `pwlin`: `list(type = "pwlin", x = ..., y = ...)`; no additional defaults.

For monotone spline transforms, there are two modes:

- `type = "ispline"`: you provide the spline shape directly via
  `knots` + `coeff` (+ optional `degree`).
- `type = "ispline_penalised"`: you provide `knots` (or `n_knots`) and
  `lambda`. If you also provide `y`, the package fits a monotone spline to the
  training pairs `(x, y)` in R. If you omit `y`, Stan estimates the monotone
  spline coefficients directly.

Meaning of arguments in `ispline_penalised`:

- `x`: input values on the raw association-feature scale (for example raw
  `cv_total` values). These are used either as training inputs or to help pick
  knot locations.
- `y`: optional desired transformed output at those `x` values. Supplying `y`
  uses the legacy plug-in fit; omitting it uses Stan estimation.
- `lambda`: smoothness penalty strength for the fitted spline coefficients.

Interpretation of `lambda`:

- smaller `lambda` (e.g., `0` to `0.1`) = more flexible curve,
- larger `lambda` (e.g., `1` to `10`) = smoother, less wiggly curve.

```{r}
# Identity: default when omitted
tf_identity <- list(
  cv_total = list(type = "identity")
)

# Functional: expr is required
tf_functional <- list(
  cv_total = list(
    type = "functional",
    expr = ~ log1p(x)
  )
)

# Direct I-spline: shape is fully user-specified by knots + coeff
tf_ispline <- list(
  cv_total = list(
    type = "ispline",
    knots = c(-1, 0, 1),
    coeff = c(0, 0.2, 0.5, 0.9, 1.2),
    degree = 3
  )
)

# Penalised I-spline: legacy plug-in mode using (x, y) and smoothness lambda
tf_ispline_pen <- list(
  cv_total = list(
    type = "ispline_penalised",
    x = c(-2, -1, 0, 1, 2),
    y = c(0.0, 0.2, 0.5, 0.9, 1.1),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)

# Penalised I-spline: Stan-estimated mode (omit y)
tf_ispline_pen_stan <- list(
  cv_total = list(
    type = "ispline_penalised",
    x = c(-2, -1, 0, 1, 2),
    n_knots = 6,
    degree = 3,
    lambda = 1.0
  )
)

# Piecewise-linear: x and y are both required
tf_pwlin <- list(
  cv_total = list(
    type = "pwlin",
    x = c(-2, -1, 0, 1, 2),
    y = c(0.2, 0.5, 1.0, 0.5, 0.2)
  )
)
```

# Random-effects cookbook (`re_params`)

`re_params` contains simulation controls for all random-effect blocks:

- `id`: from top-level `( ... | id)` terms,
- `marker`: from marker-level terms,
- `id_marker_cov`: subject-specific covariance regression for marker-by-id latents,
- `dist`: random effects in distributional regression formulas.

## `id` and `marker`

Each uses:

- `sd`: scalar or vector (length = block dimension),
- `corr`: square correlation matrix.

```{r}
re_cfg <- list(
  id = list(
    sd = c(0.6, 0.3),
    corr = matrix(c(1, 0.25,
                    0.25, 1), 2, 2)
  ),
  marker = list(sd = 0.4),
  id_marker_cov = list(
    latent = list(
      sd = c(0.7, 0.2),
      corr = diag(2)
    )
  )
)

sim <- simulate_joinme(
  n_id = 50,
  families = c("gaussian", "gaussian", "gaussian"),
  re_params = re_cfg,
  seed = 5
)
```

## `id_marker_cov` explained

This block controls the subject-specific lower-triangular matrix `L_i` used to scale marker-by-id latent random effects.

Pipeline view:

1. Draw marker-by-id latent seeds from iid standard normal values.
2. Build subject-specific `L_i` from `alpha`, `beta`, `lambda`, `diag_link`.
3. Produce final marker-by-id effects by multiplying those latent seeds with `L_i`.

This keeps the baseline covariance and the subject-specific covariance changes in one place: `L_i`.

Backward-compatibility note:

- If you still provide `re_params$id_marker_cov$latent`, the simulator translates that old baseline covariance into the baseline `alpha` intercepts before simulation.
- New code should set `alpha` directly and leave `latent` unset.

Recovery tip:

- If you want to validate coefficient recovery for a covariance-style channel,
  start with a fixed transform (for example `type = "ispline_expit"` with
  explicit `coeff`) and a low-dimensional `vcov` setup.
- Jointly estimating a penalised spline and the association coefficient is a
  harder inference problem; it is suitable for flexible modelling, but not the
  cleanest first recovery check.

Mathematically:

$$
z_{id} \sim \mathcal{N}(0, I),
$$

where $z_{id}$ is the marker-by-id latent seed vector for subject $i$ and marker $d$.

For each subject $i$ and lower-triangular element index $m$ (one diagonal or off-diagonal entry), simulation builds

$$
\eta_{im} = \alpha_m + x_i^\top \beta_m + \lambda_m z_i, \quad z_i \sim \mathcal{N}(0, 1),
$$

then maps diagonal entries through `diag_link` (`softplus` or `exp`) so diagonals are positive. Off-diagonals are left on identity scale.

Then final marker-by-id effects are

$$
b_{id} = L_i z_{id}
$$

so `alpha`/`beta`/`lambda` fully govern how $L_i$ shapes marker-by-id covariance by subject.

Fields:

- `alpha`: baseline level of each covariance element when `formulaVCov` covariates are zero and random perturbation is zero. Larger diagonal `alpha` means larger typical within-subject marker-by-id SD after the diagonal link.
- `beta`: covariate effects from `formulaVCov`; shifts covariance elements systematically across subjects by observed covariates (`x1`, `x2`, ...).
- `lambda`: multiplier on the latent subject perturbation; controls how strongly unobserved subject heterogeneity affects each covariance element.
- `diag_link`: link for diagonal elements (`"softplus"` or `"exp"`).

Practical interpretation of variability:

- Larger `|lambda|` increases subject-to-subject covariance heterogeneity.
- Set `lambda` near zero to make covariance elements mostly deterministic from `alpha + X beta`.

Dimension rule:

- full covariance: $M = Q_{idm}(Q_{idm}+1)/2$ elements,
- diagonal covariance (`||` in nested marker-id term): $M = Q_{idm}$ elements.

```{r}
sim <- simulate_joinme(
  n_id = 80,
  families = c("gaussian", "gaussian", "gaussian"),
  formulaVCov = ~ x1 + x2,
  re_params = list(
    id = list(sd = c(0.5, 0.2)),
    marker = list(sd = 0.3),
    id_marker_cov = list(
      latent = list(sd = c(0.6, 0.25)),
      alpha = c(-0.2, 0.1, -0.05),
      beta = matrix(c(0.20, 0.00,
                      0.05, 0.05,
                      0.10, -0.05), nrow = 3, byrow = TRUE),
      lambda = c(0.4, 0.4, 0.4),
      diag_link = "softplus"
    )
  ),
  assoc = c("corr"),
  assoc_coefs = list(corr = c(0.2)),
  seed = 6
)
```

# Distributional regression cookbook (`formulaDist`, `dist_coefs`, `re_params$dist`)

## Family-scoped formulas

```{r}
form_dist <- list(
  sigma[family = gaussian] ~ 1 + x1,
  sigma[family = student_t] ~ 1 + time,
  nu[family = student_t] ~ 1,
  alpha_skew[family = skew_normal] ~ 1
)
```

## Family-specific fixed coefficients in `dist_coefs`

You can assign different fixed coefficients by family for the same parameter.

```{r}
dist_beta <- list(
  sigma = list(
    gaussian = c("(Intercept)" = log(0.2), x1 = 0.3),
    student_t = c("(Intercept)" = log(0.8), time = 0.4)
  ),
  nu = c("(Intercept)" = log(6 - 2))
)

sim <- simulate_joinme(
  n_id = 80,
  families = c("gaussian", "student_t", "skew_normal"),
  formulaDist = form_dist,
  dist_coefs = dist_beta,
  seed = 7
)
```

Alternative alias keys are also accepted:

```{r}
dist_beta_alias <- list(
  "sigma[family='gaussian']" = c("(Intercept)" = log(0.2), x1 = 0.3),
  "sigma[family='student_t']" = c("(Intercept)" = log(0.8), time = 0.4)
)
```

## Distributional random effects via `re_params$dist`

Use `re_params$dist[[param]]` to control random effects in distributional formulas.

```{r}
sim <- simulate_joinme(
  n_id = 60,
  families = c("student_t", "student_t"),
  formulaDist = list(
    sigma ~ 1 + (1 | id),
    nu ~ 1
  ),
  re_params = list(
    id = list(sd = c(0.4, 0.2)),
    marker = list(sd = 0.2),
    id_marker_cov = list(latent = list(sd = c(0.4, 0.2))),
    dist = list(
      sigma = list(sd = 0.7)
    )
  ),
  seed = 8
)
```

Term-specific controls:

```{r}
re_dist <- list(
  sigma = list(
    terms = list(
      list(sd = 0.5),
      list(sd = 0.2)
    )
  )
)
```

# Covariate generation cookbook (`covariate_formulas`)

`covariate_formulas` are evaluated once per subject and merged into both event and longitudinal rows.

```{r}
sim <- simulate_joinme(
  n_id = 50,
  covariate_formulas = list(
    x1 ~ rnorm(n_id, 0, 1),
    x2 ~ rt(n_id, df = 5),
    trt ~ rbinom(n_id, size = 1, prob = 0.5)
  ),
  formulaEvent = survival::Surv(time, event) ~ x1 + x2 + trt,
  seed = 9
)
```

# End-to-end reproducible template

```{r}
set.seed(2026)

sim <- simulate_joinme(
  formulaLong = y ~ 1 + time + x1 + (1 + time | id) + (0 + x1 + (1 + time | id) | marker),
  formulaEvent = survival::Surv(time, event) ~ 1 + x1 + x2,
  formulaVCov = ~ x1 + x2,
  formulaDist = list(
    sigma[family = gaussian] ~ 1 + x1,
    sigma[family = student_t] ~ 1 + time,
    nu[family = student_t] ~ 1
  ),
  families = c("gaussian", "student_t", "student_t"),
  n_id = 100,
  n_obs_per_marker_per_id = 8,
  times_obs = seq(0, 6, length.out = 10),
  assoc = c("cv_total", "corr", "vcov"),
  assoc_coefs = list(cv_total = 0.35, corr = c(0.1), vcov = c(0.3, 0.05, 0.2)),
  transforms = joinme_tf(corr = ~ -x, vcov = "identity"),
  dist_coefs = list(
    sigma = list(
      gaussian = c("(Intercept)" = log(0.25), x1 = 0.2),
      student_t = c("(Intercept)" = log(0.9), time = 0.3)
    ),
    nu = c("(Intercept)" = log(7 - 2))
  ),
  re_params = list(
    id = list(sd = c(0.6, 0.3)),
    marker = list(sd = 0.25),
    id_marker_cov = list(
      latent = list(sd = c(0.7, 0.3)),
      alpha = c(-0.2, 0.0, -0.1),
      lambda = 0.175
    ),
    dist = list(sigma = list(sd = 0.4))
  ),
  seed = 2026
)

# Basic checks
head(sim$dataLong)
head(sim$dataEvent)
names(sim$truth)
```

# Troubleshooting

- **`corr` has no effect**: ensure nested marker-by-id random effects exist in `formulaLong`.
- **`id_marker_cov` length mismatch**: check `Q_idm` and required element count `M`.
- **family-scoped `dist_coefs` ignored**: verify family names are canonical (`gaussian`, `student_t`, `skew_normal`, ...).
- **transform errors**: for functional transforms, use supported function names (`power`, not `pow`).
- **parallel differences**: keep `seed` fixed; `simulate_JoiNMe` uses deterministic worker seeding.

# Mapping to `joinme()`

Use this rule:

- Keep `formulaLong`, `formulaEvent`, `formulaDist`, `assoc`, `transforms` as close as possible between simulation and fit.
- Encode your intended heterogeneity in simulation (`dist_coefs`, `re_params`) before fitting.
- Compare fitted summaries against `sim$truth` for calibration and recovery diagnostics.
